---
title: "lsp_prep"
output: github_document
params: 
    datasource: csv
---

```{r setup, echo = FALSE}

## Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')
library(stringr) # install.packages("stringr")
library(rgdal)  # install.packages('rgdal')
library(raster) # install.packages('raster')
library(rgeos)  # install.packages('rgeos')

# source
source('~/github/bhi/baltic2015/prep/common.r') #for create_readme() function

## rprojroot
root <- rprojroot::is_rstudio_project


## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)

dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')
dir_prep = make_path('baltic2015/prep') # replaces  file.path(dir_baltic, 'layers')

# root$find_file("README.md")
# 
# root$find_file("ao_need_gl2014.csv")
# 
# root <- find_root_file("install_ohicore.r", 
# 
# withr::with_dir(
#   root_file("DESCRIPTION"))

# NOT used?
 dir_lsp   = file.path(dir_prep, 'LSP')

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_lsp, 'lsp_prep.rmd') 
```

## 1. Background on MPA data  

### 1.1 Why MPAs?


## 2. MPA data  

### 2.1 Data source
#### 2.1.1 Shapefiles of MPA area
Shapefiles of current MPA areas were downloaded from [HELCOM MPAs Map Service](http://mpas.helcom.fi/apex/f?p=103:17::::::).  
    - Need to confirm date downloaded and date these data were last updated.  

#### 2.1.2 Management plan status csv
The status of the management plans associated with each MPA were downloaded from HELCOM's [MPA database] (http://mpas.helcom.fi/apex/f?p=103:40::::::) under the *Management Plans* tab.  Data were downloaded on 15 April 2016.  
    - Key columns in this csv file are "Site name" (MPA name) and "Management Plan status"  

There are three levels of management plan status that can be assigned to each MPA: *No plan*, *In development*, *Implemented*.  

A challenge is that each MPA can have multiple management plans associated with it. There is no limit to the number of plans not an ability to assess their relative importance. Different management plans for the same MPA can have different levels of implementation.  

**How to use this information*?*



## 3. LSP goal model overview
### 3.1 Status
Xlsp_country =  sum(w_i * MPA area)_m / Reference_pt_country  
    - Numerator is the sum over all MPAs within a country's EEZ of the MPA area weighted by the management status.  
    - w_i = value between 0 -1  
        - Need to assign weights to levels of management status.  
        - **One option**: *No plan* = 0.3, *In development* = .6, *Implemented* = 1.0.  **NEED FEEDBACK** 
    - Each country's status is applied to all BHI regions associated with that country.  

Reference_pt_country = 10% of the area in a country's EEZ is designated as an MPA and is 100% managed = 10% area country's EEZ  
    - This is based on the Convention on Biodiversity [target](https://www.cbd.int/sp/targets/rationale/target-11/)

### 3.2 Trend
##### 3.2.1 Alternative 1
MPA dbf contains the date the MPA was established. Therefore, we can calculate the cumulative area over time that are MPAs.  We can fit a cum. area ~ year. Calculate cumulative area from ealiest year of MPA established (1976) but for trend use the most recent 10 year period (2004 - 2013)?  It appears that countries mostly designate MPAs in chunks.

Only allow a positive slope since countries are not removing mpas? This would be automatic since regressing cumulative area on year (and have no info on areas removed). Also, if country reaches 10% goal, then the rate at which new area is added should go to zero. Then it is okay to have zero slope because the status will not decline in future - the required area has been allocated and status can only improve if all areas are managed. Then score of 100 is achieved.   


Cum_area_y = sum of all MPA areas established from year 1 to year y  

Cum_area_y = m*year_y + b; m = slope  

Trend = Future_year * m ; future_year = 5  

How to rescale trend?  If all values are between 0-1, do we need to rescale?  


#### 3.2.2 Alterative 2
There is limited information on MPA area from previous assessments. Need to read historic overview provided in [Baltic Sea Environment Proceedings NO. 124B Towards an ecologically coherent network of well-managed Marine Protected Areas](http://www.helcom.fi/lists/publications/bsep124b.pdf). Use this for the trend?  If area increasing, get positive trend?  

## 4. MPA data prep
Prep data layer

```{r TODO, echo=FALSE, message=FALSE}


## directory setup
dir_M <- c('Windows' = '//mazu.nceas.ucsb.edu/ohi',
           'Darwin'  = '/Volumes/ohi',    ### connect (cmd-K) to smb://mazu/ohi
           'Linux'   = '/home/shares/ohi')[[ Sys.info()[['sysname']] ]]

if (Sys.info()[['sysname']] != 'Linux' & !file.exists(dir_M)){
  warning(sprintf("The Mazu directory dir_M set in src/R/common.R does not exist. Do you need to mount Mazu: %s?", dir_M))
  # TODO: @jennifergriffiths reset dir_M to be where the SRC hold your shapefiles
}

dir_shp <- file.path(dir_M, 'git-annex/Baltic')

```

## Read in MPA and BHI regions shapefiles 
Also, get them in the same coordinate reference system for the Baltic. 

The MPA file is in the [LAEA coordinate reference system](http://spatialreference.org/ref/epsg/etrs89-etrs-laea/). 

```{r read in shapefiles, echo=FALSE, message=FALSE}

## inspect BHI regions ----
bhi <- rgdal::readOGR(dsn = path.expand(file.path(dir_shp, 'BHI_MCG_shapefile')),
                      layer = 'BHI_MCG_11052016')  
# bhi # coord. ref. : +proj=longlat


## explore/plot
# plot(bhi) #takes ~15 seconds
# bhi@data


## inspect mpa file ----
mpa = rgdal::readOGR(dsn = path.expand(file.path(dir_shp, 'bhi_MPA')),
                     layer = 'HELCOM_MPAs') 
# mpa # coord. ref. : +proj=laea

## plot
# plot(mpa) # quick
# head(mpa@data)


## transform bhi to mpa coordinate reference system (CRS) ----
## LAEA: http://spatialreference.org/ref/epsg/etrs89-etrs-laea/
## https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf
bhi = spTransform(bhi, mpa@proj4string)
# bhi # coord. ref. : +proj=laea

```

## Intersect BHI and HELCOM_MPA polygons

MPA regions were divided by with BHI region shape file, and thus we were able to calculate the total MPA area within each BHI region. MPA area per region is saved in the prep folder (`mpa_area_per_rgn.csv). 

The csv. file includes information: Area per MPA, Date established, MPA status, total MPA area per region.  


```{r intersect BHI and MPA, echo=FALSE, message=FALSE, cache=TRUE}

## plot on top of each other
plot(bhi, col = 'light blue', border = "grey", main = "BHI regions and MPAs intercept"); plot(mpa, col = 'blue', add = TRUE); legend('bottomright', c("BHI regions", "BHI borders", "MPA"), lty = c(1,1), lwd = c(2.5, 2.5, 2.5), col = c( "light blue", "grey", "blue"), text.font = 1, box.lty = 0 )

## system.time({mpa_bhi <- raster::intersect(mpa, bhi)}) # large file size and took hours to run. bhi ~ 45.5 Mb, mpa ~ 8.7 Mb.

#  plot(mpa_bhi, col ="red", border = "black")
#  head(mpa_bhi@data)

## save mpa_bhi_data file to prep/LSP folder 

#  write_csv(mpa_bhi@data, file.path(dir_lsp, 'mpa_bhi_data.csv'))

## calculate total MPA area per region and save to prep_LSP folder 

mpa_bhi_data = read.csv(file.path(dir_lsp, 'mpa_bhi_data.csv'))

### extract EEZ area (total area per BHI region) information from bhi shape file

bhi_data = bhi@data # knitting can't handle select(bhi@data)
write_csv(bhi_data, file.path(dir_lsp, 'bhi_map_data.csv'))

EEZ = bhi_data %>% data.frame() %>%
      dplyr::select(BHI_ID, 
                    eez_area_km2 = Area_km2) 

mpa_area_per_rgn = mpa_bhi_data %>%
                   dplyr::select(name_shape = Name, 
                                 BSPA_ID,
                                 BHI_ID, 
                                 country = Country,
                                 status_shape = MPA_Status,
                                 mpa_area = AreaTot_Ha,
                                 date_est = Date_Est) %>%
                    mutate(mpa_area_km2 = mpa_area * 0.01) %>%
                    dplyr::select(-mpa_area) %>%
                    full_join(EEZ, 
                             by = "BHI_ID")

write_csv(mpa_area_per_rgn, file.path(dir_prep, 'mpa_area_per_rgn.csv')) 

## count unique names in shape file 
mpa_name_shape = dplyr::select(mpa_area_per_rgn, name_shape) 
unique_mpa_shape = base::unique(mpa_name_shape) ## 163 count


```


## Status Calculation

There are three types of management status (and their corresponding weight): Designated (0.3), Designated and Partially Managed (0.6), and Designated and Managed (1).  

MPS management status data comes from ( ...?...). While combining MPA management data with MPA area data from the shapefile, 7 MPAs from the shapefile did not have a match from the management dataset. For those MPAs, status from the shapefile, which were not as detailed and updated as the status data, were used: 

| MPA | BSPA_ID | BHI_ID| Country| Status |
|----|-----|-----|-----|-----|
| Torhamns Archipelago | 110 | 14 | Sweden | Designated |
| Walkyriengrund | 171 | 10 | Germany | Designated |
| Ostseeküste am Brodtener Ufer |178 | 10 | Germany | Designated |
| Fehmarnbelt | 180 | 7 | Germany | Designated |
| Kadetrinne | 181 | 12 | Germany | Designated | 
| Jasmund National Park | 2 | 13 | Germany | Managed |
| Vorpommersche Boddenlandshaft National Park (West-Pommeranian Lagoon National Park) | 3 | 13 | Germany | Managed |

_The last two MPAs in Germany were assigned the weight of 0.6, the same as "Designated and Partially Managed"._ 

``` {r combine management status with mpa area and establishment info, echo = FALSE, message=FALSE}

mgmt_file = file.path(dir_lsp, "mpa_data_database/MPA_status_download06June2016.csv")
message(mgmt_file)
mgmt_status = read.csv(mgmt_file, sep = ";") %>%
              mutate(status = str_replace_all(Status, " ", "_")) %>%
              dplyr::select(name_csv = HELCOM_MPA_name,
                            BSPA_ID = ID,
                            status_csv = status) %>%
              dplyr::filter(!name_csv == "TEST SITE")

# count number of unique mpa names in new mpa mgmt status csv with a total of 175 observations

mpa_name = dplyr::select(mgmt_status, name_csv)
unique_mpa_csv_new = base::unique(mpa_name) ## total of 175 names that are unique. Including one "TEST SITE". 

# combine mpa_status with shape file data

mpa_mgmt = right_join(mgmt_status, mpa_area_per_rgn,  
                      by = "BSPA_ID") %>%
               mutate(mgmt_status = ifelse(is.na(status_csv), as.character(status_shape), status_csv),
                      # Mgmt_status = str_replace_all(mgmt_status, Managed, Designated_and_partially_managed),
                      date_est    = str_sub(date_est, start = -4)) # only select the year of establishment
             
write_csv(mpa_mgmt, file.path(dir_prep, 'mpa_mgmt_status_csv.csv'))

# count number of MPAs in shape file with no matches in csv file. 
# num_mismatch_new = dplyr::count(mpa_mgmt, !status_csv== "NA") 
# 13 rows (7 MPAs) have no match from the status csv file. 


````



``` {r status calculation, ECHO = FALSE, message=FALSE}

mgmt_weight = data.frame( mgmt_status = c("Designated", "Designated_and_partly_managed", "Designated_and_managed", "Managed"),
                          weight = c(0.3, 0.6, 1, 0.6) )

mpa_mgmt_with_wt = full_join( mgmt_weight, mpa_mgmt, 
                              by = 'mgmt_status') 

### calculating the reference point: 10% total eez per country

eez_data = mpa_mgmt_with_wt %>%
                 dplyr::select(BSPA_ID, BHI_ID, country, eez_area_km2) %>%
                 filter(!duplicated(BSPA_ID)) %>% #remove duplicated rows of the same MPA
                 filter(!duplicated(BHI_ID)) %>%
                 group_by(country) %>%
                 summarize(total_eez_km2 = sum(eez_area_km2)) %>%
                 ungroup() %>%
                 mutate(ref_eez_km2 = 0.1*total_eez_km2) %>%
                 dplyr::select( -total_eez_km2)
  

### calculating cumulative_mpa_area*weight, by country, by year 

cum_mpa_area_wt = mpa_mgmt_with_wt %>%
                 dplyr::select(BSPA_ID, BHI_ID, weight, country, mpa_area_km2, year = date_est) %>%
                 filter(!is.na(year), 
                        !duplicated(BSPA_ID)) %>% # the same MPA could exist in different BHI_IDs 
                 group_by(country, year) %>%
                 dplyr::summarize(sum_mpa_wt = sum(mpa_area_km2*weight)) %>%
                 arrange(year) %>%
                 mutate(cum_mpa_wt = cumsum(sum_mpa_wt)) %>%
                 ungroup()

### caculating status per country per year

status_per_year_by_country = full_join(cum_mpa_area_wt, eez_data, by = 'country') %>% 
  group_by(country, year) %>%
  summarize(status = round(max( -1, min(1, cum_mpa_wt/ref_eez_km2)) * 100, 1)) %>%
  ungroup() %>%
  arrange(country, year)

write_csv(r.status, file.path(dir_lsp, 'lsp_status_by_country_year.csv'))


# status by BHI_ID to be uploaded to layers folder
status_by_country = status_per_year_by_country %>%
  group_by(country) %>%
  filter(year == max(year)) %>%
  dplyr::select( -year) 

r.status = mpa_mgmt_with_wt %>%
           filter(!duplicated(BHI_ID)) %>%
           dplyr::select(rgn_id = BHI_ID, 
                  country) %>%
           full_join(status_by_country, 
                     by = 'country') %>% 
           dplyr::select(rgn_id, 
                         score = status) %>%
           mutate(dimension = 'status') 
                  
write_csv(r.status, file.path(dir_lsp, 'lsp_status_by_rgn.csv'))

# # to save a cleaned version of mpa_mgmt_with_wt as a separate csv file? 
# mpa_mgmt_with_wt_cleaned = mpa_mgmt_with_wt %>%
#                            select(name = name_shape, 
#                                   BHI_ID, 
#                                   BSPA_ID, 
#                                   country, 
#                                   mpa_area_km2, 
#                                   eez_area_km2, 
#                                   mgmt_status, 
#                                   weight, 
#                                   date_est)

```

The following graphs show LSP status by country and by BHI ID, as well as the distribution of MPAs by country and by country and management levels.

```{r plots, ECHO = FALSE, message=FALSE}
## plot status by country

status_by_country_plot <- ggplot(status_by_country, aes(x = country, y = status)) +
 geom_bar(stat = 'identity') +
 theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
 labs(title = 'LSP status by country',
      x = 'Country', 
      y = 'LSP status')

print(status_by_country_plot)

## plot status by BHI_ID

status_by_BHI_ID_plot <- ggplot(r.status, aes(x = rgn_id, y = score)) +
 geom_bar(stat = 'identity') +
 theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
 labs(title = 'LSP status by BHI ID',
      x = 'BHI ID', 
      y = 'LSP status')

print(status_by_BHI_ID_plot)


## plot the number of MPAs per country

num_mpa_per_country = mpa_mgmt %>%
                      group_by(country) %>%
                      summarize(count = length(!duplicated(name_shape))) %>%
                      ungroup 

mpa_per_country_plot <- ggplot(num_mpa_per_country, aes(x = country, y = count)) +
 geom_bar(stat = 'identity') +
 geom_text(aes(label = sprintf('n = %s', count), y = count), 
           size = 2, 
           angle = 90, hjust = 0, color = 'grey30') +
 theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
 labs(title = 'Number of MPAs per country',
      x = 'Country', 
      y = 'Number of MPAs')

print(mpa_per_country_plot)

## plot number of MPAs per country by mgmt levels 

mgmt_weight_alt = data.frame( mgmt_status = c("Designated", "Designated_and_partly_managed",           "Designated_and_managed"),
                          weight = c(0.3, 0.6, 1) ) # ignore "managed" category 

num_mpa_per_country_mgmt = mpa_mgmt_with_wt %>%
                           group_by(country, weight) %>%
                           summarize(count = length(!duplicated(name_shape))) %>%
                           ungroup %>%
                           full_join(mgmt_weight_alt, by = "weight")

num_country_mgmt_plot <- ggplot(num_mpa_per_country_mgmt, aes(x = country, y = count, fill = mgmt_status)) +
 geom_bar(stat = 'identity') +
 geom_text(aes(label = sprintf('n = %s', count), y = count), 
           size = 2, 
           angle = 90, hjust = 0, color = 'grey30') +
 theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
 labs(title = 'Number of MPAs by country and management level',
      x = 'Country', 
      y = 'Number of MPAs',
      fill = 'Management Level')

print(num_country_mgmt_plot)


```

## Trend calculation

Cum_area_y = sum of all MPA areas established from year 1 to year y

Cum_area_y = m*year_y + b; m = slope

Trend = Future_year * m ; future_year = 5

TODO: see the three trials of trend calculation... 

``` {r trend calculation, ECHO = FALSE, message = FALSE}

# load trend data: For most countries, there were 1-3 years of data in random years. linear model doesn't work on the original data.  Use complete and fill to fill in the blank years. 

trend_data = mpa_mgmt %>%
  dplyr::select(name = name_shape, BSPA_ID, BHI_ID, country, date_est, mpa_area_km2) %>%
  filter(!date_est == 'NA') %>% # remove one Russian MPA that has no date of establishment
  filter(!duplicated(BSPA_ID)) %>% # remove duplicated rows of the same MPA area data
  mutate(year = ifelse(date_est <2004, 2004, date_est)) %>% # select the most recent 10 years of data (2004 - 2013). any year prior to 2004 is treated as 2004. 
  group_by(country, year) %>%
  summarize(area_total = sum(mpa_area_km2)) %>%
  mutate(area_cumulative = cumsum(area_total)) %>%
  ungroup()

write_csv(trend_data, file.path(dir_lsp, 'cumulative_mpa_area_trend_data.csv'))
          
## plot: didn't work. 
# mpa_area_by_year <- ggplot(trend_data, aes(x = country, y = area_cumulative, fill = year)) +
#  geom_bar(stat = 'identity') +
#  geom_text(aes(label = sprintf('n = %s', area), y = area_cumulative), 
#            size = 2, 
#            angle = 90, hjust = 0, color = 'grey30') +
#  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
#  labs(title = '',
#       x = 'Country', 
#       y = 'Number of MPAs',
#       fill = 'Management Level')
# 
# print(mpa_area_by_year)

## TRY 1: fill in missing data from 2004-2013. Many countries started with 0 km2 MPA or very low in 2004, and all trends are therefore 1. I don't think this is a good approach for trend... 

trend_data_1 = trend_data %>%
  mutate(year = as.integer(year)) %>%
  tidyr::complete(year = full_seq(year, 1), nesting(country)) %>%
  group_by(country) %>%
  fill(area_cumulative) %>%
  mutate(area_cumulative = ifelse(is.na(area_cumulative), 0, area_cumulative))

  
r.trend.1 = trend_data_1 %>%
  group_by(country) %>%
  filter(!length(year) == 1) %>%
  do(dlm = lm(area_cumulative ~ year, data = .)) %>%
  mutate(score = pmax(-1, pmin(1, coef(dlm)[['year']]*5))) %>% 
  ungroup %>%
  dplyr::select(country, score); #head(r.trend.1)

#     country score

# 1   Denmark     1
# 2   Estonia     1
# 3   Finland     1
# 4   Germany     1
# 5    Latvia     1
# 6 Lithuania     1
# 7    Poland     1
# 8    Russia     1
# 9    Sweden     1


## TRY 2: use the most recent 5 years of cumulative area data to draw a trend. Trend is either 1 or 0. 

r.trend.2 = trend_data_1 %>%
  filter(year > (max(year) - 5)) %>% # most recent 5 years of data (2009- 2013)
  group_by(country) %>%
  do(dlm = lm(area_cumulative ~ year, data = .)) %>%
  mutate(score = round(pmax(-1, pmin(1, coef(dlm)[['year']]*5))), 1) %>% 
  ungroup %>%
  dplyr::select(country, score); #head(r.trend.2)

#     country score
# 1   Denmark     0
# 2   Estonia     0
# 3   Finland     0
# 4   Germany     0
# 5    Latvia     1
# 6 Lithuania     1
# 7    Poland     0
# 8    Russia     0
# 9    Sweden     1

## TRY 3:calculate the status of the past five years and draw a trend for those years. 

trend_data_3 = status_per_year_by_country %>%
  filter(!is.na(year)) %>% # one russian MPA didn't have info on established year
  mutate(year = as.integer(year)) %>%
  tidyr::complete(year = full_seq(year, 1), nesting(country)) %>%
  group_by(country) %>%
  fill(status) %>%
  mutate(status = ifelse(is.na(status), 0, status)) %>%
  ungroup()

r.trend.3 = trend_data_3 %>%
  filter(year > (max(year) - 5)) %>% # most recent 5 years of data (2009- 2013)
  group_by(country) %>%
  summarize(dlm = cor(status, year))
  
  do(dlm = lm(status ~ as.integer(year), data = .)) %>%
  
  mutate(score = coef(dlm[['year']]))
  mutate(score = pmax(-1, pmin(1, coef(dlm[['year']])*5))) ### this last line did not work... not sure why...

## Trial #4: compare the management status from the shape file (older) and the status csv file (newer), and using the relative status of the two to represent change: If mgmt_wt_csv > mgmt_wt_shape, trend = 0.2; if equal, trend = 0; if mgmt_wt_csv < mgmt_wt_shape, trend = -0.2. 

## might be hard to justify using two data sets that dont match.... 

mgmt_wt_csv = data.frame( status_csv = c("Designated", "Designated_and_partly_managed", "Designated_and_managed"),
                          weight_csv = c(0.3, 0.6, 1) )

mgmt_wt_shape = data.frame( status_shape = c("Designated", "Managed"), 
                            weight_shape = c(0.3, 1))
                              
trend_data_4 = mpa_mgmt %>%
  dplyr::select(BSPA_ID, status_csv, status_shape, BHI_ID, country) %>%
  full_join(mgmt_wt_csv) %>%
  full_join(mgmt_wt_shape) %>% 
  mutate(weight_csv = ifelse(is.na(weight_csv), weight_shape, weight_csv), 
         wt_combined = 0, 
         wt_combined = ifelse(weight_csv > weight_shape, 0.2, wt_combined),
         wt_combined = ifelse(weight_csv < weight_shape, -0.2, wt_combined)) %>%
  group_by(country) %>%
  filter(!duplicated(BSPA_ID)) %>%
  summarize(score = round(mean(wt_combined), 1)) %>%
  ungroup


trend_4 = mpa_mgmt %>%
  dplyr::select(rgn_id = BHI_ID, country) %>%
  full_join(trend_data_4)

         # OR:
         # score = ifelse(weight_csv > weight_shape, 0.2, -0.2), 
         # score = ifelse(weight_csv = weight_shape, 0, score))

write_csv(trend_data_4, file.path(dir_lsp, "trend_lsp_4.csv"))
  

``` 
