---
title: "lsp_prep"
output: github_document
params: 
    datasource: csv
---
```{r setup}

## Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')

# source
source('~/github/bhi/baltic2015/prep/common.r') #for create_readme() function

## rprojroot
root <- rprojroot::is_rstudio_project


## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)

dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')
dir_prep = make_path('baltic2015/prep') # replaces  file.path(dir_baltic, 'layers')

# root$find_file("README.md")
# 
# root$find_file("ao_need_gl2014.csv")
# 
# root <- find_root_file("install_ohicore.r", 
# 
# withr::with_dir(
#   root_file("DESCRIPTION"))


dir_lsp   = file.path(dir_prep, 'LSP')

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_lsp, 'lsp_prep.rmd') 
```

## 1. Background on MPA data  

### 1.1 Why MPAs?


## 2. MPA data  

### 2.1 Data source
#### 2.1.1 Shapefiles of MPA area
Shapefiles of current MPA areas were downloaded from [HELCOM MPAs Map Service](http://mpas.helcom.fi/apex/f?p=103:17::::::).  
    - Need to confirm date downloaded and date these data were last updated.  

#### 2.1.2 Management plan status csv
The status of the management plans associated with each MPA were downloaded from HELCOM's [MPA database] (http://mpas.helcom.fi/apex/f?p=103:40::::::) under the *Management Plans* tab.  Data were downloaded on 15 April 2016.  
    - Key columns in this csv file are "Site name" (MPA name) and "Management Plan status"  

There are three levels of management plan status that can be assigned to each MPA: *No plan*, *In development*, *Implemented*.  

A challenge is that each MPA can have multiple management plans associated with it. There is no limit to the number of plans not an ability to assess their relative importance. Different management plans for the same MPA can have different levels of implementation.  

**How to use this information*?*



## 3. LSP goal model overview
### 3.1 Status
Xlsp_country =  sum(w_i * MPA area)_m / Reference_pt_country  
    - Numerator is the sum over all MPAs within a country's EEZ of the MPA area weighted by the management status.  
    - w_i = value between 0 -1  
        - Need to assign weights to levels of management status.  
        - **One option**: *No plan* = 0.3, *In development* = .6, *Implemented* = 1.0.  **NEED FEEDBACK** 
    - Each country's status is applied to all BHI regions associated with that country.  

Reference_pt_country = 10% of the area in a country's EEZ is designated as an MPA and is 100% managed = 10% area country's EEZ  
    - This is based on the Convention on Biodiversity [target](https://www.cbd.int/sp/targets/rationale/target-11/)

### 3.2 Trend
##### 3.2.1 Alternative 1
MPA dbf contains the date the MPA was established. Therefore, we can calculate the cumulative area over time that are MPAs.  We can fit a cum. area ~ year. Calculate cumulative area from ealiest year of MPA established (1976) but for trend use the most recent 10 year period (2004 - 2013)?  It appears that countries mostly designate MPAs in chunks.

Only allow a positive slope since countries are not removing mpas? This would be automatic since regressing cumulative area on year (and have no info on areas removed). Also, if country reaches 10% goal, then the rate at which new area is added should go to zero. Then it is okay to have zero slope because the status will not decline in future - the required area has been allocated and status can only improve if all areas are managed. Then score of 100 is achieved.   


Cum_area_y = sum of all MPA areas established from year 1 to year y  

Cum_area_y = m*year_y + b; m = slope  

Trend = Future_year * m ; future_year = 5  

How to rescale trend?  If all values are between 0-1, do we need to rescale?  


#### 3.2.2 Alterative 2
There is limited information on MPA area from previous assessments. Need to read historic overview provided in [Baltic Sea Environment Proceedings NO. 124B Towards an ecologically coherent network of well-managed Marine Protected Areas](http://www.helcom.fi/lists/publications/bsep124b.pdf). Use this for the trend?  If area increasing, get positive trend?  

## 4. MPA data prep
Prep data layer

```{r TODO, echo=FALSE, message=FALSE}


## directory setup
dir_M <- c('Windows' = '//mazu.nceas.ucsb.edu/ohi',
           'Darwin'  = '/Volumes/ohi',    ### connect (cmd-K) to smb://mazu/ohi
           'Linux'   = '/home/shares/ohi')[[ Sys.info()[['sysname']] ]]

if (Sys.info()[['sysname']] != 'Linux' & !file.exists(dir_M)){
  warning(sprintf("The Mazu directory dir_M set in src/R/common.R does not exist. Do you need to mount Mazu: %s?", dir_M))
  # TODO: @jennifergriffiths reset dir_M to be where the SRC hold your shapefiles
}


## libraries
library(dplyr)
library(rgdal)  # install.packages('rgdal')
library(raster) # install.packages('raster')
library(rgeos)  # install.packages('rgeos')

```

## Visualize MPA data
```{r echo=FALSE, message=FALSE}
## read in data...

 tmp_filepath <- '~/Desktop/bhi_4.27.16'

HELCOM_MPAs = rgdal::readOGR(dsn = path.expand(tmp_filepath),
                             layer = 'HELCOM_MPAs') # dsn is the path name & layer is the name of the file. 
## plot
plot(HELCOM_MPAs)

## explore this spatial polygon and its data
HELCOM_MPAs
head(HELCOM_MPAs@data)

## Read in MPA management plan status file from "~/github/bhi/baltic2015/prep/LSP/mpa_data_database/MPA_management_plans_20160415.csv"

## shapefiles


## plot data...

```

