---
title: "Scores"
output:
  html_document:
    toc: true
    toc_float: true
#    keep_md: true
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
source("./_site.R")
```

These figures represent Baltic Health Index (BHI) scores as reported for different regions.  

Scores were calculated for 42 BHI regions and aggregated (using a weighted average) to create an overall score for the Baltic. Additionally, there are 9 EEZ aggregated regions and 17 SUBBASIN aggregated regions. 

Additional figures are within the [`BHI_regions`](https://github.com/OHI-Science/bhi/tree/draft/baltic2015/reports/figures/BHI_regions), [`EEZ`](https://github.com/OHI-Science/bhi/tree/draft/baltic2015/reports/figures/EEZ) and [`SUBBASIN`](https://github.com/OHI-Science/bhi/tree/draft/baltic2015/reports/figures/SUBBASIN) folders, respectively.

----


<!---
NOTE: Here the flower plots are displayed for each region and the overall assessment area (rgn_id = 0). Figures will be displayed directly from online in the reports/figures folder: to make sure you're displaying the most recent flower plots here's what to do: 
1. from the master branch, save new flower plots, and push
2. from the gh-pages branch, reknit this page or rebuild the website, and push!
--->

```{r plot_flower, results='asis'}

## regions info: order regions to start with whole study_area
regions <- 
  read_csv(sprintf('%s/spatial/regions_lookup_complete.csv', dir_scenario_gh)) %>%
    select(region_id, region_name = label)

## path for figures to display
regions <- regions %>%
  mutate(flower_png = sprintf('%s/reports/figures/BHI_regions/flower_%s.png',
                              dir_scenario_gh,
                              stringr::str_replace_all(region_name, ' - ', '-'))) %>%
  mutate(flower_png = stringr::str_replace_all(flower_png, ' ', ''))


## display header, flower ----

cat(sprintf('## %s\n\n![](%s)\n\n\n\n', regions$region_name, regions$flower_png), 
    sep = "")

```


[<img src="https://docs.google.com/drawings/d/1633uvIYgsBhiKBddQAaTlvYDJgOK1uiDB6wkooNyhFc/pub?w=576&h=288" width="100px">](http://ohi-science.org)  
[ohi-science.org home](http://ohi-science.org)
