---
title: "cs_prep"
output: github_document
params: 
    datasource: csv
---

# Carbon Storage (CS) Data prep

```{r setup}

## Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')

# source
source('~/github/bhi/baltic2015/prep/common.r') #for create_readme() function

## rprojroot
root <- rprojroot::is_rstudio_project

## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)


dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')


# root$find_file("README.md")
# 
# root$find_file("ao_need_gl2014.csv")
# 
# root <- find_root_file("install_ohicore.r", 
# 
# withr::with_dir(
#   root_file("DESCRIPTION"))



dir_cs   = file.path(dir_prep, 'CS')

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_cs, 'cs_prep.rmd') 
```

## 1. Background on Zostera data

### 1.1 Why Zostera data

### 1.2 Zostera species in the Baltic Sea

### 1.3 Related publications
[Boström et al 2014](http://onlinelibrary.wiley.com/doi/10.1002/aqc.2424/abstract)  

[HELCOM Red List Biotope Information Sheet](http://helcom.fi/Red%20List%20of%20biotopes%20habitats%20and%20biotope%20complexe/HELCOM%20Red%20List%20AA.H1B7,%20AA.I1B7,%20AA.J1B7,%20AA.M1B7.pdf)  


## 2. Data source  
### 1.1 Download information 
[HELCOM Marine Spatial Planning Map Service](http://maps.helcom.fi/website/msp/index.html)  
Select -  Marine Spatial Planning - Ecology - Ecosystem Health status  
Data layer - "Zostera Meadows"  
Downloaded on 10 May 2016 by Jennifer Griffiths

[Metadata link](http://maps.helcom.fi/website/getMetadata/htm/All/Zostera%20meadows.htm)  

### 2.2 Additional background
*Notes from Joni Kaitaranta (HELCOM)*: According to the Zostera meadows metadata there is many data sources. The dataset was compiled in 2009-2010 for the HOLAS I assessment so the dataset is not very recent. Major source was Boström et al. 2003. In: Spalding et al. World Atlas of Seagrasses but there was also national e.g. from NERI Denmark downloaded in 2009 from a URL that is no longer valid.  

*Data Structure*: Is unclear, spatial files need to be reviewed. Is it just points, or does it include areal coverage of Zostera. Data observations are accompanies by either "dense" or "sparse." 

## 3. CS goal model overview
### 3.1 Status
X_cs_region = Current_area_Zostera_meadows_region / Reference_pt_region  

Reference_pt_region = Current_area_Zostera_meadows * 1.25  
This is based upon ["During the last 50 years the distribution of the Zostera marina biotope has declined >25%. The biotope has declined to varying extents in the different Baltic Sea regions."](http://helcom.fi/Red%20List%20of%20biotopes%20habitats%20and%20biotope%20complexe/HELCOM%20Red%20List%20AA.H1B7,%20AA.I1B7,%20AA.J1B7,%20AA.M1B7.pdf)  

**If data are not areal coverage**: if data are simply points of presence, need to rethink goal model


### 3.2 Trend
Consider using the trend of NUT subcomponent for CW goal (e.g. secchi status trend).  



## Layer prep

```{r TODO, echo=FALSE, message=FALSE}
## read in data...

## plot data...

```
