---
title: "resilience_prep"
output: github_document
params: 
    datasource: csv
---

# Preparation of Resilience Data Layers

```{r setup}

## Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')

## rprojroot
root <- rprojroot::is_rstudio_project

## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)


dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')


# root$find_file("README.md")
# 
# root$find_file("ao_need_gl2014.csv")
# 
# root <- find_root_file("install_ohicore.r", 
# 
# withr::with_dir(
#   root_file("DESCRIPTION"))

source('~/github/bhi/baltic2015/prep/common.r')
dir_res    = file.path(dir_prep, 'resilience')

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_res, 'resilience_prep.rmd') 
```

## 1. Ecological 
## 1.1 Biological Integrity

## 1.2 Goal-specific Regulations

### 1.2.1 Background
This component of resilience captures the capacity of pressures to be offset by regulations/laws.  

There are three components to goal-specific regulation resilience: *(1) existence, (2) compliance, (3) enforcement*  

Three international treaties are assessed for all BHI countries for *existence* only.These agreements are´:  
[Convention on Biodiversity (CBD)](https://www.cbd.int)  
[Convention on International Trade in Endangered Species of Wild Fauna and Flora (CITES)](https://www.cites.org)  
[Paris Climate Agreement (COP21)](https://treaties.un.org/pages/ViewDetails.aspx?src=TREATY&mtdsg_no=XXVII-7-d&chapter=27&lang=en)

Two Baltic Region agreements will be assessed for *existence* and *comliance* for all countries. These are: Baltic Sea Action Plan (BSAP); Helsinki Convention (HELCOM).

13 EU directives will be assessed for *existence* and *comliance* for BHI countries that are EU member states. These are: Water Framework Directive (WFD), Marine Spatial Planning Directive (MSPD), Marine Strategy Framework Directive (MSFD), Habitat Directive (HD), Common Fisheries Policy (CFP), National Emmissions Ceilings (NEC), Industrial Emissions Directive (IED), Urban Waste Water Treatment Directive (UWWTD), Bathing Water Directive (BWD), Conservation of Wild Birds (BIRDS), Registration, Evaluation, Authorisation and Restriction of Chemicals (REACH), Peristent Organic Pollutants (POP), Nitrates Directive (ND).

### 1.2.2 Scoring
By country for each level assessed for each regulation  

#### 1.2.2.1 Existence
Scored as yes (1) or no (0)  
Score_Existence = Score / Max
Max = 1

#### 1.2.2.2 Compliance
Fail (0), Partial (1), Full (3)  
Not equal increments because Full compliance rewarded more here.

Country_compliance_directive = Mean (compliance score)_all_compliance_components

#### 1.2.2.3 Total regulation score
Score_Existence + Country_compliance_directive / (Possible_Existence + Possible_Compliance) - *Needs to be rescaled between 0-1?*  

Compliance is weighted more heavily than existence when both are scored.  


### 1.2.3 Goal Weighting
Each regulation will be weighted (1 or 2) for each goal to which is it applied.

### 1.2.4 Resilience Data Layer
Country-specific resilience score for each goal is assigned to each BHI region associated with that country.  

Country_resilience_goal_x = Sum overall all regulations for a goal[country_regulation_score * goal_weight]  

Country_resilience_goal_x  - rescale between 0-1?  


### 1.2.5 Goal-specific data layer preparation

#### 1.2.5.1 Read in data
```{r read in goal specific reg data}
## read in data

eu_baltic_scores = read.csv(file.path(dir_res, 'eu_regs_resilience_data.csv'), sep=";")

international_scores = read.csv(file.path(dir_res, 'InternationalAgreements_resilience_data.csv'), sep=";")


## 
str(eu_baltic_scores)
dim(eu_baltic_scores)

str(international_scores)
dim(international_scores)
```

#### 1.2.5.2 simply data objectives, combine into single object
```{r simplfy}
eu_baltic_lookup = eu_baltic_scores ## keep the info of text references in case needed
internat_lookup = international_scores


eu_baltic_scores = eu_baltic_scores %>%
                  mutate(Existence = "Yes",
                         ExistenceLevel = "EU_Law",
                         ExistenceNotes = "")%>% ## add existence so that compatible with international ## all EU and Baltic directives / agreements exist (if at EU level are at country level)
                  select(-DirectiveName,-ComplianceText,-ComplianceTextReference,-Notes_UncertainitiesInECreport,-Notes_AreasForFurtherInvestigation)%>%
                  select(DirectiveAbbreviation,ReportYear,CountryName,Existence,ExistenceLevel,ExistenceNotes,IsCountryAssessed,ComplianceComponent, ComplianceComponentScore,NAexplanation,Comments) %>% ## reorder
                      dplyr::rename(NAexplanationCompliance = NAexplanation,
                                    IsCountryComplianceAssessed = IsCountryAssessed)

colnames(eu_baltic_scores)

international_scores = international_scores %>%
                      mutate(IsCountryAssessed = "No")  %>% ## this is for compliance
                        select(-DirectiveName,-DataLink,-Notes_UncertainitiesInECreport,-Notes_AreasForFurtherInvestigation) %>%
                  select(DirectiveAbbreviation,ReportYear,CountryName,Existence,ExistenceLevel,ExistenceNotes,IsCountryAssessed,ComplianceComponent, ComplianceComponentScore,NAexplanation,Comments) %>%## reorder
                      dplyr::rename(NAexplanationCompliance = NAexplanation,
                                    IsCountryComplianceAssessed = IsCountryAssessed)

colnames(international_scores)

goal_spec = rbind(eu_baltic_scores, international_scores)



```

#### 1.2.5.3 assign numeric score

```{r numeric score}
goal_spec = goal_spec %>% 
            mutate(exist_numeric = ifelse(Existence == "Yes",1,0),
                   compli_numeric = ifelse(is.na(ComplianceComponentScore),NA,
                                    ifelse(ComplianceComponentScore == "Not applicable", NA,
                                    ifelse(ComplianceComponentScore == "fail",0,
                                    ifelse(ComplianceComponentScore == "partial",1,
                                    ifelse(ComplianceComponentScore == "full",3,NA))))))
            


```

#### 1.2.5.4 Plot numeric score
With the numeric score for Compliance Components, can not see overlapping scores for different components
```{r plot raw score}

ggplot(goal_spec) + 
  geom_point(aes(CountryName,exist_numeric))+
  facet_wrap(~DirectiveAbbreviation)+
  ylim(0,1)+
   theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Existence Score")

ggplot(goal_spec) + 
  geom_point(aes(CountryName,compli_numeric))+
  facet_wrap(~DirectiveAbbreviation)+
  ylim(0,1)+
   theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Compliance Score")

```

#### 1.2.5.5 Exclude Compliance Components that have NA
If a component has NA for assessment, have to exclude.
This will lead to differences in the number of components scored for different countries in the same directive. Depends on case, but could be that country did not assess or report so then review of report does not assess.
```{r exclude NA compliance}
goal_spec2 = goal_spec %>%
            filter(!is.na(compli_numeric))

dim(goal_spec);dim(goal_spec2)


## count to see if the same directive has a different number of components assessed for each country
check_compli_n = goal_spec2 %>%
                 select(DirectiveAbbreviation,CountryName,compli_numeric) %>%
                 group_by(DirectiveAbbreviation,CountryName)%>% 
                 count(DirectiveAbbreviation,CountryName)

ggplot(check_compli_n) +
  geom_point(aes(CountryName,n))+
  facet_wrap(~DirectiveAbbreviation)+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Do Compliance Components differ by Country")
```



#### 1.2.5.5 Mean Compliance score
```{r mean compliance score}
goal_spec3 = goal_spec2 %>%
            select(DirectiveAbbreviation,CountryName,exist_numeric,compli_numeric) %>%
            group_by(DirectiveAbbreviation,CountryName)%>%          
            transmute(exist_mean = mean(exist_numeric),
                      compli_mean = mean(compli_numeric))
                      
goal_compli_n = goal_spec2 %>%
            select(DirectiveAbbreviation,CountryName,compli_numeric) %>%
            filter(!is.na(compli_numeric)) %>% ## remove NAs
            group_by(DirectiveAbbreviation,CountryName)%>%  count(DirectiveAbbreviation,CountryName) ## number of compliance components assessed


goal_spec3 = goal_spec3 %>%
        left_join(., goal_compli_n, by=c("DirectiveAbbreviation", "CountryName")) ## join n_obs
                      
                     
                      
               


```

#### 1.2.5.6 Overall score
Max score is 4 (1 for existence, 3 for compliance)
```{r overal score}
goal_spec4 = goal_spec3 %>%
              mutate(max_score = 4,
                     overall_score = (exist_mean + compli_mean)/max_score)



```
#### 1.2.5.7 Plot Overall score
```{r plot overall score}
ggplot(goal_spec4)+
  geom_point(aes(CountryName,overall_score))+
  facet_wrap(~DirectiveAbbreviation)+
  ylim(0,1)+
   theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Overall Score per Regulation")

```
## TO DO 
Need to rethink exclucding compli with NA because then do not have all directives at the end!



## 2. Social
