---
title: "resilience_prep"
output:
  github_document:
    toc: true
    toc_depth: 3
params: 
    datasource: csv
---

# Preparation of Resilience Data Layers

```{r setup}

## Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')

## rprojroot
root <- rprojroot::is_rstudio_project

## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)


dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')


# root$find_file("README.md")
# 
# root$find_file("ao_need_gl2014.csv")
# 
# root <- find_root_file("install_ohicore.r", 
# 
# withr::with_dir(
#   root_file("DESCRIPTION"))

source('~/github/bhi/baltic2015/prep/common.r')
dir_res    = file.path(dir_prep, 'resilience')

dir_wgi    = file.path(dir_prep,'pressures/wgi_social') ## to source wgi data

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_res, 'resilience_prep.rmd') 
```

## 1. Ecological 
## 1.1 Biological Integrity
The biodiversity data layer will be used to represent biological integrity.  



## 1.2 Goal-specific Regulations

### 1.2.1 Background
This component of resilience captures the capacity of pressures to be offset by regulations/laws.  

There are three components to goal-specific regulation resilience: *(1) existence, (2) compliance, (3) enforcement*  

Three international treaties are assessed for all BHI countries for *existence* only.These agreements are:  
[Convention on Biodiversity (CBD)](https://www.cbd.int)  
[Convention on International Trade in Endangered Species of Wild Fauna and Flora (CITES)](https://www.cites.org)  
[Paris Climate Agreement (COP21)](https://treaties.un.org/pages/ViewDetails.aspx?src=TREATY&mtdsg_no=XXVII-7-d&chapter=27&lang=en)

Two Baltic Region agreements will be assessed for *existence* and *compliance* for all countries. These are: Baltic Sea Action Plan (BSAP); Helsinki Convention (HELCOM).

13 EU directives will be assessed for *existence* and *compliance* for BHI countries that are EU member states. These are: Water Framework Directive (WFD), Marine Spatial Planning Directive (MSPD), Marine Strategy Framework Directive (MSFD), Habitat Directive (HD), Common Fisheries Policy (CFP), National Emmissions Ceilings (NEC), Industrial Emissions Directive (IED), Urban Waste Water Treatment Directive (UWWTD), Bathing Water Directive (BWD), Conservation of Wild Birds (BIRDS), Registration, Evaluation, Authorisation and Restriction of Chemicals (REACH), Peristent Organic Pollutants (POP), Nitrates Directive (ND). 

*Note* IED and MSPD, while having compliance components can not currently have compliances assessed as the deadlines for compliance and reporting have only just recently occurred or have not occurred yet (are 2016 dates).

### 1.2.2 Scoring
By country for each level assessed for each regulation  

#### 1.2.2.1 Existence
Scored as yes (1) or no (0)  
Score_Existence = Score / Max
Max = 1

#### 1.2.2.2 Compliance
Fail (0), Partial (1), Full (3)  
Not equal increments because Full compliance rewarded more here.

Country_compliance_directive = Mean (compliance score)_all_compliance_components

#### 1.2.2.3 Total regulation score
Score_Existence + Country_compliance_directive / (Possible_Existence + Possible_Compliance)

Compliance is weighted more heavily than existence when both are scored.  

### 1.2.3 Goal Weighting & Mapping
Each regulation will be weighted (1 or 2) for each goal to which is it applied. Regulations are mapped to all the goals to which they directly influence/combat pressures.

### 1.2.4 Resilience Data Layer
Country-specific resilience score (G) for each goal is assigned to each BHI region associated with that country.  

G = sum(w_i * G_i) / sum(wi_i) ;  G_i = specific regulatory dataset total regulation score, w_i = weight for each i dataset used to assess G  




### 1.2.5 Goal-specific data layer preparation

#### 1.2.5.1 Read in data
```{r read in goal specific reg data}
## read in data

eu_baltic_scores = read.csv(file.path(dir_res, 'eu_regs_resilience_data.csv'), sep=";")

international_scores = read.csv(file.path(dir_res, 'InternationalAgreements_resilience_data.csv'), sep=";")


## 
str(eu_baltic_scores)
dim(eu_baltic_scores)

str(international_scores)
dim(international_scores)
```

#### 1.2.5.2 simply data objectives, combine into single object
```{r simplfy}
eu_baltic_lookup = eu_baltic_scores ## keep the info of text references in case needed
internat_lookup = international_scores


eu_baltic_scores = eu_baltic_scores %>%
                  mutate(Existence = "Yes",
                         ExistenceLevel = "EU_Law",
                         ExistenceNotes = "")%>% ## add existence so that compatible with international ## all EU and Baltic directives / agreements exist (if at EU level are at country level)
                  select(-DirectiveName,-ComplianceText,-ComplianceTextReference,-Notes_UncertainitiesInECreport,-Notes_AreasForFurtherInvestigation)%>%
                  select(DirectiveAbbreviation,ReportYear,CountryName,Existence,ExistenceLevel,ExistenceNotes,IsCountryAssessed,ComplianceComponent, ComplianceComponentScore,NAexplanation,Comments) %>% ## reorder
                      dplyr::rename(NAexplanationCompliance = NAexplanation,
                                    IsCountryComplianceAssessed = IsCountryAssessed)

colnames(eu_baltic_scores)

international_scores = international_scores %>%
                      mutate(IsCountryAssessed = "No")  %>% ## this is for compliance
                        select(-DirectiveName,-DataLink,-Notes_UncertainitiesInECreport,-Notes_AreasForFurtherInvestigation) %>%
                  select(DirectiveAbbreviation,ReportYear,CountryName,Existence,ExistenceLevel,ExistenceNotes,IsCountryAssessed,ComplianceComponent, ComplianceComponentScore,NAexplanation,Comments) %>%## reorder
                      dplyr::rename(NAexplanationCompliance = NAexplanation,
                                    IsCountryComplianceAssessed = IsCountryAssessed)

colnames(international_scores)

goal_spec = rbind(eu_baltic_scores, international_scores)



```


#### 1.2.5.3 assign numeric score

```{r numeric score}
goal_spec = goal_spec %>% 
            mutate(exist_numeric = ifelse(Existence == "Yes",1,0),
                   compli_numeric = ifelse(is.na(ComplianceComponentScore),NA,
                                    ifelse(ComplianceComponentScore == "Not applicable", NA,
                                    ifelse(ComplianceComponentScore == "fail",0,
                                    ifelse(ComplianceComponentScore == "partial",1,
                                    ifelse(ComplianceComponentScore == "full",3,NA))))))
            


```

#### 1.2.5.4 Plot numeric score
With the numeric score for Compliance Components, can not see overlapping scores for different components
```{r plot raw score}

ggplot(goal_spec) + 
  geom_point(aes(CountryName,exist_numeric))+
  facet_wrap(~DirectiveAbbreviation)+
  ylim(0,1)+
   theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Existence Score")

ggplot(goal_spec) + 
  geom_point(aes(CountryName,compli_numeric))+
  facet_wrap(~DirectiveAbbreviation)+
  ylim(0,1)+
   theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Compliance Score")

```

#### 1.2.5.5 Separate Existance and Compliance
```{r seperate objects for exist and compli}

goal_spec_exist = goal_spec %>%
                  select(DirectiveAbbreviation:ExistenceNotes, exist_numeric)

goal_spec_compli = goal_spec %>%
                  select(DirectiveAbbreviation:CountryName, IsCountryComplianceAssessed:Comments, compli_numeric)


```

#### 1.2.5.6 How many compliance components assessed?
Some directives have no components assessed. Others have a subset that could not be assessed - these might differ by country. Minimal differences by country. WDF not assessed for Denmark. Germany assessed for one more POP component. Poland asssesse for one fewer UWWTD component.  
**Gray bars are components not assessed**
```{r}

goal_spec_compli_n1 = goal_spec_compli %>%
                      count(DirectiveAbbreviation, CountryName,compli_numeric)

ggplot(goal_spec_compli, aes(CountryName,fill=as.character(compli_numeric)))+ 
  geom_bar(position="stack")+
  facet_wrap(~DirectiveAbbreviation)+
  ylab("Count of Compliance Componenets")+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Compliance Components Assessed by Country & Directive")

```


#### 1.2.5.7 Mean Compliance score
```{r mean compliance score}
## calculate the mean value for all scored compliance by directive and country
goal_spec_compli_mean = goal_spec_compli%>%
            select(DirectiveAbbreviation,CountryName,compli_numeric) %>%
            group_by(DirectiveAbbreviation,CountryName)%>%          
            summarise(compli_mean = mean(compli_numeric, na.rm=TRUE)) %>%
            ungroup()

                      
## get the number of components assessed
goal_spec_compli_n2 = goal_spec_compli%>%
                      select(DirectiveAbbreviation,CountryName,compli_numeric) %>%
                      filter(!is.na(compli_numeric))%>%
                      count(DirectiveAbbreviation,CountryName)

## merge the mean score and the count of components
goal_spec_compli_mean = goal_spec_compli_mean %>%
                        left_join(., goal_spec_compli_n2, 
                                  by= c("DirectiveAbbreviation","CountryName"))
               

dim(goal_spec_compli_mean)
```

#### 1.2.5.8 Plot Mean Compliance score
**Max compliance score can be 3**
```{r plot mean compliance score}
ggplot(goal_spec_compli_mean)+
  geom_point(aes(CountryName,compli_mean,size=n))+
  facet_wrap(~DirectiveAbbreviation)+
  ylim(0,4)+
  ylab("Mean compliance score")+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Mean Compliance Score")

```


#### 1.2.5.9 Existence score
```{r clean exist score}
dim(goal_spec_exist)#818   7

goal_spec_exist = goal_spec_exist %>%
                  select(DirectiveAbbreviation,CountryName,exist_numeric,ExistenceLevel) %>%
                  distinct(.)
                  
dim(goal_spec_exist)#149   4

head(goal_spec_exist)

```


#### 1.2.5.10 Overall score
Max score is 4 (1 for existence, 3 for compliance).  

All directives are evaluated based on a maximum score of 4. Therefore, for directives where only existence can be measured (no compliance metrics), the maximum score possible is 1/4. These directives are: *IED, MSPD, CBD, CITES, COP21*.  Directives that are scored on both existence and compliance are: *BIRDS, BSAP, BWD, CFP, HD, HELCOM, MSFD, ND, NEC, POP, REACH, UWWTD, WFD*.  

In two directives scored for compliance, there was a country where compliance could not be scored. For UWWTD, no compliance report was available for Poland. For WFD, no compliance report was available for Denmark. However, these two countries are still evaluated on both existence and compliance so that their score is comparable to other countries for these directives.  

```{r overal score}

## merge existence and compliance
goal_spec_overall = full_join(goal_spec_exist, goal_spec_compli_mean,
                              by=c("DirectiveAbbreviation","CountryName")) %>%
                    mutate(compli_mean2 = ifelse(is.na(compli_mean),0,compli_mean)) %>% ## make a column where NA are zero so that it does not screw up the total score
                    mutate(total = exist_numeric + compli_mean2, 
                           max_score =4)

## This code not necessary if all directives get a max score of 4. 
# ## for UWWTD, Poland and WFD, Denmark, change max score to 4 so that are comparable
# goal_spec_overall = goal_spec_overall %>%
#                     mutate(max_score = ifelse(CountryName=="Poland" & DirectiveAbbreviation =="UWWTD",4,max_score),
#                            max_score = ifelse(CountryName=="Denmark" & DirectiveAbbreviation =="WFD",4,max_score))
# 


## calculate overall score
goal_spec_overall = goal_spec_overall %>%
                    mutate(overall_score = total/max_score) %>% 
                    mutate(score_level = ifelse(DirectiveAbbreviation %in% c("IED","MSPD","CBD","CITES","COP21"), "existence", "existence and compliance")) %>% ## score level defined by directives scored only for existence or for both
  select(DirectiveAbbreviation, CountryName,overall_score, score_level)

## change Helcom to HELCOM

goal_spec_overall = goal_spec_overall %>%
                    mutate(DirectiveAbbreviation = as.character(DirectiveAbbreviation))%>%
                    mutate(DirectiveAbbreviation = ifelse(DirectiveAbbreviation == "Helcom", "HELCOM", DirectiveAbbreviation))

```   

#### 1.2.5.11 Plot Overall score
```{r plot overall score}
ggplot(goal_spec_overall)+
  geom_point(aes(CountryName,overall_score, colour=score_level))+
  facet_wrap(~DirectiveAbbreviation)+
  ylim(0,1)+
   theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Overall Score per Regulation")

```

#### 1.2.5.12 Plot Overall score by country across all directives
```{r plot overall score by country}
ggplot(goal_spec_overall)+
  geom_point(aes(DirectiveAbbreviation,overall_score, colour=score_level))+
  facet_wrap(~CountryName)+
  ylim(0,1)+
   theme(axis.text.x = element_text(colour="grey20", size=6, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Overall Score per Regulation")

```

### 1.2.6 Exploring alternative mapping and weighting outcomes
**Mapping**  
Mapping the directives to goals can be done with different criteria. We explore 2 alternative mapping critera.  
(1) Direct effect of directive on goal status - expert opinion  
(2) Direct or indirect effect directive on goal status - expert opinion and key word search of the directives  

*Note* for this exploration this is done at the subgoal level. NUT, CON, TRA are treated as subgoals (not subcomponents) here.  


**Weights**  
Weights among different directives can be set using different criteria.  We calculate *G* in the resilience equation for two different mapping criteria.  

G = sum(w_i * G_i) / sum(wi_i) ;  G_i = specific regulatory dataset, w_i = weight for each i dataset used to assess G  

In OHI, weights are based on "quality of information contained in the dataset with regard to estimates of regulation effectiveness." In BHI, we are modifying this approach. Instead w_i are based on the overall importance and policy focus among the different directives.


#### 1.2.6.1 Load mapping and weighting layers
```{r load mapping and weighting layers}
map_direct = read.csv(file.path(dir_res, 'resilience_matrix_direct.csv'), stringsAsFactors = FALSE)

map_indirect_direct = read.csv(file.path(dir_res, 'resilience_matrix_indirect_direct.csv'), stringsAsFactors = FALSE)


weights_importance = read.csv(file.path(dir_res, 'resilience_weights_directive_importance.csv'))


```

#### 1.2.6.2 Calculate *G* for each country for direct mapping
```{r direct mapping G}
## vector of unique regulations
unique_regs = map_direct %>%
              select(WFD:IED)%>%
              colnames()
unique_regs

## long data format for mapping with columns goal, regulation, map
map_direct_long = map_direct %>%
                  select(-component,-spp_status,-wgi_all)%>% ## remove component column, remove environment and social resilience
                  gather(regulation,map, -goal) %>%
                  filter(map %in% unique_regs) %>%
                  arrange(goal)
  
                  
head(map_direct_long)

## clean weights_importance object
weights_importance = weights_importance %>%
                      filter(layer != 'spp_status')%>%
                      filter(layer != 'wgi_all' )


## join mapping to weights
map_direct_weights = left_join(map_direct_long, weights_importance,
                        by=c("regulation"="layer"))


## join country scoring to mapping and weights
map_direct_weights_score = inner_join(goal_spec_overall, map_direct_weights,
                                     by=c("DirectiveAbbreviation"="regulation"))%>%
                           dplyr::rename(regulation = DirectiveAbbreviation,
                                         country= CountryName)%>%
                           select(-score_level) %>%
                           arrange(country,goal)
## calculate country G

country_G_direct = map_direct_weights_score %>%
                   select(-map,-type)%>%
                   mutate(wi_Gi = weight*overall_score)%>%
                   group_by(country,goal)%>%
                   summarise(G = sum(wi_Gi)/sum(weight))%>%
                   ungroup()
              
## plot country G direct 
ggplot(country_G_direct)+
  geom_point(aes(goal,G))+
  facet_wrap(~country)+
  ylim(0,1)+
  theme(axis.text.x = element_text(colour="grey20", size=6, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Country G value (overall resilience) per Goal with direct mapping")

ggplot(country_G_direct)+
  geom_point(aes(country,G))+
  facet_wrap(~goal)+
  ylim(0,1)+
  theme(axis.text.x = element_text(colour="grey20", size=6, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Country G value (overall resilience) per Goal with direct mapping")


```

#### 1.2.6.3 Calculate *G* for each country for direct and indirect mapping
```{r calcuate G indirect and direct}
## vector of unique regulations
unique_regs = map_indirect_direct %>%
              select(WFD:IED)%>%
              colnames()
unique_regs

## long data format for mapping with columns goal, regulation, map
map_indirect_direct_long = map_indirect_direct %>%
                  select(-component,-spp_status,-wgi_all)%>% ## remove component column, remove environment and social resilience
                  gather(regulation,map, -goal) %>%
                  filter(map %in% unique_regs) %>%
                  arrange(goal)
  
                  
head(map_indirect_direct_long)

## clean weights_importance object
weights_importance = weights_importance %>%
                      filter(layer != 'spp_status')%>%
                      filter(layer != 'wgi_all' )


## join mapping to weights
map_indirect_direct_weights = left_join(map_indirect_direct_long, weights_importance,
                        by=c("regulation"="layer"))


## join country scoring to mapping and weights
map_indirect_direct_weights_score = inner_join(goal_spec_overall, map_indirect_direct_weights,
                                     by=c("DirectiveAbbreviation"="regulation"))%>%
                           dplyr::rename(regulation = DirectiveAbbreviation,
                                         country= CountryName)%>%
                           select(-score_level) %>%
                           arrange(country,goal)
## calculate country G

country_G_indirect_direct = map_indirect_direct_weights_score %>%
                   select(-map,-type)%>%
                   mutate(wi_Gi = weight*overall_score)%>%
                   group_by(country,goal)%>%
                   summarise(G = sum(wi_Gi)/sum(weight))%>%
                   ungroup()
              
## plot country G direct 
ggplot(country_G_indirect_direct)+
  geom_point(aes(goal,G))+
  facet_wrap(~country)+
  ylim(0,1)+
  theme(axis.text.x = element_text(colour="grey20", size=6, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Country G value (overall resilience) per Goal w/ indirect + direct mapping")

ggplot(country_G_direct)+
  geom_point(aes(country,G))+
  facet_wrap(~goal)+
  ylim(0,1)+
  theme(axis.text.x = element_text(colour="grey20", size=6, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Country G value (overall resilience) per Goal w/ indirect +direct mapping")

```

#### 1.2.6.4 Plot comparison of *G* by mapping approach
```{r G comparison}

country_G_compare = bind_rows(
                          mutate(country_G_direct,mapping="direct"),
                          mutate(country_G_indirect_direct,mapping= "indirect and direct"))

## plot country G comparison 
ggplot(country_G_compare)+
  geom_point(aes(goal,G, color=mapping, shape=mapping))+
  facet_wrap(~country)+
  ylim(0,1)+
  scale_shape_manual(values=c(16,6))+
  theme(axis.text.x = element_text(colour="grey20", size=6, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Country G Mapping Comparison")

## plot country G direct 
ggplot(country_G_compare)+
  geom_point(aes(country,G, colour=mapping,shape=mapping))+
  facet_wrap(~goal)+
  ylim(0,1)+
   scale_shape_manual(values=c(16,6))+
  theme(axis.text.x = element_text(colour="grey20", size=6, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Country G Mapping Comparison")


```


## TO DO
1. Decide on weight and mapping
2. Prepare final data layers (scores for country applied to BHI regions)




## 2. Social
The [Worldwide Governance Indicators (WGI) project](http://info.worldbank.org/governance/wgi/index.aspx#home) reports aggregate and individual governance indicators for 215 economies over the period 1996–2014, for six dimensions of governance: 

Voice and Accountability  
Political Stability and Absence of Violence/Terrorism  
Government Effectiveness  
Regulatory Quality  
Rule of Law  
Control of Corruption  

### 2.1 Data  
OHI has already extracted these data by country.  For BHI, we obtained these data on 13 July 
2016 from [the file 'wgi_combined_scores_by_country.csv' on the OHI github site](https://github.com/OHI-Science/ohiprep/tree/master/globalprep/worldbank_wgi/intermediate).

#### 2.1.1 Data attributes  
Colname names in 'wgi_combined_scores_by_country.csv'
    "country": reporting country
    "year": reporting year
    "score_wgi_scale": mean score of six indicators:
        Voice and Accountability
        Political Stability and Absence of Violence/Terrorism
        Government Effectiveness
        Regulatory Quality
        Rule of Law
        Control of Corruption
    "score_ohi_scale": score_wgi_scale rescaled between 0 and 1 (score_wgi_scale range was -
2.5 and 2.5)


### 2.2 Resilience layer 
The rescaled value of the WGI score is a resilience data layer (eg. higher WGI, greater resilience).

#### 2.2.1 Current value
The value for each country in 2013  

#### 2.2.2 Rescaling
The WGI scores have a range -2.5 to 2.5. They need to be rescaled to between 0 and 1. Use minimum and maximum value in the dataset.  We compare the pressure layer difference if all countries are included in the dataset when selecting min and max to rescale or if only Baltic region values are used obtain min and max values to rescale.    

### 2.3 Prepare Data layer

#### 2.3.1 Read in and organize data

##### 2.3.1.1 Read in data
Data are in the folder 'pressures' under 'wgi_social/data_database'
```{r read in data}
## read in data
wgi_global = read.csv(file.path(dir_wgi, 'data_database/wgi_combined_scores_by_country.csv'), sep=";", stringsAsFactors = FALSE)
str(wgi_global)

bhi_lookup = read.csv(file.path(dir_wgi, 'bhi_basin_country_lookup.csv'), sep=";", stringsAsFactors = FALSE) %>%
            select(BHI_ID, rgn_nam) %>%
            dplyr::rename(rgn_id=BHI_ID,
                          country=rgn_nam)


```

##### 2.3.1.2 Select Baltic Sea countries
```{r select baltic countries}

wgi_baltic = wgi_global %>%
            filter(grepl("Denmark|Estonia|Finland|Germany|Latvia|Lithuania|Poland|Sweden|Russia",country))

wgi_baltic %>% select(country) %>% distinct()
```


####  2.3.2 Plot WGI scores

#####  2.3.2.1  Plot WGI scores (WGI scale)
```{r Plot raw Baltic WGI scores}
ggplot(wgi_baltic) + 
  geom_point(aes(year, score_wgi_scale, colour = country ))+
   geom_line(aes(year, score_wgi_scale, colour = country ))+
  ylab("WGI Score")+
  ylim(-2.5,2.5)+
  ggtitle("WGI Scores (WGI scale")

```


#####  2.3.2.2  Plot WGI scores  - OHI global rescale
```{r Plot Baltic WGI ohi global rescale scores}
ggplot(wgi_baltic) + 
  geom_point(aes(year, score_ohi_scale, colour = country ))+
   geom_line(aes(year, score_ohi_scale, colour = country ))+
  ylab("Score")+
  ylim(0,1)+
  ggtitle("WGI Scores (OHI global rescale")

```


#### 2.3.3 Rescale to the Baltic region

##### 2.3.3.1 Select Min and Max
```{r select min and max from the Baltic}
## min
baltic_min = wgi_baltic %>%
             select(score_wgi_scale)%>%
             min()%>%
             as.numeric()
baltic_min

## country and year of min
filter(wgi_baltic, score_wgi_scale == baltic_min)
#         country year score_wgi_scale score_ohi_scale
# Russian Federation 2000      -0.8633578       0.3273284


##max
baltic_max = wgi_baltic %>%
             select(score_wgi_scale)%>%
             max()%>%
             as.numeric()

## country and year of max
filter(wgi_baltic, score_wgi_scale == baltic_max)

# country year score_wgi_scale score_ohi_scale
#Finland 2004        1.985394       0.8970788

```

##### 2.3.3.2 Rescale Baltic
```{r rescale baltic}
wgi_baltic = wgi_baltic %>%
             mutate(min = baltic_min,
                    max=baltic_max,
             score_bhi_scale = (score_wgi_scale - baltic_min)/(baltic_max - baltic_min))%>%
             select(-min,-max)


```


##### 2.3.3.3 Plot rescaled to Baltic Region
```{r plot rescaled to baltic }
ggplot(wgi_baltic) + 
  geom_point(aes(year, score_bhi_scale, colour = country ))+
   geom_line(aes(year, score_bhi_scale, colour = country ))+
  ylab("Score")+
  ylim(0,1)+
  ggtitle("WGI Scores (BHI region rescale")


```


##### 2.3.3.4 Plot and Compare Rescaling options
Using the BHI region rescaling penalizes Russia. Does also spread the scores somewhat.  

```{r Plot and Compare Rescaling options}
ggplot(gather(wgi_baltic, score_type, score, -country,-year) %>% filter(score_type != "score_wgi_scale")) + 
  geom_point(aes(year, score, colour = country ))+
   geom_line(aes(year, score, colour = country ))+
  facet_wrap(~score_type)+
  ylab("Score")+
  ylim(0,1)+
  ggtitle("Rescale comparison WGI Scores")

```

#### 2.3.4 Assign country scores to BHI regions

##### 2.3.4.1 Join BHI region lookup to country scores
```{r join BHI region lookup to country scores}
wgi_rgn = wgi_baltic %>%
          select(-score_wgi_scale)%>%
          gather(., score_type, resilience_score, -country,-year)%>%
          mutate(country = ifelse(country == "Russian Federation", "Russia", country))%>%
          full_join(., bhi_lookup, by = "country")

```


#### 2.3.5 Prepare and save Objects
Select global rescaling so that Russia is not so heavily penalized.  

##### 2.3.5.1 Prepare object
```{r prepare object}

wgi_all = wgi_rgn %>%
         filter(score_type == "score_ohi_scale")%>%
         select(-score_type, -country)%>%
         group_by(rgn_id) %>%
         filter(year == max(year))%>%
         ungroup() %>%
         select(rgn_id, year, resilience_score) %>%
         arrange(rgn_id)

wgi_all = wgi_all %>%
        select(-year)
        

```

#### 5.2 Save object
```{r save object}

write.csv(wgi_all, file.path(dir_layers, 'wgi_all_bhi2015.csv'), row.names=FALSE)
```


