---
title: "resilience_prep"
output: github_document
params: 
    datasource: csv
---

# Preparation of Resilience Data Layers

```{r setup}

## Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')

## rprojroot
root <- rprojroot::is_rstudio_project

## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)


dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')


# root$find_file("README.md")
# 
# root$find_file("ao_need_gl2014.csv")
# 
# root <- find_root_file("install_ohicore.r", 
# 
# withr::with_dir(
#   root_file("DESCRIPTION"))

source('~/github/bhi/baltic2015/prep/common.r')
dir_res    = file.path(dir_prep, 'resilience')

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_res, 'resilience_prep.rmd') 
```

## 1. Ecological 
## 1.1 Biological Integrity


## 1.2 Goal-specific Regulations

### 1.2.1 Background
This component of resilience captures the capacity of pressures to be offset by regulations/laws.  

There are three components to goal-specific regulation resilience: *(1) existence, (2) compliance, (3) enforcement*  

Three international treaties are assessed for all BHI countries for *existence* only.These agreements are´:  
[Convention on Biodiversity (CBD)](https://www.cbd.int)  
[Convention on International Trade in Endangered Species of Wild Fauna and Flora (CITES)](https://www.cites.org)  
[Paris Climate Agreement (COP21)](https://treaties.un.org/pages/ViewDetails.aspx?src=TREATY&mtdsg_no=XXVII-7-d&chapter=27&lang=en)

Two Baltic Region agreements will be assessed for *existence* and *compliance* for all countries. These are: Baltic Sea Action Plan (BSAP); Helsinki Convention (HELCOM).

13 EU directives will be assessed for *existence* and *compliance* for BHI countries that are EU member states. These are: Water Framework Directive (WFD), Marine Spatial Planning Directive (MSPD), Marine Strategy Framework Directive (MSFD), Habitat Directive (HD), Common Fisheries Policy (CFP), National Emmissions Ceilings (NEC), Industrial Emissions Directive (IED), Urban Waste Water Treatment Directive (UWWTD), Bathing Water Directive (BWD), Conservation of Wild Birds (BIRDS), Registration, Evaluation, Authorisation and Restriction of Chemicals (REACH), Peristent Organic Pollutants (POP), Nitrates Directive (ND).

### 1.2.2 Scoring
By country for each level assessed for each regulation  

#### 1.2.2.1 Existence
Scored as yes (1) or no (0)  
Score_Existence = Score / Max
Max = 1

#### 1.2.2.2 Compliance
Fail (0), Partial (1), Full (3)  
Not equal increments because Full compliance rewarded more here.

Country_compliance_directive = Mean (compliance score)_all_compliance_components

#### 1.2.2.3 Total regulation score
Score_Existence + Country_compliance_directive / (Possible_Existence + Possible_Compliance)

Compliance is weighted more heavily than existence when both are scored.  

### 1.2.3 Goal Weighting & Mapping
Each regulation will be weighted (1 or 2) for each goal to which is it applied. Regulations are mapped to all the goals to which they directly influence/combat pressures.

### 1.2.4 Resilience Data Layer
Country-specific resilience score for each goal is assigned to each BHI region associated with that country.  

Country_resilience_goal_x = Sum overall all regulations for a goal[country_regulation_score * goal_weight]  

Country_resilience_goal_x  - rescale between 0-1?  


### 1.2.5 Goal-specific data layer preparation

#### 1.2.5.1 Read in data
```{r read in goal specific reg data}
## read in data

eu_baltic_scores = read.csv(file.path(dir_res, 'eu_regs_resilience_data.csv'), sep=";")

international_scores = read.csv(file.path(dir_res, 'InternationalAgreements_resilience_data.csv'), sep=";")


## 
str(eu_baltic_scores)
dim(eu_baltic_scores)

str(international_scores)
dim(international_scores)
```

#### 1.2.5.2 simply data objectives, combine into single object
```{r simplfy}
eu_baltic_lookup = eu_baltic_scores ## keep the info of text references in case needed
internat_lookup = international_scores


eu_baltic_scores = eu_baltic_scores %>%
                  mutate(Existence = "Yes",
                         ExistenceLevel = "EU_Law",
                         ExistenceNotes = "")%>% ## add existence so that compatible with international ## all EU and Baltic directives / agreements exist (if at EU level are at country level)
                  select(-DirectiveName,-ComplianceText,-ComplianceTextReference,-Notes_UncertainitiesInECreport,-Notes_AreasForFurtherInvestigation)%>%
                  select(DirectiveAbbreviation,ReportYear,CountryName,Existence,ExistenceLevel,ExistenceNotes,IsCountryAssessed,ComplianceComponent, ComplianceComponentScore,NAexplanation,Comments) %>% ## reorder
                      dplyr::rename(NAexplanationCompliance = NAexplanation,
                                    IsCountryComplianceAssessed = IsCountryAssessed)

colnames(eu_baltic_scores)

international_scores = international_scores %>%
                      mutate(IsCountryAssessed = "No")  %>% ## this is for compliance
                        select(-DirectiveName,-DataLink,-Notes_UncertainitiesInECreport,-Notes_AreasForFurtherInvestigation) %>%
                  select(DirectiveAbbreviation,ReportYear,CountryName,Existence,ExistenceLevel,ExistenceNotes,IsCountryAssessed,ComplianceComponent, ComplianceComponentScore,NAexplanation,Comments) %>%## reorder
                      dplyr::rename(NAexplanationCompliance = NAexplanation,
                                    IsCountryComplianceAssessed = IsCountryAssessed)

colnames(international_scores)

goal_spec = rbind(eu_baltic_scores, international_scores)



```


#### 1.2.5.3 assign numeric score

```{r numeric score}
goal_spec = goal_spec %>% 
            mutate(exist_numeric = ifelse(Existence == "Yes",1,0),
                   compli_numeric = ifelse(is.na(ComplianceComponentScore),NA,
                                    ifelse(ComplianceComponentScore == "Not applicable", NA,
                                    ifelse(ComplianceComponentScore == "fail",0,
                                    ifelse(ComplianceComponentScore == "partial",1,
                                    ifelse(ComplianceComponentScore == "full",3,NA))))))
            


```

#### 1.2.5.4 Plot numeric score
With the numeric score for Compliance Components, can not see overlapping scores for different components
```{r plot raw score}

ggplot(goal_spec) + 
  geom_point(aes(CountryName,exist_numeric))+
  facet_wrap(~DirectiveAbbreviation)+
  ylim(0,1)+
   theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Existence Score")

ggplot(goal_spec) + 
  geom_point(aes(CountryName,compli_numeric))+
  facet_wrap(~DirectiveAbbreviation)+
  ylim(0,1)+
   theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Compliance Score")

```

#### 1.2.5.5 Separate Existance and Compliance
```{r seperate objects for exist and compli}

goal_spec_exist = goal_spec %>%
                  select(DirectiveAbbreviation:ExistenceNotes, exist_numeric)

goal_spec_compli = goal_spec %>%
                  select(DirectiveAbbreviation:CountryName, IsCountryComplianceAssessed:Comments, compli_numeric)


```

#### 1.2.5.6 How many compliance components assessed?
Some directives have no components assessed. Others have a subset that could not be assessed - these might differ by country. Minimal differences by country. WDF not assessed for Denmark. Germany assessed for one more POP component. Poland asssesse for one fewer UWWTD component.  
**Gray bars are components not assessed**
```{r}

goal_spec_compli_n1 = goal_spec_compli %>%
                      count(DirectiveAbbreviation, CountryName,compli_numeric)

ggplot(goal_spec_compli, aes(CountryName,fill=as.character(compli_numeric)))+ 
  geom_bar(position="stack")+
  facet_wrap(~DirectiveAbbreviation)+
  ylab("Count of Compliance Componenets")+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Compliance Components Assessed by Country & Directive")

```


#### 1.2.5.7 Mean Compliance score
```{r mean compliance score}
## calculate the mean value for all scored compliance by directive and country
goal_spec_compli_mean = goal_spec_compli%>%
            select(DirectiveAbbreviation,CountryName,compli_numeric) %>%
            group_by(DirectiveAbbreviation,CountryName)%>%          
            summarise(compli_mean = mean(compli_numeric, na.rm=TRUE)) %>%
            ungroup()

                      
## get the number of components assessed
goal_spec_compli_n2 = goal_spec_compli%>%
                      select(DirectiveAbbreviation,CountryName,compli_numeric) %>%
                      filter(!is.na(compli_numeric))%>%
                      count(DirectiveAbbreviation,CountryName)

## merge the mean score and the count of components
goal_spec_compli_mean = goal_spec_compli_mean %>%
                        left_join(., goal_spec_compli_n2, 
                                  by= c("DirectiveAbbreviation","CountryName"))
               

dim(goal_spec_compli_mean)
```

#### 1.2.5.8 Plot Mean Compliance score
**Max compliance score can be 3**
```{r plot mean compliance score}
ggplot(goal_spec_compli_mean)+
  geom_point(aes(CountryName,compli_mean,size=n))+
  facet_wrap(~DirectiveAbbreviation)+
  ylim(0,4)+
  ylab("Mean compliance score")+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Mean Compliance Score")

```


#### 1.2.5.9 Existence score
```{r clean exist score}
dim(goal_spec_exist)#818   7

goal_spec_exist = goal_spec_exist %>%
                  select(DirectiveAbbreviation,CountryName,exist_numeric,ExistenceLevel) %>%
                  distinct(.)
                  
dim(goal_spec_exist)#149   4

head(goal_spec_exist)

```


#### 1.2.5.10 Overall score
Max score is 4 (1 for existence, 3 for compliance).  

Directives that can only be scored on existence are evaluated on only existence a max score of 1. These directives are: IED, MSPD, CBD, CITES, COP21.  

Directives that are scored on both existence and compliance are evaluated on the combined score with a mx of 4 possible. These directives are: BIRDS, BSAP, BWD, CFP, HD, HELCOM, MSFD, ND, NEC, POP, REACH, UWWTD, WFD.  

In two directives scored for compliance, there was a country where compliance could not be scored. For UWWTD, no compliance report was available for Poland. For WFD, no compliance report was available for Denmark. However, these two countries are still evaluated on both existence and compliance so that their score is comparable to other countries for these directives.  

```{r overal score}

## merge existence and compliance
goal_spec_overall = full_join(goal_spec_exist, goal_spec_compli_mean,
                              by=c("DirectiveAbbreviation","CountryName")) %>%
                    mutate(compli_mean2 = ifelse(is.na(compli_mean),0,compli_mean)) %>% ## make a column where NA are zero so that it does not screw up the total score
                    mutate(total = exist_numeric + compli_mean2, 
                           max_score = ifelse(!is.na(compli_mean),4,1))

## for UWWTD, Poland and WFD, Denmark, change max score to 4 so that are comparable
goal_spec_overall = goal_spec_overall %>%
                    mutate(max_score = ifelse(CountryName=="Poland" & DirectiveAbbreviation =="UWWTD",4,max_score),
                           max_score = ifelse(CountryName=="Denmark" & DirectiveAbbreviation =="WFD",4,max_score))

## calculate overall score
goal_spec_overall = goal_spec_overall %>%
                    mutate(overall_score = total/max_score) %>% ## if compli_mean is not NA, the max total score is 4; if it is NA, then only existence assessend and max is 1
                    mutate(score_level = ifelse(max_score == 1, "existence", "existence and compliance")) %>%
  select(DirectiveAbbreviation, CountryName,overall_score, score_level)

```   

#### 1.2.5.11 Plot Overall score
```{r plot overall score}
ggplot(goal_spec_overall)+
  geom_point(aes(CountryName,overall_score, colour=score_level))+
  facet_wrap(~DirectiveAbbreviation)+
  ylim(0,1)+
   theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Overall Score per Regulation")

```

#### 1.2.5.12 Plot Overall score by country across all directives
```{r plot overall score by country}
ggplot(goal_spec_overall)+
  geom_point(aes(DirectiveAbbreviation,overall_score, colour=score_level))+
  facet_wrap(~CountryName)+
  ylim(0,1)+
   theme(axis.text.x = element_text(colour="grey20", size=6, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Overall Score per Regulation")

```


## TO DO
1. Weight each regulation for each goal  
2. Map each regulation to goals  
3. Prepare final data layers (scores for country applied to BHI regions)


## 2. Social
