---
title: "open_sea_anoxia_prep"
output:
  github_document:
    toc: true
    toc_depth: 3
params: 
    datasource: csv
---
```{r setup}

## Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')
library(raster)

# source
source('~/github/bhi/baltic2015/prep/common.r') #for create_readme() function

## rprojroot
root <- rprojroot::is_rstudio_project


## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)

dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')
dir_prep = make_path('baltic2015/prep') # replaces  file.path(dir_baltic, 'layers')


# root$find_file("README.md")
# 
# root$find_file("ao_need_gl2014.csv")
# 
# root <- find_root_file("install_ohicore.r", 
# 
# withr::with_dir(
#   root_file("DESCRIPTION"))

dir_anox   = file.path(dir_prep, 'pressures/open_sea_anoxia')

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_anox, 'open_sea_anoxia_prep.rmd') 
```

## 1. Background on open sea anoxia data  

### 1.1 Why anoxia?


## 2. Anoxia data  

### 2.1 Data source 
#### 2.1.1 Anoxia data layer
These data are from [Carstensen et al. 2014](http://www.pnas.org/content/111/15/5628.abstract). Jacob Carstensen provided dbf files containing O2 (mg⋅L−1) for a series of years for the Baltic Proper and the Gulf of Finland.  

Other areas are not included in this study but at the present time have no anoxia problems - therefore they will recieve a pressure value of 0.  

``` {r anoxia data setup, echo = FALSE, message = FALSE}
## directory setup
dir_M <- c('Windows' = '//mazu.nceas.ucsb.edu/ohi',
           'Darwin'  = '/Volumes/ohi',    ### connect (cmd-K) to smb://mazu/ohi
           'Linux'   = '/home/shares/ohi')[[ Sys.info()[['sysname']] ]]

if (Sys.info()[['sysname']] != 'Linux' & !file.exists(dir_M)){
  warning(sprintf("The Mazu directory dir_M set in src/R/common.R does not exist. Do you need to mount Mazu: %s?", dir_M))
  # TODO: @jennifergriffiths reset dir_M to be where the SRC hold your shapefiles
}

dir_shp <- file.path(dir_M, 'git-annex/Baltic')

## libraries
library(dplyr)
library(rgdal)  # install.packages('rgdal')
library(raster) # install.packages('raster')
library(rgeos)  # install.packages('rgeos')

```

#### 2.1.2 Bathymetry
Downloaded from [HELCOM](http://maps.helcom.fi/website/mapservice/index.html) 28 April 2016 by Thorsten Blenckner

``` {r read in bathymetry data, echo = FALSE}

## read in bathymetry raster data
bathymetry = raster::raster(file.path(dir_M, 'git-annex/Baltic/anoxia/Baltic_Sea_Bathymetry_Database_v091.tif'))
# anox
# plot(bathymetry)

```

## 3. Open sea anoxia spatial data scale
We will work on the BHI region scale. O2 rasters and Batlic bathymetry will be overlayed on BHI regions.  Different approaches will be used for BHI regions in the Baltic Proper and in the Gulf of Finland.  

In "baltic_rgns_to_bhi_rgns_lookup_holas_basins.csv" in "~\github\bhi\baltic2015\prep\pressures\open_sea_anoxia" the BHI regions associated with the Baltic Proper and Gulf of Finland are listed.

``` {r read in anoxia data, echo = FALSE}


  # install.packages('foreign')
  # library(foreign)

anox_file_list = list.files(path= file.path(dir_shp, 'anoxia'),
                            pattern="CodRVmap_")

  bal_proj = CRS("+init=epsg:32634 +proj=utm +zone=34 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0 ")
  bal_res = 1000
  dir_temp = make_path('baltic2015/prep/pressures/open_sea_anoxia/O2_bottom_files')

for (anox_file in anox_file_list) {
   
  # anox_file = 'CodRVmap_1906.dbf'
  anox = foreign::read.dbf(file.path(dir_shp, 'anoxia', anox_file))  %>%
       dplyr::select(x_utm, y_utm, O2_bottom) # head(anox) 

  t <- raster::rasterFromXYZ( anox, res = bal_res, crs = bal_proj) 
  # plot(t)
  
  rast_file = file.path(dir_temp, paste0(file_path_sans_ext(anox_file), ".tif"))
  writeRaster(t, rast_file,  overwrite=TRUE)
  
}


```

## 4. Open sea anoxia data rescaling
Note, we may need to treat the Baltic Proper and the Gulf of Finland separately.  In the Baltic Proper there is a strong halocline at 70m and it is the anoxic area below the halocline that is of interest (therefore we focus on areas with depths >= 70).  In the Gulf of Finland, there a less strong halocline and thermal stratification is important so we may have to think differently about which areas to focus on.  

```{r anoxia rescaling, echo = FALSE}

# isolate Baltif Proper and Gulf of Finland

## inspect BHI regions ----
bhi <- rgdal::readOGR(dsn = path.expand(file.path(dir_shp, 'BHI_MCG_shapefile')),
                      layer = 'BHI_MCG_11052016') 

## subset Baltic Proper (BP) and Gulf of Finland (GoF) regions, and rasterize them to the same coor system and resolution as O2_bottom files. 

O2_raster = raster::raster(file.path(dir_anox, 'O2_bottom_files', 'CodRVmap_2014.tif'))

GoF = bhi %>%
   subset(Subbasin == "Gulf of Finland") %>%
  spTransform(bal_proj) %>%
  rasterize(O2_raster, progress='text', field = 'BHI_ID') #plot(GoF)

BP = bhi %>% 
  subset(!Subbasin == "Gulf of Finland") %>%
  spTransform(bal_proj) %>%
  rasterize(O2_raster, progress='text', field = 'BHI_ID') #plot(BP)

# create bathymetry mask: remove cells with depth higher than -70 m

halocline_70 = bathymetry 
halocline_70[halocline_70 > -70] = NA

halocline_70 = projectRaster(halocline_70, O2_raster) # transform to the same projection as O2 data
plot(halocline_70)

# Loop for each O2_bottom tif file, mask with halocline_70

O2_file_list = list.files(path = file.path(dir_anox, 'O2_bottom_files'),
                          pattern ="CodRVmap_[0-9]+\\.tif")

for (i in O2_file_list){

## read in file 
## test: i = 'CodRVmap_2014.tif'
O2_bottom = raster::raster(file.path(dir_anox, 'O2_bottom_files', i))

## select area with O2<0 
O2_below_0 = O2_bottom
O2_below_0[O2_below_0 >= 0] = NA
# plot(O2_below_0)
# plot(halocline_70, add=T)

#### for 1906 and 1955, file coord system are different from bal_proj. To change them to the same coor system:  
if (i == 'CodRVmap_1906.tif' | i =='CodRVmap_1955.tif' ) {
  
  O2_below_0 = O2_below_0 %>% projectRaster(halocline_70)
  
}

## mask with depth >70m and BP 
O2_below_70 = mask(O2_below_0, halocline_70) %>%
  mask(BP)
# plot(O2_below_70)

## Calcualte the total area within each BP BHI region
# since we know cell resolution of O2_below_70 is 1000 x 1000 m, we can just count number of cells per region and sum

# rgn_ras <- bhi%>%
#             spTransform(bal_proj) %>% 
#               rasterize(O2_bottom, progress='text', field = 'BHI_ID') # to the same resolution as O2_bottom

# use zonal to get statistics per 'zone' - in this case we have rasterized our BHI polygon to 1000x1000m cells, with BHI_ID as cell values. This acts as a zone layer
# that can be used to extract different information (mean, max, min, count, sum) for rasters of the same res and crs.
area_rgn <- zonal(O2_below_70, BP, fun='count') %>%
  data.frame(.) %>%
  dplyr::select(BHI_ID = zone, 
         area_km2 = count) %>%
  mutate(year = substr(i, 10, 13)) # add year

write_csv(data.frame(area_rgn), file.path(dir_temp, paste0(file_path_sans_ext(i), '_BP.csv')))

}

```

Gulf of Finland: 

``` {r Gulf of Finland Anoxia rescaling, ECHO = FALSE, message = FALSE }

O2_file_list = list.files(path = file.path(dir_anox, 'O2_bottom_files'),
                          pattern ="CodRVmap_[0-9]+\\.tif")

for (i in O2_file_list){

## read in O2 file 
## test: i = 'CodRVmap_1906.tif'
O2_bottom = raster::raster(file.path(dir_anox, 'O2_bottom_files', i))

## select area with O2<0 
O2_below_0 = O2_bottom
O2_below_0[O2_below_0 >= 0] = NA
# plot(O2_below_0)

#### for 1906 and 1955, file coord system are different from the rest of the O2 files. To change them to the same coor system:  
if (i == 'CodRVmap_1906.tif' | i =='CodRVmap_1955.tif' ) {
  
  O2_below_0 = O2_below_0 %>% projectRaster(O2_raster)
  #plot(O2_below_0)
}

## mask with GoF
O2_below_0_GoF = mask(O2_below_0, GoF) 
# plot(O2_below_0_GoF)

## Calcualte the total area within each BP BHI region
# since we know cell resolution of O2_below_70 is 1000 x 1000 m, we can just count number of cells per region and sum. 
# use zonal to get statistics per 'zone' - in this case we have rasterized our BHI polygon to 1000x1000m cells, with BHI_ID as cell values. This acts as a zone layer
# that can be used to extract different information (mean, max, min, count, sum) for rasters of the same res and crs.

area_rgn <- zonal(O2_below_0_GoF, GoF, fun='count') %>%
  data.frame(.) %>%
  dplyr::select(BHI_ID = zone, 
         area_km2 = count) %>%
  mutate(year = substr(i, 10, 13)) # add year

write_csv(data.frame(area_rgn), file.path(dir_temp, paste0(file_path_sans_ext(i), '_GoF.csv')))

}


```

Combine Gulf of Finland and Baltic Proper data

``` {r combine GoF and BP data, ECHO = FALSE, message = FALSE}





```
### 4.1 Current value  
    - Baltic Proper BHI regions =  of the area >= 70m deep the area with O2 <0 mg⋅L−1  
        - Use the mean area for the most recent 3 years (2012-2014)  
    - Gulf of Finland BHI regions = of the total surface area in each BHI region, the area with O2 <0 mg⋅L−1  
        - Use the mean area for the most recent 3 years (2012-2014)  
        
``` {r Current value, ECHO= FALSE, message = FALSE}




```

### 4.2 Max value
    - Baltic Proper BHI regions =  of area >= 70m depth, the maximum extent of area O2 <0 mg⋅L−1 in last 10 years  
    - Gulf of Finland BHI regions = of the surface area, the maximum extent of area O2 <0 mg⋅L−1 in last 10 years  

### 4.3 Min value
     - Baltic Proper BHI regions =  of area >= 70m deep the area with O2 <0 mg⋅L−1  in 1906  
     - Gulf of Finland BHI regions = total surface area, the area with O2 <0 mg⋅L−1  in 1906


## 5. Open Sea Anoxia layer prep

```{r TODO, echo=FALSE, message=FALSE}
## read in data...

## plot data...

```

UTM 34 North. 
