---
title: "eco_prep"
output: github_document
params: 
    datasource: csv
---

# ECO subgoal data preparation

```{r setup}

## Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')

source('~/github/bhi/baltic2015/prep/common.r')


## rprojroot
root <- rprojroot::is_rstudio_project


## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)

dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')


# root$find_file("README.md")
# 
# root$find_file("ao_need_gl2014.csv")
# 
# root <- find_root_file("install_ohicore.r", 
# 
# withr::with_dir(
#   root_file("DESCRIPTION"))



dir_eco    = file.path(dir_prep, 'ECO')

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_eco, 'eco_prep.rmd')

```

## 1. Background

## 2. Data

### 2.1 Regional Data
Eurostat regional (NUTS3) GDP
Data downloaded on 12 May 2016 from Eurostat database [nama_10r_3gdp](http://ec.europa.eu/eurostat/data/database?p_auth=EgN81qAf&p_p_id=estatsearchportlet_WAR_estatsearchportlet&p_p_lifecycle=1&p_p_state=maximized&p_p_mode=view&_estatsearchportlet_WAR_estatsearchportlet_action=search&text=nama_10r_3gdp)  

Nominal GDP data in millions of Euros.

No data for Russia regions.  

### 2.2 Country Level data
Eurostat Country GDP data
Data downloaded 22 March 2016 from Eurostat Database [nama_10_gdp](http://ec.europa.eu/eurostat/data/database?p_auth=sHLAepWT&p_p_id=estatsearchportlet_WAR_estatsearchportlet&p_p_lifecycle=1&p_p_state=maximized&p_p_mode=view&_estatsearchportlet_WAR_estatsearchportlet_action=search&text=nama_10_gdp)  

Nominal GDP in millions of Euros.  

Data does not include Russia

### 2.3 Russian country level data

Eurostat database [naida_10_gdp](http://ec.europa.eu/eurostat/web/products-datasets/-/naida_10_gdp)  

[Metadata link](http://ec.europa.eu/eurostat/cache/metadata/en/naid_10_esms.htm)  

Download criteria:  
Russia; Gross domestic product at market prices; CP_MEUR (Current prices, million euro ); 2000-2014


### 2.4 Population density data

#### 2.4.1 Fine scale
Population density data obtained from the [HYDE database](http://themasites.pbl.nl/tridion/en/themasites/hyde/download/index-2.html)

Year of data = 2005. Data were a 5' resolution. Erik Smedberg with the Baltic Sea Center re-gridded to a 10 x 10 km grid.

Population density within a 25km buffer from the coast will be used.

References:
Klein Goldewijk, K. , A. Beusen, M. de Vos and G. van Drecht (2011). The HYDE 3.1 spatially explicit database of human induced land use change over the past 12,000 years, Global Ecology and Biogeography20(1): 73-86. DOI: 10.1111/j.1466-8238.2010.00587.x.

Klein Goldewijk, K. , A. Beusen, and P. Janssen (2010). Long term dynamic modeling of global population and built-up area in a spatially explicit way, HYDE 3 .1. The Holocene20(4):565-573. http://dx.doi.org/10.1177/0959683609356587

#### 2.4.2 National
**EU countries**  
Downloaded on March 31 2016 from Eurostat database [demo_gind](http://ec.europa.eu/eurostat/data/database?p_auth=whAQQAX7&p_p_id=estatsearchportlet_WAR_estatsearchportlet&p_p_lifecycle=1&p_p_state=maximized&p_p_mode=view&_estatsearchportlet_WAR_estatsearchportlet_action=search&text=demo_gind)  

Variables:  
1990-2015 
Population on Jan1  

[Metadata Link](http://ec.europa.eu/eurostat/cache/metadata/en/demo_gind_esms.htm)  

*Population on 1 January:*  
Eurostat aims at collecting from the EU-28's Member States' data on population on 31st December, which is further published as 1 January of the following year. The recommended definition is the 'usual resident population' and represents the number of inhabitants of a given area on 31st December . However, the population transmitted by the countries can also be either based on data from the most recent census adjusted by the components of population change produced since the last census, either based on population registers.


**Russia**  
Downloaded on 10 June 2016 from Eurostat database:   [naida_10_pe](http://ec.europa.eu/eurostat/web/products-datasets/-/naida_10_pe)  
population (thousands of people)  
employment (thousands of people) *need to exclude this*  
 

### 2.5 Aligning BHI regions with NUTS3 regions and population density

UPDATE with Marc's methods or link


## 3. Goal Model

'Xeco = (GDP_Region_c/GDP_Region_r)/(GDP_Country_c/GDP_Country_r)  
'c = current year, r=reference year  '
'reference point is a moving window (single year value)
'Region is the BHI region which is comprised of GDP data from the associated NUTS3 regions
'data can be in nominal GDP because is a ratio value (adjusting by a deflator would cancel out)  
'each BHI region is composed by one or more NUTS3 regions, these are allocated by population density from each NUTS3 region associated with a given BHI region




## 4. Other
NEED TO UPDATE - this may change
### 4.1 Interpreting NA and zero
#### 4.1.1 Status Score of Zero
'Status scores of zero were assigned when the region had no data or insufficient data but the indicator is applicable

#### 4.1.2 Trend value of NA
'Trend values of NA are assigned if there are not 5 years of data available to calculate a trend




## 5. Regional GDP prep

### 5.1 Data organization

#### 5.1.1 Read in data
This should be check and updated when data are updated in database. Data should be extract from database by ´eco_prep_database_call.r´ and saved to folder ´eco_data_database´

```{r read in regional gdp}
regional_gdp = read.csv(file.path(dir_eco, 'eco_data_database/nama_10r_3gdp_1_Data_download05-12-2016.csv'), sep=";")

dim(regional_gdp) #20760     6

str(regional_gdp)
```

#### 5.1.2 clean data object
```{r clean regional gdp object}
unique(regional_gdp$UNIT) ## Million euro 

regional_gdp = regional_gdp %>%
              dplyr::rename(year = TIME, nuts3 = GEO, nuts3_name = GEO_LABEL,
                            unit = UNIT, value = Value, flag_notes = Flag_and_Footnotes)%>%
              mutate(nuts3 = as.character(nuts3), 
                     nuts3_name = as.character(nuts3_name))


```

### 5.2 Data associations with Baltic and BHI
#### 5.2.1 Filter for Baltic countries
```{r filter for baltic countries}
## read in EU country abbreviations and names
eu_lookup = read.csv(file.path(dir_eco, 'EUcountrynames.csv'), sep=";")  

## add country names to regional_gdp and filter for Baltic countries
regional_gdp2 = regional_gdp %>%
               mutate(country_abb = substr(nuts3,1,2))%>%
               left_join(., eu_lookup, by="country_abb") %>%
               filter(grepl("Denmark|Estonia|Finland|Germany|Latvia|Lithuania|Poland|Sweden",country))
               
str(regional_gdp2)
unique(regional_gdp2$country)


```

#### 5.2.2 Filter for regions associated with BHI regions

#### 5.2.3 Plot regional GDP, divide by country
```{r plot regional gdp raw}



```


### 5.3 BHI region GDP

#### 5.3.1 Population density in 25km for each NUTS3 region

#### 5.3.2 Percent NUTS3 region buffer pop den per BHI region

#### 5.3.3 GDP allocated to BHI regions

#### 5.3.4 BHI region total GDP

#### 5.3.5 Select years




## 6.Country GDP prep

### 6.1 Organize data

#### 6.1.1 Read in data

```{r read in country gdp}
## read EU national GDP
country_gdp = read.csv(file.path(dir_eco, 'eco_data_database/nama_10_gdp_1_Data.csv'), sep=";")
 dim(country_gdp) #[1] 4428    8
 str(country_gdp)
 
## read in Russian national GDP
russian_nat_gdp = read.csv(file.path(dir_eco, 'eco_data_database/naida_10_gdp_1_Data_cleaned.csv'), sep=";")

dim(russian_nat_gdp) #16  6
str(russian_nat_gdp)

```

#### 6.1.2 Clean data
```{r clean country gdp data}
country_gdp = country_gdp %>%
              dplyr::rename(year = TIME, country_abb = GEO, country= GEO_LABEL,
                            unit=UNIT, na_item = NA_ITEM, na_item_label = NA_ITEM_LABEL,
                            value = Value, flag_notes = Flag.and.Footnotes) %>%
              mutate(country_abb = as.character(country_abb),
                     country = as.character(country))%>%
              mutate(country = ifelse(country =="Germany (until 1990 former territory of the FRG)","Germany",country ))

russian_nat_gdp = russian_nat_gdp %>%
                  dplyr::rename(year = TIME, country= GEO,
                            unit=UNIT, na_item = NA_ITEM,
                            value = Value, flag_notes = Flag.and.Footnotes) %>%
                            mutate(country_abb = "RU")

  


```

#### 6.1.3 Restrict years to >= 2000
For EU country GDP
```{r years to greater than or equal to 2000}
country_gdp = country_gdp %>%
              filter(year >=2000) %>%
               filter(grepl("Denmark|Estonia|Finland|Germany|Latvia|Lithuania|Poland|Sweden",country))
              

```

#### 6.1.4 Select unit type - 'current prices, million euro
For EU country GDP
```{r select unit type}

##EU country GDP
unique(country_gdp$unit)
##[1] Current prices, million euro              Chain linked volumes (2010), million euro
##[3] Chain linked volumes, index 2010=100

country_gdp = country_gdp %>%
              filter(unit == "Current prices, million euro")%>%
              mutate(unit = "million euro")

dim(country_gdp) #[1] 576   8

## Russia GDP, clean unit code
russian_nat_gdp = russian_nat_gdp %>%
                  mutate(unit = "million euro")

```

#### 6.1.5 assign data flag codes
```{r data flag codes}
country_gdp %>% select(flag_notes)%>%distinct() ## p e
country_gdp = country_gdp %>%
              mutate(flag_notes = ifelse(flag_notes == 'p', "provisional",
                                  ifelse(flag_notes == 'e', "estimated", "")))
country_gdp %>% filter(flag_notes == "provisional" | flag_notes == "estimated")  ## no baltic countries


russian_nat_gdp %>% select(flag_notes) %>% distinct() #e
russian_nat_gdp = russian_nat_gdp %>%
                  mutate(flag_notes = ifelse(flag_notes == 'e', "estimated", ""))
russian_nat_gdp %>% filter(flag_notes == "estimated")  ## two years estimated 2000, 2001




```

### 6.2 Baltic regions

#### 6.2.1 Filter for Baltic countries
```{r filter country gdp for Baltic countries}

country_gdp2 = country_gdp %>%
                filter(grepl("Denmark|Estonia|Finland|Germany|Latvia|Lithuania|Poland|Sweden",country))
dim(country_gdp2);dim(country_gdp)                

colnames(country_gdp2)
colnames(russian_nat_gdp)

```

#### 6.2.2 Bind rows EU Baltic countries and Russia data
```{r join eu and russia}

country_gdp3 = bind_rows(select(country_gdp2, year, country, country_abb, unit, value, flag_notes),
                    select(russian_nat_gdp, year, country, country_abb, unit, value, flag_notes))
str(country_gdp3)

```  

#### 6.2.3 Plot National GDP
```{r plot national GDP}
ggplot(country_gdp3)+
  geom_point(aes(year,value))+
  facet_wrap(~country, scales="free_y") +
  ggtitle("National Nominal GDP")

```

### 6.3 Join BHI region id to countries

### Population data - get per capita GDP

### 6.4 Create and export country GDP data layer
