---
title: "contaminants_prep"
output: github_document
params: 
    datasource: csv
---
```{r setup}

## source common libraries, directories, functions, etc
# Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')

## rprojroot
root <- rprojroot::is_rstudio_project


## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)

dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')

source('~/github/bhi/baltic2015/prep/common.r')
dir_cw    = file.path(dir_prep, 'CW')
dir_con    = file.path(dir_prep, 'CW/contaminants')

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_con, 'contaminants_prep.rmd')

```

# 1. Contaminant Data Prep

## 1.2 Indicators
3 indicators are proposed which would then be combined to give an overall comtanimant sub-component status.  

### 1.2.1 (1) PCB concentration indicator
**CB-153 concentration**
    - Suggested by Anders Bignet and Elisabeth Nyberg.  

This would be a good indicator because it is abundant (so almost always measured) and will have few detection / quantification limit problems.  

Use the EU threshold for human health as the reference point.  



#### 1.2.1a Original PCB indicator: ICES-6 PCB
Sum  of  PCB28, PCB52,  PCB101, PCB138,  PCB153 and  PCB180.  

This is similar to the ICES-7 except that PCB 118 is excluded (since it is metabolized by mammals).  

75 ng/g wet weight is the [EU threshold for fish muscle. See Section 5 Annex, 5.3](http://eur-lex.europa.eu/LexUriServ/LexUriServ.do?uri=OJ:L:2011:320:0018:0023:EN:PDF)  

**Additional information from HELCOM on the Core Indicators**  
[HELCOM Core Indicator of Hazardous Substances Polychlorinated biphenyls (PCB) and dioxins and furans 2013](http://www.helcom.fi/Core%20Indicators/HELCOM-CoreIndicator_Polychlorinated_biphenyls_and_dioxins_and_furans.pdf)

*Relevant text from the HELCOM core indicator report*  
"The PCBs included in this core indicator report are the 7 PCB congeners that have been monitored since the beginning of the HELCOM and OSPARCOM monitoring programmes, carefully selected mainly by ICES working groups due to their relatively uncomplicated identification and quantification in gas chromatograms and as they usually contribute a very high proportion of the total PCB content in environmental samples. These are the ‘ICES 7’: CB-28, CB-52, CB-101, CB-118, CB-138, CB-153 and CB-180."  

*Determination of GES boundary*
The CORESET expert group decided that, due to uncertainties in the target setting on the OSPAR and EU working groups, the seven PCBs should be monitored and concentrations analysed but the core indicator assesses primarily two congeners only: CB-118 (dioxin like) and 153 (non-dioxin like). Tentatively the OSPAR EACs for these two congeners are suggested to be used.


### 1.3 (2) TEQ value for PCBs and Dioxins
[HELCOM Core Indicator of Hazardous Substances Polychlorinated biphenyls (PCB) and dioxins and furans](http://www.helcom.fi/Core%20Indicators/HELCOM-CoreIndicator_Polychlorinated_biphenyls_and_dioxins_and_furans.pdf)  *There should be a more recent document*  

Dioxins are included in several international agreements, of which the Stockholm Convention and the Convention on Long Range Transboundary Air are among the most important for the control and reduction of sources to the environment. WHO and FAO have jointly established a maximum tolerable human intake level of dioxins via food, and within the EU there are limit values for dioxins in food and feed stuff (EC 2006). Several other EU legislations regulate dioxins, e.g. the plan for integrated pollution prevention and control (IPPC) and directives on waste incineration (EC, 2000, 2008). The EU has also adopted a Community Strategy for dioxins, furans and PCBs (EC 2001).

**Determination of GES boundary**
For dioxins, it was decided to use the GES boundary of 4.0 ng kg-1 ww WHO-TEQ for dioxins and 8.0 ng kg-1 ww WHO-TEQ for dioxins and dl-PCBs.

### 1.4 (3) PFOS indicator

[HElCOM PFOS core indicator document](http://www.helcom.fi/Core%20Indicators/PFOS_HELCOM%20core%20indicator%202016_web%20version.pdf)

### 1.5 Additional references
[Faxneld et al. 2014](http://www.diva-portal.org/smash/record.jsf?pid=diva2%3A728508&dswid=1554) Biological effects and environmental contaminants in herring and Baltic Sea top predators  


# 2. Data

## 2.1 Data sources
ICES  
[ICES database](http://dome.ices.dk/views/ContaminantsBiota.aspx)  
Downloaded 20 October 2015 by Cornelia Ludwig

IVL (Svenska Miljönstitutet / Swedish Environmental Research Institute)  
[IVL database](http://dvsb.ivl.se/)  
Downloaded 2 December 2015 by Cornelia Ludwig  
[IVL detection limit information](http://www.ivl.se/sidor/lab--analys/miljo/biota.html)  

## 2.2 Data prep prior to database
Data were cleaned and harmonized by Cornelia Ludwig prior to being put into the BHI database.  
**PCBs**  
(1) Swedish data were given in lipid weight and were converted to wet weight.  Not all samples had a Extrlip (fat) percentage and therefore could not be converted.  These samples were not included in the database  
(2) Data were in different units and were standardized to mg/kg (however the values listed for the detection limit and the quantification limit were not updated to reflect the value unit change)


# 3. Other Info 
........

# 4. Data Prep - PCB Indicator


```{r TODO, echo=FALSE, message=FALSE}
## read in data...

## plot data...

```
