---
title: "contaminants_prep"
output: github_document
params: 
    datasource: csv
---
```{r setup}

## source common libraries, directories, functions, etc
# Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')

## rprojroot
root <- rprojroot::is_rstudio_project


## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)

dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')

source('~/github/bhi/baltic2015/prep/common.r')
dir_cw    = file.path(dir_prep, 'CW')
dir_con    = file.path(dir_prep, 'CW/contaminants')

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_con, 'contaminants_prep.rmd')

```

# Contaminant Data Prep
## TABLE OF CONTENTS
Section 1.......................Indicator Overview  

Section 2.......................Data Overview  

Section 3.......................Other Info  

Section 4.......................ICES6 Indicator prep  

Section 5.......................Dioxin Indicator prep  

Section 6.......................PFOS Indicator prep  



## 1 Indicators
3 indicators are proposed, describing different aspects of toxicity, and then would be combined to give an overall contamimant sub-component status.  

### 1.2 (1) PCB concentration indicator

#### 1.2.1 ICES-6 PCB
Non-dioxin like PCBs: sum of congeners (28, 52, 101, 138, 153, 180) 75 μg/kg ww fish muscle  

This is similar to the ICES-7 except that PCB 118 is excluded (since it is metabolized by mammals).  

75 ng/g wet weight is the [EU threshold for fish muscle. See Section 5 Annex, 5.3](http://eur-lex.europa.eu/LexUriServ/LexUriServ.do?uri=OJ:L:2011:320:0018:0023:EN:PDF). This threshold was also agreed upon as GES boundary at the most recent meeting of the [Working Group on the State of the Environment and Nature Conservation](http://helcom.fi/helcom-at-work/groups/state-and-conservation) April 11-15, 2016. *Recevied the draft report from Elisabeth Nyberg*  


### 1.3 (2) TEQ value for PCBs and Dioxins
Dioxin and dioxin-like compounds:0.0065 TEQ ug /kg ww fish, crustaceans or molluscs (source of target:EQS biota human health). Secondary GES boundary: CB-118 24 μg/kg lw fish liver or muscle (source: EAC).  

This threshold was agreed upon as GES indicator at the most recent meeting of the [Working Group on the State of the Environment and Nature Conservation](http://helcom.fi/helcom-at-work/groups/state-and-conservation) April 11-15, 2016. *Recevied the draft report from Elisabeth Nyberg*  

Thisis consistent with the [EU human health thresholds for dioxin and dioxin-like compounds - 6.5 pg/g](http://eur-lex.europa.eu/LexUriServ/LexUriServ.do?uri=OJ:L:2011:320:0018:0023:EN:PDF)  

TEQ values from the [World Health Organization 2005](http://www.who.int/ipcs/assessment/tef_values.pdf)  


### 1.4 (3) PFOS indicator

[HELCOM PFOS core indicator document](http://www.helcom.fi/Core%20Indicators/PFOS_HELCOM%20core%20indicator%202016_web%20version.pdf)

### 1.5 Additional references
[Faxneld et al. 2014](http://www.diva-portal.org/smash/record.jsf?pid=diva2%3A728508&dswid=1554) Biological effects and environmental contaminants in herring and Baltic Sea top predators  

[Bignert, A., Nyberg, E., Sundqvist, K.L., Wiberg, K., 2007. Spatial variation in concentrations and patterns of the PCDD/F and dioxin-like PCB content in herring from the northern Baltic Sea. J. Environ. Monit. 9, 550–556.](http://pubs.rsc.org/en/Content/ArticleLanding/2007/EM/b700667e#!divAbstract)

## 2. Data

### 2.1 Data sources

#### 2.1.1 PCB data
**ICES**
[ICES database](http://dome.ices.dk/views/ContaminantsBiota.aspx)  
Downloaded 22 April 2016 by Jennifer Griffiths  
Data selections:  
Year - 1990-2014  
Purpose of monitoring = All  
Country = ALL  
Monitoring Program = ALL  
Parameter Group = Chlorobiophenyls  
Reporting Laboratory = All  
Analytical laboratory = All  
Geographical Areas = (HELCOM) ALL HELCOM Sub-basins  

#### 2.1.2 Dioxins
**ICES**
[ICES database](http://dome.ices.dk/views/ContaminantsBiota.aspx)  
Downloaded on 11 May 2016 by Jennifer Griffiths
Selections:  
Year - 1998-2015  (earliest allowed)  
Purpose of monitoring = All  
Country = ALL  
MOnitoring Program = ALL  
Parameter Group = Dioxins  
Reporting Laboratory = All  
Analytical laboratory = All  
Geographical Areas = (HELCOM) ALL HELCOM Sub-basins  

#### 2.1.3 PFOS


#### Other sources
IVL (Svenska Miljönstitutet / Swedish Environmental Research Institute)  
[IVL database](http://dvsb.ivl.se/)  
Downloaded 2 December 2015 by Cornelia Ludwig  


### 2.2 Data prep prior to database
#### 2.2.1 PCB
Raw data from ICES were prepared in `~/github/bhi/baltic2015/prep/CW/contaminants/raw_contaminants_data_prep`  
    - If duplicate measurements were made from the same sample, values were averaged (all were two measurement of LIPIDWT% per sample, all from 2014, Poland).  
    -  Data were standardized to the same unit ug/kg  
    - If data were presented in lipid weight, they were converted to wet weight by: (EXLIP% / 100)*(CB-conc lipid weight).  

**Note**: Detection limit (detect_lim) and Quantification limit (quant_lim) values are in the original units.  They are not standardized to ug/kg and not converted to wet weight if originally in lipid weight

## 3. Other Info 
### 3.1 Congers in ICES database  
#### 3.1.1 PCB Congeners
CB101       2 2' 4 5 5'-pentachlorobiphenyl  
CB105       2 3 3' 4 4'-pentachlorobiphenyl  
CB118       2 3' 4 4' 5-pentachlorobiphenyl  
CB128     2 2' 3 3' 4 4'-hexachlorobiphenyl  
CB138     2 2' 3 4 4' 5'-hexachlorobiphenyl   
CB149     2 2' 3 4' 5' 6-hexachlorobiphenyl  
CB153     2 2' 4 4' 5 5'-hexachlorobiphenyl  
CB156      2 3 3' 4 4' 5-hexachlorobiphenyl  
CB157     2 3 4 3' 4' 5'-hexachlorobiphenyl  
CB167     2' 3 4 4' 5 5'-hexachlorobiphenyl  
CB169     3 3' 4 4' 5 5'-hexachlorobiphenyl  
CB170  2 2' 3 3' 4 4' 5-heptachlorobiphenyl  
CB180  2 2' 3 4 4' 5 5'-heptachlorobiphenyl  
CB28               2 4 4'-trichlorobiphenyl  
CB52          2 2' 5 5'-tetrachlorobiphenyl  
CB77          3 3' 4 4'-tetrachlorobiphenyl  

#### 3.1.2 Dioxin Congeners
CDD1N         1 2 3 7 8-pentachlorodibenzo-p-dioxin  
CDD4X        1 2 3 4 7 8-hexachlorodibenzo-p-dioxin  
CDD6P     1 2 3 4 6 7 8-heptachlorodibenzo-p-dioxin  
CDD6X        1 2 3 6 7 8-hexachlorodibenzo-p-dioxin  
CDD9X        1 2 3 7 8 9-hexachlorodibenzo-p-dioxin  
CDDO    1 2 3 4 6 7 8 9-octachlorodibenzo-p-dioxin   
CDF2N             2 3 4 7 8-pentachlorodibenzofuran  
CDF2T              2 3 7 8-tetrachloro-dibenzofuran  
CDF4X            2 3 4 6 7 8-hexachlorodibenzofuran  
CDF6P         1 2 3 4 6 7 8-heptachlorodibenzofuran  
CDF6X            1 2 3 6 7 8-hexachlorodibenzofuran  
CDF9P         1 2 3 4 7 8 9-heptachlorodibenzofuran  
CDF9X            1 2 3 7 8 9-hexachlorodibenzofuran  
CDFDN  1 2 3 7 8/1 2 3 4 8-pentachloro-dibenzofuran  
CDFO               octachloro-dibenzofuran (group)   
TCDD   2 3 7 8-tetrachlorodibenzo-p-dioxin   

## 4. Data Prep - PCB Indicator

### 4.1 Basic data cleaning and organization

#### 4.1.1 Read in PCB data
```{r read in data, echo=FALSE, message=FALSE}
## read in PCB data
pcb1 = read.csv(file.path(dir_con, 'contaminants_data_database/ices_herring_pcb.csv'))
dim(pcb1)


## Read in lookup - BHI region to HOLAS basins
lookup_basins = read.csv(file.path(dir_con,'baltic_rgns_to_bhi_rgns_lookup_holas.csv'))

## Read in station impact
lookup_impact = read.csv(file.path(dir_con,'raw_prep/station_impact_cleaned.csv')) ## maybe move to database??


```

#### 4.1.2 Organize
Remove and rename columns, change column data types
```{r organize data}
str(pcb1)

#format date as date
pcb2 = pcb1 %>% 
      dplyr::rename(date=Date, lat=Latitude, lon= Longitude, 
                    month=Month, day=Day, doy=DoY,
                    bhi_id = BHI_ID, helcom_id=HELCOM_ID, 
                    helcom_coastal_code=HELCOM_COASTAL_CODE, ices_area = ICES_AREA,
                    agmax_y = AGMAX_y, agmea_y=AGMEA_y, agmin_y = AGMIN_y,
                    drywt_percent = DRYWT._., exlip_percent = EXLIP._.,fatwt_percent=FATWT._.,
                    lipidwt_percent = LIPIDWT._., lnmax_cm = LNMAX_cm,
                    lnmea_cm=LNMEA_cm, lnmin_cm=LNMIN_cm, wtmax_g=WTMAX_g, wtmea_g=WTMEA_g,
                    wtmin_g=WTMIN_g) %>% # rename columns
      select(-date_ices, -X) %>% #remove columns not needed
      mutate(date =as.Date(date,"%Y-%m-%d"),
             sub_samp_id= as.character(sub_samp_id)) %>%
      select(country:lon,bhi_id:ices_area,date:year,month:doy,species:value)## reorder columns

head(pcb2)

```

#### 4.1.3 Plot data
Explore raw data.
**Why are there unassigned BHI regions?**
```{r plot data}

ggplot(pcb2) + geom_point(aes(date,value, colour=congener)) +
  facet_wrap(~bhi_id)

```

#### 4.1.4 Filter for ICES 6 PCBs
```{r filter ices 6}
ices6pcb = c("CB28_ug/kg", "CB52_ug/kg", "CB101_ug/kg", "CB138_ug/kg",
            "CB153_ug/kg", "CB180_ug/kg")
  
pcb3 = pcb2 %>% 
        filter(congener %in% ices6pcb)

dim(pcb3); dim(pcb2)

```

#### 4.1.5 Plot data -ICES 6 pcbs
```{r plot ices 6}
ggplot(pcb3) + geom_point(aes(date,value, colour=congener)) +
  facet_wrap(~bhi_id)
```

#### 4.1.6 Stations with No BHI ID
Why were some data not assigned a BHI ID (through script in BHI database).  
Will check with Marc, might be because very near coast and BHI regions don't extend?
```{r  bhiid na}

## which stations have no BHI ID assigned
pcb3 %>% filter(is.na(bhi_id)) %>% select(country,station, lat, lon) %>% distinct(.)

##country            station     lat     lon
##1  Sweden     Gaviksfjaerden 62.8645 18.2412
##2  Sweden Kinnbaecksfjaerden 65.0568 21.4750
##3  Sweden  Langvindsfjaerden 61.4554 17.1622

```

#### 4.1.7 Manually assign BHI to stations missing
```{r assign missing bhi regions}

##Gaviksfjaerden is BHI region 37
##Kinnbaecksfjaerden is BHI region 41
##Langvindsfjaerden is BHI region 37

pcb4 = pcb3 %>%
      mutate(bhi_id = ifelse(station=="Gaviksfjaerden",37,
                      ifelse(station=="Kinnbaecksfjaerden", 41,
                      ifelse(station=="Langvindsfjaerden",37,bhi_id))))
pcb4 %>% filter(is.na(bhi_id)) %>% select(country,station, lat, lon) %>% distinct(.)

##plot again
ggplot(pcb4) + geom_point(aes(date,value, colour=congener)) +
  facet_wrap(~bhi_id)


```


#### 4.1.8 Join to HOLAS basins and evaluate data coverage

```{r join to HOLAS basins}
pcb5 = full_join(select(lookup_basins, rgn_id, basin_name), pcb4, by= c("rgn_id"="bhi_id"))
head(pcb5)




```


#### 4.1.9 Plot by basin
```{r plot by basin}

ggplot(pcb5) + geom_point(aes(date,value, colour=congener)) +
  facet_wrap(~basin_name)

```

### 4.2 Evaluate sites sampled
What are the site representing: baseline/reference conditions or local impact.  

#### 4.2.1 Join PCB sites, station dictionary, and impact codes
Some sites have had the site type recorded in the ICES station dictionary (see below). It is important to know which sites are catagorized as:  
    1. *RH* = WFD R(HZ) - Representative of general conditions in terms of hazardous substances  
    2. *B* =  WFD B - Baseline/Reference station  
    3. Any of the codes containing "I" (IH, IH-A, IH-C, IH-D, IH-E, IH-F, IH-H, IH-I, IH-M, IH-O, IH-P, IH-S, IH-W, IP, IP-B, IP-N, IP-T) which refers to a specific type of impact at the site.  

It appears that *only Swedish sites have this information entered* (see below),  Most are RH (19), while B  (4) and RP (1).  RP = WFD R(PHY) Representative of general conditions for nutrients/organic matter.  

**Sites to Include/Exclude**  
Given only Swedish sites have this information recorded, seems difficult to use this information to include or exclude sites.  
```{r join sites with site info }
pcb_sites = pcb5 %>% 
            select(country, station,basin_name,lat,lon,rgn_id) %>%
            filter(!is.na(station)) %>%  ## These are NA because when joined to BHI regions, were no station
            mutate(station2 = tolower(station)) %>% ## station name all lower case to join with station_dictionary
            distinct(.)
pcb_sites
dim(pcb_sites) #[1] 55  7

## clean lookup table
lookup_impact = lookup_impact %>%
                mutate(Station_Name2 = as.character(Station_Name2))


## join with station dictionary

pcb_sites = pcb_sites %>% left_join(., lookup_impact, by=c("station2"="Station_Name2", "country"="Country"))
dim(pcb_sites) # 55 26

pcb_sites %>% select(country, station, basin_name, lat, lon, rgn_id,MSTAT,All_Biota_Data:impact_I) %>% arrange(impact_RH, impact_B, impact_I)

## many sites do not have purpose entered in ICES
## the Finnish sites with no station name could then not be found in the impact lookup
## Many of the sites have "FALSE" under 'Contaminant_parameters_in_biota'
## These sites are also "FALSE" under All_Biota_data 

## from the station dictionary definitions:
## All_Biota_Data: Data type (DTYPE) CF - all parameters - contaminants and biological effects of contaminants including disease in biota
##Contaminant_parameters_in_biota: Data type (DTYPE) CF - Contaminant parameter groups


```


### 4.3 Data flagged for quality

#### 4.3.1 Evaluate data with Qflags
Codes:  
**<** = less than  

**>** = greather than  

**D** = reported value is less than the detection limit (detli)  

**Q** = reported value is less than the limit of quantification (lmqnt)  

**~** separates multiple flags  

No information is given for what **<** implies when is it is used alone (e.g. is it related to the detection limit?)
**Values for detect_lim and quant_lim**: Remember these are in the original units (not standardized to ug_kg or to wet weight)
```{r qflag eval}

## what qflags are present
pcb5 %>% select(qflag) %>% distinct(.) 
##1      
##2     D
##3  <NA>
##4     <
##5     Q
##6   <~D


## "~"  separates multiple flags

## which entries use '<'
pcb5 %>% 
  filter(.,grepl('<', qflag)) %>% select(country,station,date,sub_samp_ref,qflag,detect_lim,congener,value) %>% head()

## appears to be variable usage of '<', values are not always provided for detect_lim when used, values of detect_lim are in original units so are not always comparable to the value column

## which entries use "D"
pcb5 %>% 
  filter(.,grepl('D', qflag)) %>%
select(country,station,date,sub_samp_ref,qflag,detect_lim,congener,value) %>% head()


```

#### 4.3.2 New qflag column
Create second qflag column with "yes" if any symbol used to flag the data, and "no" if not

```{r new qflag column}
pcb6 = pcb5 %>%
      mutate(qflag = as.character(qflag))%>%
      mutate(qflag2 = ifelse(qflag == "", "no",
                      ifelse(qflag== "D", "yes",
                      ifelse(qflag== "<", "yes",
                      ifelse(qflag== "Q", "yes",
                      ifelse(qflag=="<~D", "yes", "")))))) %>%
      mutate(qflag2 = replace(qflag2, is.na(qflag2),"no"))
head(pcb6)

pcb6 %>% filter(qflag2 == "yes") %>% select(qflag,qflag2) %>% distinct(.)
pcb6 %>% filter(qflag2 == "no") %>% select(qflag,qflag2) %>% distinct(.)
pcb6 %>% filter(qflag2 == "") %>% select(qflag,qflag2) %>% distinct(.)
pcb6 %>% filter(is.na(qflag)) %>% select(qflag,qflag2) %>% distinct(.)

```

#### 4.3.3 Plot data with qflag 
Plot each congener separately. CB 52, and CB 28 in particular have many flagged congeners.
```{r}
## congeners to plot
# 1 CB101_ug/kg
# 2 CB138_ug/kg
# 3 CB153_ug/kg
# 4 CB180_ug/kg
# 5  CB28_ug/kg
# 6  CB52_ug/kg

ggplot(filter(pcb6, congener =="CB101_ug/kg")) +
  geom_point(aes(date,value,shape=qflag2, colour= qflag2)) + 
  facet_wrap(~basin_name) +
  scale_shape_manual(values=c(1,2))+
  ggtitle("CB101_ug/kg")

ggplot(filter(pcb6, congener =="CB138_ug/kg")) +
  geom_point(aes(date,value,shape=qflag2, colour= qflag2)) + 
  facet_wrap(~basin_name) +
  scale_shape_manual(values=c(1,2))+
   ggtitle("CB138_ug/kg")

ggplot(filter(pcb6, congener =="CB153_ug/kg")) +
  geom_point(aes(date,value,shape=qflag2, colour= qflag2)) + 
  facet_wrap(~basin_name) +
  scale_shape_manual(values=c(1,2))+
   ggtitle("CB153_ug/kg")

ggplot(filter(pcb6, congener =="CB180_ug/kg")) +
  geom_point(aes(date,value,shape=qflag2, colour= qflag2)) + 
  facet_wrap(~basin_name) +
  scale_shape_manual(values=c(1,2))+
   ggtitle("CB180_ug/kg")

ggplot(filter(pcb6, congener =="CB28_ug/kg")) +
  geom_point(aes(date,value,shape=qflag2, colour= qflag2)) + 
  facet_wrap(~basin_name) +
  scale_shape_manual(values=c(1,2))+
   ggtitle("CB28_ug/kg")

ggplot(filter(pcb6, congener =="CB52_ug/kg")) +
  geom_point(aes(date,value,shape=qflag2, colour= qflag2)) + 
  facet_wrap(~basin_name) +
  scale_shape_manual(values=c(1,2))+
   ggtitle("CB52_ug/kg")



```

#### 4.3.4 Number of observations w/ and w/o qflag
*2123* samples (not unique sample dates) where all 6 congeners measured; only *1031* have no qflagged congeners.  
```{r obs }
## number of C153 observations
## including qflag observations
pcb6 %>% filter(congener=="CB153_ug/kg")%>%nrow()
##2199

## w/o qflag observatons
pcb6 %>% filter(congener=="CB153_ug/kg" & qflag2 =="no")%>%nrow()
##2129



## number of observations with all ICES 6
ices6_obs = pcb6 %>% 
  select(rgn_id:bulk_id,congener,value) %>%
  filter(!is.na(congener)) %>% ## this is for bhi_regions with no data
  group_by(rgn_id, basin_name ,country,monit_program,monit_purpose, report_institute,station,
           lat,lon, helcom_id ,helcom_coastal_code,ices_area,date, monit_year, year, month,
            day,doy,species, sub_samp_ref,sub_samp_id,samp_id, num_indiv_subsample, bulk_id)%>%
  spread(congener, value) %>%
  ungroup() %>%
  mutate(sum_ices6 =  (`CB101_ug/kg` + `CB138_ug/kg` + `CB153_ug/kg` + `CB180_ug/kg` + `CB28_ug/kg` + `CB52_ug/kg`)) %>%  ## sum all the congeners, if any are NA, result is NA
  filter(!is.na(sum_ices6))
ices6_obs %>% nrow() #2123

## Number of observations without qflagged data
ices6_obs_noq = pcb6 %>% 
  select(rgn_id:bulk_id,qflag2,congener,value) %>%
  filter(!is.na(congener)) %>% ## this is for bhi_regions with no data
  filter(qflag2 == "no") %>%  ## filter for only data not flagged
  select(-qflag2) %>%
  group_by(rgn_id, basin_name ,country,monit_program,monit_purpose, report_institute,station,
           lat,lon, helcom_id ,helcom_coastal_code,ices_area,date, monit_year, year, month,
            day,doy,species, sub_samp_ref,sub_samp_id,samp_id, num_indiv_subsample, bulk_id)%>%
  spread(congener, value) %>%
  ungroup() %>%
  mutate(sum_ices6 =  (`CB101_ug/kg` + `CB138_ug/kg` + `CB153_ug/kg` + `CB180_ug/kg` + `CB28_ug/kg` + `CB52_ug/kg`)) %>%  ## sum all the congeners, if any are NA, result is NA
  filter(!is.na(sum_ices6))

  ices6_obs_noq %>% nrow()  #1031
  
  
```

#### 4.3.5 Second value column with qflag adjustment
If the data have any qflag, apply LOD/2.  However, LOD (eg. detect_lim column) value is not always provided and values are in original units.  Instead, apply transformation to the reported data value.

```{r ajdust for qflag}
 pcb7 = pcb6 %>%
        mutate(value2= ifelse(qflag2=="yes", value/2, value))

```

#### 4.3.5 Plot qflag adjusted values and not adjusted values
Each congener plotted in separate plot.  

**CB101**  
Plot 1. First both raw values (red) and adjusted (black).  
Plot 2 .Second, only adjusted values where blue trianges are those where LOD/2 applied.  
```{r plot adjusted values CB101}
ggplot(filter(pcb7, congener =="CB101_ug/kg")) +
  geom_point(aes(date,value), color="red", shape=19) + 
   geom_point(aes(date,value2), color="black",shape=1)+
  facet_wrap(~basin_name, scales="free_y") +
  ggtitle("Plot 1. CB101_ug/kg")

ggplot(filter(pcb7, congener =="CB101_ug/kg")) +
   geom_point(aes(date,value2, color=qflag2,shape=qflag2))+
  facet_wrap(~basin_name, scales="free_y") +
  scale_shape_manual(values=c(1,2))+
  ggtitle("Plot 2. CB101_ug/kg")

```

**CB138**  
Plot 1.First both raw values (red) and adjusted (black).  
Plot 2.Second, only adjusted values where blue trianges are those where LOD/2 applied.  

```{r plot adjusted values CB138}
ggplot(filter(pcb7, congener =="CB138_ug/kg")) +
  geom_point(aes(date,value), color="red", shape=19) + 
   geom_point(aes(date,value2), color="black",shape=1)+
  facet_wrap(~basin_name, scales="free_y") +
  ggtitle("Plot 1. CB138_ug/kg")

ggplot(filter(pcb7, congener =="CB138_ug/kg")) +
   geom_point(aes(date,value2, color=qflag2,shape=qflag2))+
  facet_wrap(~basin_name, scales="free_y") +
  scale_shape_manual(values=c(1,2))+
  ggtitle("Plot 2. CB138_ug/kg")


```

**CB153**  
Plot 1.First both raw values (red) and adjusted (black).  
Plot 2.Second, only adjusted values where blue trianges are those where LOD/2 applied.  

```{r plot adjusted values CB153}
ggplot(filter(pcb7, congener =="CB153_ug/kg")) +
  geom_point(aes(date,value), color="red", shape=19) + 
   geom_point(aes(date,value2), color="black",shape=1)+
  facet_wrap(~basin_name, scales="free_y") +
  ggtitle("Plot 1. CB153_ug/kg")

ggplot(filter(pcb7, congener =="CB153_ug/kg")) +
   geom_point(aes(date,value2, color=qflag2,shape=qflag2))+
  facet_wrap(~basin_name, scales="free_y") +
  scale_shape_manual(values=c(1,2))+
  ggtitle("Plot 2.CB153_ug/kg")


```

**CB180**  
Plot 1.First both raw values (red) and adjusted (black).  
Plot 2.Second, only adjusted values where blue trianges are those where LOD/2 applied.  

```{r plot adjusted values CB180}
ggplot(filter(pcb7, congener =="CB180_ug/kg")) +
  geom_point(aes(date,value), color="red", shape=19) + 
   geom_point(aes(date,value2), color="black",shape=1)+
  facet_wrap(~basin_name, scales="free_y") +
  ggtitle("Plot 1. CB180_ug/kg")

ggplot(filter(pcb7, congener =="CB180_ug/kg")) +
   geom_point(aes(date,value2, color=qflag2,shape=qflag2))+
  facet_wrap(~basin_name, scales="free_y") +
  scale_shape_manual(values=c(1,2))+
  ggtitle("Plot 2. CB180_ug/kg")

```

**CB28**  
Plot 1. First both raw values (red) and adjusted (black).  
Plot 2. Second, only adjusted values where blue trianges are those where LOD/2 applied.  

```{r plot adjusted values CB28}

ggplot(filter(pcb7, congener =="CB28_ug/kg")) +
  geom_point(aes(date,value), color="red", shape=19) + 
   geom_point(aes(date,value2), color="black",shape=1)+
  facet_wrap(~basin_name, scales="free_y") +
  ggtitle("Plot 1. CB28_ug/kg")

ggplot(filter(pcb7, congener =="CB28_ug/kg")) +
   geom_point(aes(date,value2, color=qflag2,shape=qflag2))+
  facet_wrap(~basin_name, scales="free_y") +
  scale_shape_manual(values=c(1,2))+
  ggtitle("Plot 2. CB28_ug/kg")

```

**CB52**  
Plot 1. First both raw values (red) and adjusted (black).  
Plot 2. Second, only adjusted values where blue trianges are those where LOD/2 applied.  

```{r plot adjusted values CB52}
ggplot(filter(pcb7, congener =="CB52_ug/kg")) +
  geom_point(aes(date,value), color="red", shape=19) + 
   geom_point(aes(date,value2), color="black",shape=1)+
  facet_wrap(~basin_name, scales="free_y") +
  ggtitle("Plot 1. CB52_ug/kg")

ggplot(filter(pcb7, congener =="CB52_ug/kg")) +
   geom_point(aes(date,value2, color=qflag2,shape=qflag2))+
  facet_wrap(~basin_name, scales="free_y") +
  scale_shape_manual(values=c(1,2))+
  ggtitle("Plot 2. CB52_ug/kg")


```

#### 4.4 ICES6 PCB

#### 4.4.1 ICES6 PCB sum
1. Sum only samples with all 6 PCBs observed.  
2. Sum1 = sum value2 (qflag adjusted)
3. Sum2 = sum only observation with no qflag values
```{r ices6 sum}

ices6_1 =pcb7 %>%
        select(rgn_id:bulk_id,congener,value2) %>%
        filter(!is.na(congener)) %>% ## this is for bhi_regions with no data
        group_by(rgn_id, basin_name ,country,monit_program,monit_purpose, 
                 report_institute,station,lat,lon,helcom_id,helcom_coastal_code,
                 ices_area,date, monit_year, year, month,day,doy,species,sub_samp_ref,
                 sub_samp_id,samp_id, num_indiv_subsample, bulk_id)%>%
        spread(congener, value2) %>%
        ungroup() %>%
        mutate(sum_ices6_ug_kg =  (`CB101_ug/kg` + `CB138_ug/kg` + `CB153_ug/kg` + `CB180_ug/kg` + `CB28_ug/kg` + `CB52_ug/kg`)) %>%
        filter(!is.na(sum_ices6_ug_kg))%>% ## remove any observations with NA (eg not all congeners observed)
        arrange(country,station,date,sub_samp_ref)

dim(ices6_1) #[1] 2123   31


ices6_2 =pcb7 %>%
        select(rgn_id:bulk_id,qflag2,congener,value2) %>%
        filter(!is.na(congener)) %>% ## this is for bhi_regions with no data
        filter(qflag2 == "no") %>%  ## filter for only data not flagged
        select(-qflag2) %>%
        group_by(rgn_id, basin_name ,country,monit_program,monit_purpose, 
                 report_institute,station,lat,lon,helcom_id,helcom_coastal_code,
                 ices_area,date, monit_year, year, month,day,doy,species,sub_samp_ref,
                 sub_samp_id,samp_id, num_indiv_subsample, bulk_id)%>%
        spread(congener, value2) %>%
        ungroup() %>%
        mutate(sum_ices6_ug_kg =  (`CB101_ug/kg` + `CB138_ug/kg` + `CB153_ug/kg` + `CB180_ug/kg` + `CB28_ug/kg` + `CB52_ug/kg`))    %>%
        filter(!is.na(sum_ices6_ug_kg)) %>% ## remove any observations with NA (eg not all congeners observed)            
        arrange(country,station,date,sub_samp_ref)

dim(ices6_2)#1031   31


```

#### 4.4.2 Plot ICES6

```{r plot ices6 with qflag}

ices6_1= ices6_1 %>%
              select(-`CB101_ug/kg` ,-`CB138_ug/kg` , -`CB153_ug/kg` ,
                     -`CB180_ug/kg`, -`CB28_ug/kg`,- `CB52_ug/kg`) %>% ## remove congener columns
          mutate(health_threshold=75)  ## Add EU human health threshold 75 ug/kg

ggplot(ices6_1) + 
  geom_point(aes(date,sum_ices6_ug_kg))+
  geom_line(aes(date,health_threshold))+
  facet_wrap(~basin_name, scales="free_y")+
  ggtitle("ICES6 including qflag-adjusted")+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))


ices6_2= ices6_2 %>%
              select(-`CB101_ug/kg` ,-`CB138_ug/kg` , -`CB153_ug/kg` ,
                     -`CB180_ug/kg`, -`CB28_ug/kg`,- `CB52_ug/kg`) %>% ## remove congener columns
          mutate(health_threshold=75)  ## Add EU human health threshold 75 ug/kg

ggplot(ices6_2) + 
  geom_point(aes(date,sum_ices6_ug_kg))+
  geom_line(aes(date,health_threshold))+
  facet_wrap(~basin_name, scales="free_y")+
  ggtitle("ICES6 excluded qflag")+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))

```



#### 4.4.3 ICES6 mean value by date and location
Take the mean value by date, do for both qflag-adjusted and no-qflag dataset

```{r mean ices 6 by date}

##including qflagged adjusted
ices6_1datemean = ices6_1 %>% 
                  select(-sub_samp_ref,-sub_samp_id,-samp_id, -num_indiv_subsample, -bulk_id) %>% ##remove unique id
                  group_by(rgn_id, basin_name ,country,monit_program,monit_purpose, 
                 report_institute,station,lat,lon,helcom_id,helcom_coastal_code,
                 ices_area,date, monit_year, year, month,day,doy,species) %>%
                summarize(ices6_datemean_ug_kg = mean(sum_ices6_ug_kg))%>%
                ungroup()

dim(ices6_1datemean) # 227  20


## excluding qflagged
ices6_2datemean = ices6_2 %>% 
                  select(-sub_samp_ref,-sub_samp_id,-samp_id, -num_indiv_subsample, -bulk_id) %>% ##remove unique id
                  group_by(rgn_id, basin_name ,country,monit_program,monit_purpose, 
                 report_institute,station,lat,lon,helcom_id,helcom_coastal_code,
                 ices_area,date, monit_year, year, month,day,doy,species) %>%
                summarize(ices6_datemean_ug_kg = mean(sum_ices6_ug_kg))%>%
                ungroup()

dim(ices6_2datemean) #145  20


```

#### 4.4.4 Plot ICES6 mean value by date and location
**Plot By Basin**
Gray dots are the ices6 concentration mean by date and location including qflagg-adjusted values.  
Red dots are the ices6 concentration mean by date and location excluding qflagged values.  
    - Including the qflag-adjusted values lowers the mean concentration by date and location, provides more observations in the Kattegat, The Quark, and W. Gotland Basin 
    
    - Outlier in Eastern Gotland Basin is from Polish observations 2014. Have checked unit conversions etc, have found not error.

```{r plot ices6 date mean by basin plot}

##temporily merge ices6_1datemean and ices6_2datemean to more easily compare in a plot
plotdata_temp = full_join(ices6_1datemean,ices6_2datemean,
                          by=c("rgn_id", "basin_name", "country", "monit_program", 
                               "monit_purpose", "report_institute", "station", "lat", 
                               "lon", "helcom_id", "helcom_coastal_code", "ices_area", 
                               "date", "monit_year", "year", "month", "day", "doy", "species")) %>%
                          dplyr::rename(ices6_datemean_qflag_adj=ices6_datemean_ug_kg.x,
                                        ices6_datemean_no_qflag=ices6_datemean_ug_kg.y) %>%
                          mutate(health_threshold=75)

##not showing human health threshold
# ggplot(plotdata_temp) +
#   geom_point(aes(date,ices6_datemean_qflag_adj),colour="gray")+
#   geom_point(aes(date,ices6_datemean_no_qflag),colour="red")+
#   facet_wrap(~basin_name, scales="free_y")+
#   ggtitle("Mean ICES6 concentration ug/kg by date & location")

##plot by year
ggplot(plotdata_temp) +
  geom_point(aes(year,ices6_datemean_qflag_adj),colour="gray")+
  geom_point(aes(year,ices6_datemean_no_qflag),colour="red")+
  facet_wrap(~basin_name, scales="free_y")+
  ggtitle("Mean ICES6 concentration ug/kg by date & location plot by year")

##showing the human health threshold                    
ggplot(plotdata_temp) +
  geom_point(aes(date,ices6_datemean_qflag_adj),colour="gray")+
  geom_point(aes(date,ices6_datemean_no_qflag),colour="red")+
  geom_line(aes(date,health_threshold))+
  facet_wrap(~basin_name, scales="free_y")+
  ggtitle("Mean ICES6 concentration ug/kg by date & location w/ Human Health threshold")

##Plot variation in the 5 year status period
ggplot(filter(plotdata_temp, year >2008 & year < 2013)) + 
  geom_boxplot(aes(basin_name, ices6_datemean_qflag_adj))+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Mean ICES6 concentration (ug/kg) variation by basin, 2009-2013")

```

**Plot By BHI Region**
Gray dots are the ices6 concentration mean by date and location including qflagg-adjusted values.  
Red dots are the ices6 concentration mean by date and location excluding qflagged values.  
    - More observations for Regions 1, 11,26,35,36,39,41,42 when including qflagged values.  
```{r ices6 mean date and location plot by bhi region}


#not showing human health threshold
ggplot(plotdata_temp) +
  geom_point(aes(date,ices6_datemean_qflag_adj),colour="gray")+
  geom_point(aes(date,ices6_datemean_no_qflag),colour="red")+
  facet_wrap(~rgn_id, scales="free_y")+
  ggtitle("Mean ICES6 concentration ug/kg by date & location")

 #plot by year
ggplot(plotdata_temp) +
  geom_point(aes(year,ices6_datemean_qflag_adj),colour="gray")+
  geom_point(aes(year,ices6_datemean_no_qflag),colour="red")+
  facet_wrap(~rgn_id, scales="free_y")+
  ggtitle("Mean ICES6 concentration ug/kg by date & location plot by year")


##showing the human health threshold                    
ggplot(plotdata_temp) +
  geom_point(aes(date,ices6_datemean_qflag_adj),colour="gray")+
  geom_point(aes(date,ices6_datemean_no_qflag),colour="red")+
  geom_line(aes(date,health_threshold))+
  facet_wrap(~rgn_id, scales="free_y")+
  ggtitle("Mean ICES6 concentration ug/kg by date & location w/ Human Health threshold")


```



#### 4.4.5 Most recent year obs by each location
Using 2009 -2013 for status assessment is most consistent (and 2014 are from Poland and look like extreme outliers).  There will be *no* status for Gulf of Finland, most recent data 2007.

```{r most recent year}
ices6_1datemean %>% arrange(year) %>%group_by(basin_name) %>% summarise(last(year))%>%ungroup()
#                basin_name last(year)
#                    (fctr)      (int)
# 1               Aland Sea       2013
# 2            Arkona Basin       2013
# 3          Bornholm Basin       2014
# 4            Bothnian Bay       2013
# 5            Bothnian Sea       2013
# 6   Eastern Gotland Basin       2014
# 7         Gulf of Finland       2007
# 8                Kattegat       2013
# 9  Northern Baltic Proper       2013
# 10              The Quark       2013
# 11  Western Gotland Basin       2013

ices6_2datemean %>% arrange(year) %>% group_by(basin_name) %>% summarise(last(year))%>%ungroup()

#                basin_name last(year)
#                    (fctr)      (int)
# 1               Aland Sea       2011
# 2            Arkona Basin       2013
# 3          Bornholm Basin       2013
# 4            Bothnian Bay       2013
# 5            Bothnian Sea       2013
# 6   Eastern Gotland Basin       2012
# 7         Gulf of Finland       2007
# 8                Kattegat       2013
# 9  Northern Baltic Proper       2013
# 10  Western Gotland Basin       2011

```

### 4.5 Status and Trend Evaluation
**Status Alternatives**
Compare calculating the status by BHI region or Basin.  
Use data including qflag-adjusted (this could lower values).  
Use 5 year mean ICES6 concentration.  

Steps:  
1. Have mean ICES conc. by date and location (our unique observations)  
2. Take mean of all unique observations for 2009-2013 in either BHI region or basin  
3. If done by basin, give basin score to all regions within basin  
4. Compare results  
5. Assess number of data points contributing  

**Trend**
Compare calculating trend by BHI region or basin.  
Use data including qflag-adjusted (this could lower values).  

Steps:  
1. Linear regression of unique observations in a basin or region  
2. Use a 10 year period: 2004-2013  
3. Minimum of different observation years  
4. Scale all observations relative to the human health threshold?  
**Why Scale?**  If values are all below the threshold would not want an increase trend in observed values to suggest that the future status will be worse??
4. **Need to have mixed-effects to account for different stations?**

#### 4.5.1  Mean Basin
Calculate the 2009-2013 mean value by basin and the number of observations. Assign basin values to BHI regions.
```{r  mean basin }
## calculate the mean value by basin
ices6_mean_basin = ices6_1datemean %>% 
                filter(year >=2009 & year <= 2013) %>% ## select years 2009
                select(basin_name,ices6_datemean_ug_kg) %>%
                group_by(basin_name) %>%
                summarise(ices6_mean_ug_kg = round(mean(ices6_datemean_ug_kg),2))%>%
                ungroup()
                
## number of observations by basin
ices6_mean_basin_num_obs = ices6_1datemean %>%
                             filter(year >=2009 & year <= 2013) %>% ## select years 2009 
                             select(basin_name,ices6_datemean_ug_kg) %>%
                             count(basin_name)
## number of locations sampled by basin
ices6_mean_basin_num_loc = ices6_1datemean %>%
                             filter(year >=2009 & year <= 2013) %>% ## select years 2009 
                             select(basin_name,lat,lon) %>%
                             distinct(.)%>%
                             count(basin_name)

## join mean and number of observations and number of locations
ices6_mean_basin  = ices6_mean_basin %>%
                      full_join(., ices6_mean_basin_num_obs, by="basin_name") %>% 
                      dplyr::rename(num_obs=n) %>%
                      full_join(.,ices6_mean_basin_num_loc, by="basin_name" ) %>%
                      dplyr::rename(num_loc=n)
                                    
ices6_mean_basin

## assign basin values to BHI regions
ices6_mean_basin_rgn = ices6_mean_basin %>%
                         full_join(., select(lookup_basins,rgn_id,basin_name),by="basin_name")%>%
                          mutate(mean_type="basin")
ices6_mean_basin_rgn




```

##### 4.5.1.1 Plot Mean Basin
Symbols represent the number of unique locations sampled within the basin. Size indicates the number of observations in the basin (unique date x location)
```{r plot mean basin}

ggplot(ices6_mean_basin_rgn) + 
  geom_point(aes(basin_name,ices6_mean_ug_kg, size=num_obs,shape=factor(num_loc)))+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
    ggtitle("Mean ICES6 conc calculated by basin 2009-2013")


```


#### 4.5.2 Mean BHI Region
Calculate the 2009-2013 mean value by BHI region and the number of observations. 
```{r mean bhi region }
ices6_mean_region = ices6_1datemean %>% 
                filter(year >=2009 & year <= 2013) %>% ## select years 2009
                select(rgn_id,ices6_datemean_ug_kg) %>%
                group_by(rgn_id) %>%
                summarise(ices6_mean_ug_kg = round(mean(ices6_datemean_ug_kg),2))%>%
                ungroup()
                
## number of observations by region
ices6_mean_region_num_obs = ices6_1datemean %>%
                             filter(year >=2009 & year <= 2013) %>% ## select years 2009 
                             select(rgn_id,ices6_datemean_ug_kg) %>%
                             count(rgn_id)
## number of observations by region
ices6_mean_region_num_loc= ices6_1datemean %>%
                             filter(year >=2009 & year <= 2013) %>% ## select years 2009 
                             select(rgn_id,lat,lon) %>%
                             distinct(.)%>%
                             count(rgn_id)


## join mean and number of observations
ices6_mean_region  = ices6_mean_region %>%
                      full_join(., ices6_mean_region_num_obs, by="rgn_id") %>%
                      dplyr::rename(num_obs=n)%>%
                      full_join(., ices6_mean_region_num_loc, by="rgn_id")%>%
                     dplyr::rename(num_loc=n)
ices6_mean_region

## join with lookup_basins so have rgns with NA
ices6_mean_region = ices6_mean_region %>%
                      full_join(., select(lookup_basins, rgn_id), by="rgn_id") %>%
                      mutate(mean_type= "region")


```

##### 4.5.2.1 Plot Mean Region
Symbols represent the number of unique locations sampled within the region. Size indicates the number of observations in the region (unique date x location)
```{r plot mean region}

ggplot(ices6_mean_region) + 
  geom_point(aes(rgn_id,ices6_mean_ug_kg, size=num_obs,shape=factor(num_loc)))+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
    ggtitle("Mean ICES6 conc calculated by BHI Region 2009-2013")


```


#### 4.5.3  Mean BHI Region
Compare mean calculations using region or basin.  
```{r compare mean calculations}
##join data so can plot together
ices6_mean = bind_rows(ices6_mean_region,select(ices6_mean_basin_rgn, -basin_name))

#ggplot(ices6_mean) + 
  #geom_point(aes(rgn_id,ices6_mean_ug_kg, size=num_obs))+
  #facet_wrap(~mean_type)+
  #ggtitle("Mean ICES6 conc calculated by basin or BHI region")


ggplot(ices6_mean) + 
  geom_point(aes(rgn_id,ices6_mean_ug_kg, size=num_obs, colour=mean_type,shape=mean_type))+
   scale_shape_manual(values=c(1,2))+
    ggtitle("Mean ICES6 conc calculated by basin or BHI region")

ggplot(ices6_mean) + 
  geom_point(aes(rgn_id,ices6_mean_ug_kg, size=num_loc, colour=mean_type,shape=mean_type))+
   scale_shape_manual(values=c(1,2))+
    ggtitle("Mean ICES6 conc calculated by basin or BHI region")


```

#### 4.5.4 Calculate Status
Xices6 = 1/ (Mean_ICES6_region / Reference_pt  )
Reference_pt = health_threshold=  75ug/kg

Scale between 0 and 1.  If value is below 75, score = 1
```{r  status calculation}
health_threshold = 75

ices6_mean = ices6_mean %>%
              mutate(ices6_status = pmin(1,1/(ices6_mean_ug_kg/health_threshold))) ## status calculation, scaled between 0-1

## because no values exceed the human health threshold, status score is 1.      

```

#### 4.5.5 Plot status
```{r plot status}

ggplot(ices6_mean)+ 
  geom_point(aes(rgn_id, ices6_status*100))+
  facet_wrap(~mean_type)+
  ylim(0,100)+
  ggtitle("ICE6 Status based on 2009-2013, calculated for basin or region")

```


#### 4.5.6 Trend calculation by Basin
Need to think abou the interpretation of the data treatment.  
1. If use raw observations, then normalize (zscore data), then fit trend, if get increase or decrease but all values are below the threshold, does it make sense to apply a change in the trend to the status?  Would we really think the future status will be lower?  

2. If take all raw observations, calculate "status" as done for the mean value, then fit trend, is this more true to the idea that variation below the human health threshold should not affect the trajectory of the future status?  

3. Need to think if simple linear regression is okay, or if need to account for site?
```{r basin trend calculation}
## Key constant
lag_win = 5 ## the value of the future year we are interested in (five years in the future)



## number of years of observations ## Check to make sure all 5 or >
ices_basin_num_years=ices6_1datemean %>%
  filter(year >=2004 & year <= 2013) %>%
  select(basin_name,year)%>%
  distinct(.)%>%
  group_by(basin_name,year) %>%
  summarise(num=n())%>%
  ungroup() %>% 
  select(-num)%>%
  group_by(basin_name)%>%
  count(basin_name)%>%ungroup()

## Exclude Gulf of Finland -- 3 year (also no status)

## also exclude basins for which no status was calculated
    basin_list = ices6_mean_basin_rgn %>% 
                  filter(!is.na(ices6_mean_ug_kg)) %>%
                  select(basin_name)%>%
                  distinct()
    basin_list = as.character(basin_list$basin_name)

    

## Calculate trend with zscore of data    
ices6_basin_trend_alt1 = ices6_1datemean %>% 
                    filter(year >=2004 & year <= 2013) %>% ## select years 2004-2013
                    filter(basin_name %in%  basin_list ) %>%
                    select(basin_name,year,date,ices6_datemean_ug_kg)%>%
                    group_by(basin_name) %>%
                    mutate(mean = mean(ices6_datemean_ug_kg),
                           sd = sd(ices6_datemean_ug_kg),
                           ices6_zscore = (ices6_datemean_ug_kg - mean)/sd) %>% ## zscore values
                      select(basin_name, year,ices6_zscore) %>%
                    do(trend_mdl = lm(ices6_zscore~year, data=.)) %>%  ## regression model to get trend from raw observations
                    summarize(basin_name=basin_name,
                              trend_score = coef(trend_mdl)['year']*lag_win) %>%  ## lag_win defined above, numbe of years in the future for future status
                    ungroup()





## Calculate trend by making each observation a "status" value
ices6_basin_trend_alt2 = ices6_1datemean %>% 
                    filter(year >=2004 & year <= 2013) %>% ## select years 2004-2013
                    filter(basin_name %in%  basin_list ) %>%
                    select(basin_name,year,date,ices6_datemean_ug_kg)%>%
                    mutate(obs_status= pmin(1,1/(ices6_datemean_ug_kg/health_threshold))) %>%
                    group_by(basin_name)%>%
                    do(trend_mdl = lm(obs_status~year, data=.)) %>%  ## regression model to get trend from raw observations
                    summarize(basin_name=basin_name,
                              trend_score = coef(trend_mdl)['year']*lag_win) %>%  ## lag_win defined above, numbe of years in the future for future status
                    ungroup()




```

### 4.5.7 Plot linear model trends - 2 approaches

```{r  plot trend alternatives}
ices6_basin_trend_alt1 = ices6_basin_trend_alt1 %>%
                          mutate(trend_approach = "zscore")
ices6_basin_trend_alt2 = ices6_basin_trend_alt2 %>%
                          mutate(trend_approach = "status")
trend_alts = bind_rows(ices6_basin_trend_alt1,ices6_basin_trend_alt2)

ggplot(trend_alts)+
  geom_point(aes(basin_name,trend_score))+
  facet_wrap(~trend_approach)+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Alternative Trend Model Fitting (2004-2013)")
      


```


#### 4.5.8 Trend calculation by Basin with mixed-effects model
IN PROGRESS.......errors  
Only fitting for the z-score approach because with status approach all values are 1 and this causes model fitting problems.  
Current error for fitting the zscore data seems to be an issue if there is only 1 factor (does not want fit the random effect) - so tried the if() statement.  But now get error "Error: cannot modify grouping variable"
```{r mixed effects trend model}

## calculate the trend on the zscored raw values, with a mixed effects model so that there is a random intercept by location.  Use lat because not all have a station name  (REML used for coefficient estimates)
# library(lme4)
# library(broom)
# ices6_basin_trend_alt1b = ices6_1datemean %>% 
#                     filter(year >=2004 & year <= 2013) %>% ## select years 2004-2013
#                     filter(basin_name %in%  basin_list ) %>%
#                     select(basin_name,year,date,lat, ices6_datemean_ug_kg)%>%
#                     group_by(basin_name) %>%
#                     mutate(mean = mean(ices6_datemean_ug_kg),
#                            sd = sd(ices6_datemean_ug_kg),
#                            ices6_zscore = (ices6_datemean_ug_kg - mean)/sd) %>% ## zscore values
#                     ungroup()%>%
#                     select(basin_name,lat, year,ices6_zscore) %>%
#                     group_by(basin_name) %>%
#                     #mutate(f_lat = factor(lat))%>% 
#                     do({if(distinct(.,lat) %>% nrow() >1)
#                       (trend_mdl = lmer(ices6_zscore ~ year + (1|lat), data =.))
#                       else
#                        (trend_mdl = lm(ices6_zscore ~ year, data =.))}) %>%  ## regression model to get trend from raw observations
#                     summarize(basin_name=basin_name,
#                               trend_score = coef(trend_mdl)['year']*lag_win) %>%  ## lag_win defined above, numbe of years in the future for future status
#                     ungroup()

```


### 4.6 Method discussion with Anna Sobek  
1. Indicator choice: *We agreed that ICES6 is the best option *  
2. Decision about use of qflagg-adjusted data: *use qflagged data with the adjustement of (congener conc/2)*  
3. Decision about spatial scale of the data: *decide best approach is to calculate for each basin*  
4. Trend decision: *best approach is first convert individual observations to a "status" relative to the human health threshold, then fit linear model by basin for 10 year period* **TREND CHECK  - Does the trend value need to be rescaled to between -1 and 1? Does not exceed now but need to consider if method broadly works?)**  


### 4.7 Final Status and Trend obejcts
Create final status and trend objects to export as csv to layers folder. These layers are then registered in layers.csv

```{r final ices6 objects}

## STATUS
## status calculated in 4.5.4
status_ices6 = ices6_mean %>%
              filter(mean_type=="basin")%>%  ## select status calculated by basin
              select(rgn_id, ices6_status)%>%
              dplyr::rename(score = ices6_status) %>%
              mutate(dimension = "status",
                     score = score*100) %>%
              select(rgn_id, dimension,score) %>%
              arrange(rgn_id)
  

## TREND
## Trend calculated iin 4.5.6
trend_ices6 = ices6_basin_trend_alt2 %>%
              full_join(.,select(lookup_basins, rgn_id,basin_name), by= "basin_name")%>%
              select(rgn_id, trend_score)%>%
              rename(score = trend_score) %>%
              mutate(dimension = "trend",
                     score = round(score,2))%>%
              select(rgn_id,dimension,score)%>%
              arrange(rgn_id)
              


## WRITE AS CSV

write.csv(status_ices6, file.path(dir_layers, 'cw_con_ices6_status_bhi2015.csv'), row.names=FALSE)

write.csv(trend_ices6, file.path(dir_layers, 'cw_con_ices6_trend_bhi2015.csv'), row.names=FALSE)

```



## 5. Data Prep - Dioxin Indicator
### TO DO  
read in dioxin and pcb data  
read in congener TEF values  
select only dioxin-like pcbs  
can I align sub samp ref for pcb data and dioxins  
how many congeners measured by sample?  
methods for how to sum up (fixed set of congeners or anything measured)?  



## 6. Data Prep - PFOS Indicator


