---
output: html_document
---

## Food Provision: Fisheries

```{r, child='goals/FIS.Rmd', results='asis', echo=FALSE}
```

[**Full data preparation information and code**](https://github.com/OHI-Science/bhi/blob/draft/baltic2015/prep/FIS/fis_np_prep.md).   
<!--- For goal model code, click [here](link to functions.r in specific commit/release). --->  


## Food Provision: Mariculture

```{r, child='goals/MAR.Rmd', results='asis', echo=FALSE}
```

[**Full data preparation information and code**](https://github.com/OHI-Science/bhi/blob/draft/baltic2015/prep/MAR/mar_prep.md)


## Artisanal Fishing Opportunity 

[**Full data preparation information and code**]()

## Natural Products

```{r, child='goals/NP.Rmd', results='asis', echo=FALSE}
```

[**Full data preparation information and code**](https://github.com/OHI-Science/bhi/blob/draft/baltic2015/prep/NP/np_prep.md).   
<!--- For goal model code, click [here](link to functions.r in specific commit/release). --->  

## Carbon Storage

[**Full data preparation information and code**]()

## Tourism & Recreation

[**Full data preparation information and code**]()

## Livelihoods & Economies: Livelihoods

[**Full data preparation information and code**]()

## Livelihoods & Economies: Economies

[**Full data preparation information and code**]()

## Sense of Place: Iconic Species

[**Full data preparation information and code**]()

## Sense of Place: Lasting Special Places

[**Full data preparation information and code**]()

## Clean Waters: Nutrients

```{r, child='goals/NUT.Rmd', results='asis', echo=FALSE}
```

[**Full data preparation information and code**](https://github.com/OHI-Science/bhi/blob/draft/baltic2015/prep/CW/nutrient/nutrient_prep.Rmd)


## Clean Waters: Trash

```{r, child='goals/TRA.Rmd', results='asis', echo=FALSE}
```

[**Full data preparation information and code**](https://github.com/OHI-Science/bhi/blob/draft/baltic2015/prep/CW/trash/tra_prep.md)


## Clean Waters: Contaminants

```{r, child='goals/CON.Rmd', results='asis', echo=FALSE}
```

[**Full data preparation information and code**](https://github.com/OHI-Science/bhi/blob/draft/baltic2015/prep/CW/contaminants/contaminants_prep.Rmd)


## Biodiversity

[**Full data preparation information and code**]()
