---
title: "secchi_prep"
output:
  github_document:
    toc: true
    toc_depth: 3
params: 
    datasource: csv
---

# Nutrient subgoal data layer prep - Secchi data

```{r setup}
## source common libraries, directories, functions, etc
source('~/github/bhi/baltic2015/prep/common.r')
dir_cw    = file.path(dir_prep, 'CW')
dir_secchi    = file.path(dir_prep, 'CW/secchi')

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_secchi, 'secchi_prep.rmd')
```

## 1. Background on using Secchi

[HELCOM Water Clarity Core Indicator](http://www.helcom.fi/baltic-sea-trends/indicators/water-clarity)  Mean Summer Secchi (June-September)  
**HELCOM Good Environmental Status**
"Good environmental status is measured in relation to scientifically based and commonly agreed sub-basin-wise target levels.  

These GES boundaries were based on the results obtained in the TARGREV project (HELCOM 2013a), taking also advantage of the work carried out during the EUTRO PRO project (HELCOM 2009) and national work for WFD. The final targets were set through an expert evaluation process done by the intersessional activity on development of core eutrophication indicators (HELCOM CORE EUTRO) and the targets were adopted by the HELCOM Heads of Delegations 39/2012."  

[Approaches and methods for eutrophication target setting in the Baltic Sea region](http://www.helcom.fi/Documents/Ministerial2013/Associated%20documents/Background/Eutorophication%20targets_BSEP133.pdf)

[Fleming-Lehtinen and Laamanen. 2012. Long-term changes in Secchi depth and the role of phytoplankton in explaining light attenuation in the Baltic Sea. Estuarine, Coastal, and Shelf Science 102-103:1-10](http://www.sciencedirect.com/science/article/pii/S0272771412000418)  

[EUTRO-OPER](http://helcom.fi/helcom-at-work/projects/eutro-oper/) Making HELCOM Eutrophication Assessments opestatusnal.  Check here for HELCOM status calculations and assessment. Included on this page is a link to the [Eutrophication Assessment Manual](http://helcom.fi/Documents/Eutrophication%20assessment%20manual.pdf)  


## 2. Secchi Data

### 2.1 Data sources
**ICES**  
Data extracted from database and sent by Hjalte Parner.  
* "extraction from our database classified into HELCOM Assessment Units – HELCOM sub basins with coastal WFD water bodies or water types"

**SMHI**  
Downloaded from [SMHI Shark database](http://www.smhi.se/klimatdata/oceanografi/havsmiljodata/marina-miljoovervakningsdata) on 23 February 2016 by Lena Viktorsson.  
* Download notes: datatyp: Physical and Chemical; Parameter: secchi depth  
Lena  did not exclude any data when she downloaded it.

### 2.2 Data Cleaning and decision-making
**Duplicates in the Data**  
ICES data contains profile data (eg temperature,but secchi is only measured once). Need only unique secchi records.  It appears the SMHI also contains profiles. Also check to see if any SMHI data already in the ICES records.

**Coastal data**  
    - non-coastal (offshore) data are flagged with the code "0" under the column *"HELCOM_COASTAL_CODE"*  
        - HOLAS basin shape files with coastal and non-coastal areas were overlayed with the secchi sampling locations, all locations were flagged with a code indicating coastal or offshore.  
    - Coastal data are removed from the analysis.  
    - This should result in a similar dataset used as Fleming-Lehtinen and Laamanen 2012 (see map).  

**Sampling frequency**
*Can / should these data decisions be implemented?  We have not implemented this so far*  
Fleming-Lehtinen and Laamanen (2012) do the following:  
    - If several observations were made on the same day in the vicinity of one another, they set max observation to 1 per day.  
    - If trips were made with objective to study seasonal algae blooms, a maximum of two observations were accepted to avoid bias.  
    
## 3. Data Prep 

### 3.1 Read in data
Prelimary filtering to remove duplicate values within datasets (eg profiles) and between datasets.  
    - Check of initial smhi and ices datasets for observations from BHI regions 5 & 6 (The Sound) - region 5 all have a coastal designation. ICES data for region 6 includes both coastal and offshore but only 3 data points offshore from 2000- present. SMHI data contains no observations from region 6.

Some data points do not have a BHI ID assigned. Appears that almost all have a coastal code that != 0. If they are very close to the coast - they may fall outside the BHI shapefile. Because we exclude all sites that are coastal - only one site will have to have BHI added manually.
```{r data load}
## read in secchi data
data1 = readr::read_csv(file.path(dir_secchi, 'secchi_data_database/ices_secchi.csv'))
data2 = readr::read_csv(file.path(dir_secchi, 'secchi_data_database/smhi_secchi.csv'))

## Data overview
dim(data1)
colnames(data1)
str(data1)

dim(data2)
colnames(data2)
str(data2)

## Initial filtering
ices <- data1 %>% data.frame()%>%
  dplyr::select(bhi_id= BHI_ID, secchi, year= Year, month= Month, 
         lat= Latitude, lon = Longitude, 
         cruise= Cruise, station = Station, date= Date, coast_code=HELCOM_COASTAL_CODE) %>%
  mutate(date = as.Date(date, format= "%Y-%m-%d"))%>%
  mutate(supplier = 'ices')
head(ices)


##which ices data have BHI_ID of NA
ices.na <- ices %>%
           filter(is.na(bhi_id))
  
    dim(ices.na) # 1684   11
    ices.na.loc = ices.na %>% dplyr::select(lat,lon) %>% distinct() ## unique locations
    dim(ices.na.loc) # 86  2
    ices.na %>% dplyr::select(coast_code)%>% distinct()  ## at least one location is off shore
    ices.na %>% filter(coast_code==0)

    ## will need to manally add BHI id for site with NA and coastal code of 0



smhi <- data2 %>% data.frame()%>%
  rename(secchi = value) %>%
  dplyr::select(bhi_id= BHI_ID, secchi, year= Year, month= Month, 
        lat= Latitude, lon= Longitude, 
         cruise = Provtagningstillfaelle.id, 
         station = Stationsnamn, date= Date, coast_code=HELCOM_COASTAL_CODE) %>%
  mutate(supplier = 'smhi', cruise = as.character(cruise))
head(smhi)

## is na smhi
smhi.na <- smhi %>%
           filter(is.na(bhi_id))
  
    dim(smhi.na) # 35034    11
    smhi.na.loc = smhi.na %>% dplyr::select(lat,lon) %>% distinct() ## unique locations
    dim(smhi.na.loc) #615   2
    smhi.na %>% dplyr::select(coast_code)%>% distinct()  ## none are offshore
    smhi.na %>% filter(coast_code==0)
    ## no coastal code of zero




## Look for duplicate data

## is any data duplicated in ices itself
ices.duplicated = duplicated(ices)
sum(ices.duplicated==TRUE) #181855  ## MANY duplicates 

ices.duplicated = duplicated(dplyr::select(ices,-station))
sum(ices.duplicated==TRUE) #181977 ## more duplicated when remove station columns
    ## it is not because of multiple cruises on same day and location
    ## tried by removing lat and lon and keeping station, fewer duplicates detected

## duplicates because ICES table includes deptp
new_ices = unique(dplyr::select(ices,-station)); nrow(new_ices)  #take only unique records # 33566


## is any data duplicated in smhi itself
smhi.duplicated = duplicated(dplyr::select(smhi, -station))
sum(smhi.duplicated==TRUE) #85691 ## MANY duplicates  ## removing station does not affect it
new_smhi = unique(dplyr::select(smhi, -station)); nrow(new_smhi) #take only unique records # 17099

## use setdiff() to indentify data smhi not in ices
new_smhi = setdiff(dplyr::select(new_smhi,-supplier,-cruise), dplyr::select(new_ices,-supplier,-cruise)) %>%
            mutate(supplier = "smhi")
nrow(new_smhi) #  16627
## it appears 461 records are duplicates (if remove cruise and station)
## if date, lat, lon, secchi all match, I think they are duplicates

## Now create a new allData, bind only the new_smhi object to ices
allData = bind_rows(new_ices,new_smhi)
nrow(allData) # 50193
allData %>% dplyr::select(year, month, date, cruise, lat, lon,secchi) %>% distinct() %>%nrow(.)  #50193

## what if remove cruise
allData %>% dplyr::select(year, month, date, lat, lon,secchi) %>% distinct() %>%nrow(.)
# 50193

```

### 3.2 Remove coastal observations
Select only data with coastal code "0"  
```{r  remove coastal data points}
dim(allData) #[1]  50193    10

## Do any observations have NA for coast_code
allData %>% filter(is.na(coast_code) & is.na(bhi_id)) %>% dim() # 3567   10
allData %>% filter(is.na(coast_code) & !is.na(bhi_id)) %>% dim() #  3 10
  
  ## 3567 observations with no coast_code or BHI_ID, all from SMHI, are 292 distinct locations
      loc_no_coastcode_nobhi =allData %>% 
                                filter(is.na(coast_code) & is.na(bhi_id))%>%
                                dplyr::select(lat,lon)%>%
                                  distinct()
     
        ## check locations
         library('ggmap')
        map = get_map(location = c(8.5, 53, 32, 67.5))

         plot_map1 = ggmap(map) +
          geom_point(aes(x=lon, y=lat), data=loc_no_coastcode_nobhi,size = 2.5)
      
          plot_map1
      ## these locations are very coastal or outside of the Baltic Sea
    
  ##3 observations with NA for the coast_code but have BHI_ID
     loc_no_coastcode_bhi =  allData %>% 
                              filter(is.na(coast_code) & !is.na(bhi_id)) %>% 
                              dplyr::select(lat,lon)%>%
                              distinct()
      
     plot_map2 = ggmap(map) +
      geom_point(aes(x=lon, y=lat), data=loc_no_coastcode_bhi,size = 2.5)
      
      plot_map2
      ## these are clearly coastal stations
  
      
## What are coastal codes for The Sound (BHI regions 5,6)
##Region 6
allData %>% filter(bhi_id %in% 6) %>% dplyr::select(bhi_id,year,date,lat, lon,coast_code, supplier)%>% arrange(desc(year))%>%distinct(.)
    ## three observations in BHI region 6 after 2000
    ## Not summer observations
allData %>% filter(bhi_id %in% 6) %>% dplyr::select(coast_code, supplier)%>%distinct(.)

#Region 5
allData %>% filter(bhi_id %in% 5) %>% dplyr::select(bhi_id,year,lat, lon,coast_code, supplier)%>% arrange(desc(year))%>%distinct(.)

allData %>% filter(bhi_id %in% 5) %>% dplyr::select(coast_code, supplier)%>%distinct(.)
    ## All region 5 codes are coastal



## Filter data that are only offshore, coast_code == 0
allData = allData %>% filter(coast_code==0)

dim(allData)#14019    10 

## This is a substantial reduction in the number of observations

## find data points without BHI ID
allData %>% filter(is.na(bhi_id))  ##manual check is just barely within Latvian EEZ so is region 27

allData = allData %>%
          mutate(bhi_id = ifelse(is.na(bhi_id),27, bhi_id))
allData %>% filter(is.na(bhi_id))  

```


### 3.3 Target values
These are the values that will be used as a reference point.
```{r helcom target secchi}
target <- readr::read_csv(file.path(dir_secchi, "eutro_targets_HELCOM.csv"))
head(target)

#select just summer_seccchi target
target = target %>% dplyr::select(basin, summer_secchi)%>%
        mutate(basin = str_replace_all(basin,"_"," "))

```

### 3.4 HELCOM HOLAS Basin
These basins are the relevant physical units.  
Secchi data will be first assessed at this level and then assigned to BHI region. EEZ divisions may result in some BHI regions that have no data but they are physically the same basin as a BHI region with data.

```{r basin lookup}
basin_lookup = read.csv(file.path(dir_secchi,'bhi_basin_country_lookup.csv'), sep=";")
basin_lookup=basin_lookup %>% dplyr::select(bhi_id = BHI_ID, basin_name=Subbasin)

```


### 3.5 dplyr::select summer data and plot
    - Months 6-9 (June, July, August, September)  
    - Years >= 2000  
    - Data is sparse for BHI regions 4,22,25  
    - No data BHI regions 5 (all coastal), 6 (offshore observations rare after from 2000 and not in summer).  
```{r dplyr::select summer data}
summer = allData %>% filter(month %in%c(6:9)) %>%
        filter(year >=2000)
head(summer)



#Plot
ggplot(summer) + geom_point(aes(month,secchi, colour=supplier))+
  facet_wrap(~bhi_id, scales ="free_y")

ggplot(summer) + geom_point(aes(year,secchi, colour=supplier))+
  facet_wrap(~bhi_id)


```

### 3.6 Assign secchi data to a HOLAS basin
    - Data coverage appears better at the basin scale.  
    - With coastal data excluded, there are **no data points observed in the The Sound**  
    - Some basins have missing data or limited data for the most recent years: Aland Sea, Great Belt, Gulf of Riga, Kiel Bay, The Quark  
```{r assign summer data to a HOLAS basin}

summer = summer %>% full_join(., basin_lookup, by="bhi_id")

#Plot
ggplot(summer) + geom_point(aes(month,secchi, colour=supplier))+
  facet_wrap(~basin_name)

ggplot(summer) + geom_point(aes(year,secchi, colour=supplier))+
  facet_wrap(~basin_name)


```


### 3.7 Restrict data to before 2014
    - The Sound does not appear in the plot - No Data
    - There are still basins with limited or not data from 2010 onwards (*Great Belt*) but this at least removes the potential for not having data reported in the past 2 years
```{r restrict data before 2014}
summer = summer %>% filter(year < 2014)

#Plot
ggplot(summer) + geom_point(aes(year,secchi, colour=supplier))+
  facet_wrap(~basin_name, scales ="free_y")


## SAVE DATA FOR VISUALIZE

nut_space_data = summer %>%
                 dplyr::select(lat,lon)%>%
                 distinct()%>%
                 mutate(data_descrip = "summer secchi unique sampling locations 2000-2013",
                        bhi_goal = "NUT")

write.csv(nut_space_data, file.path(dir_baltic,'visualize/nut_space_data.csv'),row.names=FALSE)

nut_time_data = summer %>%
                dplyr::select(rgn_id=bhi_id,basin=basin_name,year,variable=month,value=secchi)%>%
                mutate(unit="secchi depth m",
                       data_descrip = "summer secchi observations",
                       bhi_goal ="NUT")


write.csv(nut_time_data, file.path(dir_baltic,'visualize/nut_time_data.csv'),row.names=FALSE)

```

### 3.8 Evaluate number of stations sampled in each basin
Very different number of unique lat-lon locations by month and basin.  
Sometimes lat-lon is not good to use because recording specific ship location which might be vary even though ship is at the same station. More duplicates were detected in the data however when station was not included, than when lat and lon were not included as the location identifier.  


```{r samples and stations by basin}

basin_summary = summer %>% group_by(basin_name,year,month)%>%
                dplyr::select(year, month,lat,lon,basin_name)%>%
                summarise(loc_count = n_distinct(lat,lon))
basin_summary

#plot sampling overview
ggplot(basin_summary) + geom_point(aes(year,loc_count, colour=factor(month)))+
  facet_wrap(~basin_name, scales ="free_y")+
  ylab("Number Sampling Locations")



```


### 3.9 Mean secchi Calculation

### 3.9.1 Calculate mean monthly value for each summer month 
basin monthly mean = mean of all samples within month and basin


```{r calculate monthly mean secchi}

mean_months = summer %>% dplyr::select(year, month,basin_name,secchi)%>%
              group_by(year,month,basin_name)%>%
              summarise(mean_secchi = round(mean(secchi,na.rm=TRUE),1))%>%
              ungroup()
head(mean_months)

```

### 3.9.2 Plot mean monthly value by basin
    - Limited July sampling in a number of basins 
    - The Quark only sampled in June in early part of time series, August later half of time series.  
    
```{r plot mean monthly}
#Plot
ggplot(mean_months) + geom_point(aes(year,mean_secchi, colour=factor(month)))+
  geom_line(aes(year,mean_secchi, colour=factor(month)))+
  facet_wrap(~basin_name)+
  scale_y_continuous(limits = c(0,10))

```


### 3.9.3 Calculate summer mean secchi (basin)
basin summer mean = mean of basin monthly mean values
```{r calculate mean summer secchi}

mean_months_summer = mean_months %>% dplyr::select(year, basin_name,mean_secchi) %>%
                      group_by(year,basin_name)%>%
                      summarise(mean_secchi = round(mean(mean_secchi,na.rm=TRUE),1)) %>%
                      ungroup()  #in mean calculation all some months to have NA, ignore for that years calculation

```

### 3.9.4 Plot summer mean secchi
```{r plot mean summer secchi}
ggplot(mean_months_summer) + geom_point(aes(year,mean_secchi))+
  geom_line(aes(year,mean_secchi))+
  facet_wrap(~basin_name)+
  scale_y_continuous(limits = c(0,10))
```



### 3.9.5 Plot summer secchi with target values indicated
Horizontal lines are HELCOM target values.  
```{r summer secchi with target }

secchi_target = left_join(mean_months_summer,target, by=c("basin_name" = "basin"))%>%
                dplyr::rename(target_secchi = summer_secchi)
head(secchi_target)

ggplot(secchi_target) + geom_point(aes(year,mean_secchi))+
  geom_line(aes(year,target_secchi))+
  facet_wrap(~basin_name)+
  scale_y_continuous(limits = c(0,10))

```


## 4. Status and Trend explostatusn

### 4.1 Goal Model & Trend
**The Goal Model is how the status is calculated**  
Xnut_b = mean_summer_secchi_y_b / Reference point_r

mean_summer_secchi_y = mean summer secchi in year (y) in a basin (b)
Reference point = HELCOM target for that basin   

Xnut_bhi_region = Xnut_b  
*Each BHI region will receive the status value of the HOLAS basin it belongs to*  

**Trend**
The trend is calculated based on a linear regression.  
Status_b ~ m*Year + intercept  
In our approach here, we use a 10 year period to calculate the trend with a minimum of 5 data points.  In most cases, the BHI framework uses a 5 year period for the trend, but as secchi is a slow response variable, we use a longer time period.  
The trend value is the slope (m) of the linear regression multiplied by the year of future interest (5 years from status year) and this value is constrained to between -1 and 1.  


### 4.2 What year will the status be calculated for for each basin?
    - This is if there is no temporal gapfilling.  
    - Most basins can have the status calculated for **2013** with the **exception** of: *Aland Sea (2012)*, *Great Belt (2011)*, *Gulf of Finland (2012), *Gulf of Riga (2012)*, *The Quark (2012)*  
    - No data for the Sound  
    - One option is no gap filling and calculating the status for differenet final years and over a different 5 year period
```{r what is the last year}
## get the last year of non-NA data
last_year = secchi_target%>%
            filter(!is.na(mean_secchi))%>%
            group_by(basin_name)%>%
            summarise(last_year = last(year)) %>%
            print(n=15)

##which are not in 2013
last_year %>% filter(last_year < 2013)

 
```

### 4.3 Calculate status
**Status calculation with raw (non-modeled) mean summer secchi by basin**  
Status must be calculated in data prep because calculation for a basin and then applied to all regions.  
*Status code based on code Lena Viktorsson developed for functions.r*

```{r Status calculation}

## Define constants for status calculation

  min_year = 2000        # earliest year to use as a start for regr_length timeseries
                          ##data already filtered for 
  regr_length = 10       # number of years to use for regression
  future_year = 5        # the year at which we want the likely future status
  min_regr_length = 5    # min actual number of years with data to use for regression.

  
## Basin data with target
  secchi_target
  
  
## Calculate basin status
  ## Xnut = basin_mean/basin_target
  
  basin_status = secchi_target %>%
                 mutate(., status =  pmin(1, mean_secchi/target_secchi)) %>%
      dplyr::select(basin_name, year, status)
  
## Calculate basin trend
  
  basin_trend =
    basin_status %>%
    group_by(basin_name) %>%
    do(tail(. , n = regr_length)) %>%  # calculate trend only if there is at least X years of data (min_regr_length) in the last Y years of time serie (regr_length)
    do({if(sum(!is.na(.$status)) >= min_regr_length)
    data.frame(trend_score = 
                 max(-1, min(1, coef(lm(status ~ year, .))['year'] * future_year)))
         else data.frame(trend_score = NA)}) %>%
    ungroup() 

  
  ## Assign basin status and trend to BHI regions
    bhi_status = basin_status %>%
                group_by(basin_name)%>%
                summarise_each(funs(last), basin_name, status)%>% #select last year of data for status in each basin (this means status year differs by basin)
                mutate(status = round(status*100))%>% #status is whole number 0-100
                ungroup()%>%
                left_join(basin_lookup,.,by="basin_name")%>% #join bhi regions to basins
                mutate(dimension = 'status') %>%
                dplyr::select(rgn_id = bhi_id, dimension, score=status)
                
  
    bhi_trend = left_join(basin_lookup,basin_trend, by="basin_name") %>%
                 mutate(score = round(trend_score,2),
                        dimension = "trend")%>%
                dplyr::select(rgn_id = bhi_id, dimension, score )
               


```

### 4.4 Plot Basin status over time
Basin status is initially a value between 0 - 1.  Calculated for each year between 2000 and 2013.  
```{r plot basin status}

ggplot(basin_status) + geom_point((aes(year,status)))+
  facet_wrap(~basin_name) +
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))

```

### 4.5 Plot BHI region status and trend values
    - Status values can range from 0-100 -- this is the status for the *most recent* year. In most cases this is 2013.  
    - Trend values can be between -1 to 1
    - No status or trend for 5 or 6 - these are The Sound
    
```{r bhi status and trend plot}

ggplot(full_join(bhi_status,bhi_trend, by=c("rgn_id","dimension","score"))) + geom_point(aes(rgn_id,score),size=2)+
  facet_wrap(~dimension, scales="free_y")+
  xlab("BHI region")


```


### 4.6 Save csv files
These csv files will be used as a first cut for the secchi status and trend.
**Files to save**  
1. Status and trend for each BHI region based on the basin level calculations
```{r write csv file to layers}
## Write csv files to layers
    readr::write_csv(bhi_status, 
                 file.path(dir_layers, "cw_nu_status_bhi2015.csv"))
    
    readr::write_csv(bhi_trend, 
                 file.path(dir_layers, "cw_nu_trend_bhi2015.csv"))

```


## 5. Next steps

**Trend calculation**  
1. Have calculated the trend using data spanning 10 years (minimum of 5 data points). Is it agreed that we should use the longer time window for the trend?  

**No Observations for The Sound**  
1. The Sound has no summer offshore observations after 2000 in either ICES or SMHI data.  
    - How is it evaluated by HOLAS?  
    - BHI regions are 5 & 6.  5 is all coastal, 6 has both coastal and offshore observations before 2000 and from 2000 forward, 3 offshore observations in spring (not summer).  
  
**Should/How to model the data?**  
Not all months are sampled in all years (see above plot).  
If model, do a linear model.  
*Linear model options*
1. Take mean summer secchi, ignore that different months sampled in different years, just average the months that are sampled in any given year.  Model by basin and year.  
2. Take mean monthly value by year, model by basin + year + month. Average the modelled monthly value to get a summer mean.  
3. Do the above but just model all the data points, don't take the mean value and instead use a random effect to account for location?  



# 6. Rescale Mean secchi data and recalculate status

Rescale secchi depth to target level and change the goal model equation to:  

Xnut_b = (mean_summer_secchi_y_b - min_secchi_b) / (Reference point_r - min_secchi_b)

mean_summer_secchi_y = mean summer secchi in year (y) in a basin (b)
min_secchi_b = minimum mean summer secchi in a basin (b)
Reference point = HELCOM target for that basin   

Xnut_bhi_region = Xnut_b  
*Each BHI region will receive the status value of the HOLAS basin it belongs to*  

```{r testing non linear weighting, warning = FALSE}

## non-linear transformation: decay and logisitic

## decay: 
## high interests in general. scores are always disproprotionally low. 
## status_1 & status_2 = ratio^m
## m_1 = 2
## m_2 = 4

## logistic:
## status_3 & status_4 = 1/(1 + exp( -(ratio - a)/b))
## a_3 = 0.5, b_3 = 0.1 -> medium-interests: water quality considered bad when secchi depth <50% of target value and are given disproportionally bad scores. 
## a_4 = 0.7, b_4 = 0.08 -> high-interests: water quality considered bad when secchi depth <80% of target value and are given disproportionally bad scores. 

## general comparison of decay and logistic

basin_status_rescaled = data.frame(ratio = seq(0, 1, 0.1)) %>%
  mutate(linear = ratio,
         transformation_1 = ratio^2,
         transformation_2 = ratio^4,
         transformation_3 = 1/(1+ exp(-(ratio - 0.5)/0.1)),
         transformation_4 = 1/(1+ exp(-(ratio - 0.7)/0.08))) %>% 
  gather(key = transformation, value = status, 2:6)

plot_rescaled = qplot(x = ratio, y = status, color = transformation, data =  basin_status_rescaled, geom = "point") +
  geom_line(stat = "identity", position = "identity") +
  labs(title = 'Example non-linear transformations of Nutrients status score',
      x = 'Non-transformed status score (mean_secchi/target_secchi)', 
      y = 'Transformed status score',
      fill = 'Transformation')

print(plot_rescaled)


 ## setup to plot different tranformations using PlotMap function

 source('~/github/bhi/baltic2015/PlotMap.r')
 source('~/github/bhi/baltic2015/PrepSpatial.R') 
 #install.packages('maptools')
 #install.packages('broom')

## transformation 1: status = (mean_secchi/target_secchi) ^ 2

 basin_status_orig = secchi_target %>%
                 mutate(ratio =  pmin(1, mean_secchi/target_secchi)) %>%
      dplyr::select(basin_name, year, ratio)

 bhi_status_1 = basin_status_orig %>%
                group_by(basin_name) %>%
                summarise_each(funs(last), basin_name, ratio) %>% #select last year of data for status in each basin (this means status year differs by basin)
                mutate(status = round(ratio^2*100))%>% #status is whole number 0-100
                ungroup()%>%
                left_join(basin_lookup,.,by="basin_name")%>% #join bhi regions to basins
                # mutate(dimension = 'status') %>%
                dplyr::select(rgn_id = bhi_id, score=status)
 
 if (!require(gpclib)) install.packages("gpclib", type="source")
gpclibPermit()

 plot_transf_1 =  PlotMap(bhi_status_1, map_title = expression('Transformation 1: status = (mean/target)' ^ 2), 
                           rgn_poly        = PrepSpatial(path.expand('~/github/bhi/baltic2015/spatial/regions_gcs.geojson')))
                          # fig_path       = path.expand('~/github/bhi/baltic2015/prep/CW/secchi/transformation_figs')) 
                      
  
 ## transformation 2: status =  (mean_secchi/target_secchi) ^ 4

  bhi_status_2 = basin_status_orig %>%
                group_by(basin_name) %>%
                summarise_each(funs(last), basin_name, ratio) %>% #select last year of data for status in each basin (this means status year differs by basin)
                mutate(status = round(ratio^4*100))%>% #status is whole number 0-100
                ungroup()%>%
                left_join(basin_lookup,.,by="basin_name")%>% #join bhi regions to basins
                # mutate(dimension = 'status') %>%
                dplyr::select (rgn_id = bhi_id, score=status)

 plot_transf_2 =  PlotMap(bhi_status_2, map_title = expression('Transforamtion 2: status = (mean/target)' ^ 4), 
                           rgn_poly = PrepSpatial(path.expand('~/github/bhi/baltic2015/spatial/regions_gcs.geojson')))


 ## transformation 3: status = 1/(1+ exp(-(mean/target - 0.5)/0.1))

 bhi_status_3 = basin_status_orig %>%
                group_by(basin_name) %>%
                summarise_each(funs(last), basin_name, ratio) %>% #select last year of data for status in each basin (this means status year differs by basin)
                mutate(status = round(1/(1+ exp(-(ratio - 0.5)/0.1)) *100))%>% #status is whole number 0-100
                ungroup()%>%
                left_join(basin_lookup,.,by="basin_name")%>% #join bhi regions to basins
                # mutate(dimension = 'status') %>%
                dplyr::select (rgn_id = bhi_id, score=status)

 plot_transf_3 =  PlotMap(bhi_status_3, 
                          map_title = 'Transformation 3: \n status = 1/(1+ exp(-(mean/target - 0.5)/0.1))', 
                          rgn_poly = PrepSpatial(path.expand('~/github/bhi/baltic2015/spatial/regions_gcs.geojson')))



## transformation 4: status = 1/(1+ exp(-(mean/target - 0.7)/0.08))

 bhi_status_4 = basin_status_orig %>%
                group_by(basin_name) %>%
                summarise_each(funs(last), basin_name, ratio) %>% #select last year of data for status in each basin (this means status year differs by basin)
                mutate(status = round(1/(1+ exp(-(ratio - 0.7)/0.08)) *100))%>% #status is whole number 0-100
                ungroup()%>%
                left_join(basin_lookup,.,by="basin_name")%>% #join bhi regions to basins
                # mutate(dimension = 'status') %>%
                dplyr::select (rgn_id = bhi_id, score=status)

 plot_transf_4 =  PlotMap(bhi_status_4,  
                          map_title = 'Transformation 4: \n status = 1/(1+ exp(-(mean/target - 0.7)/0.08))', 
                           rgn_poly = PrepSpatial(path.expand('~/github/bhi/baltic2015/spatial/regions_gcs.geojson')))


  
```
