---
title: "climate_change_prep"
output: github_document
params: 
    datasource: csv
---

# Climate Change Pressure Layers
## Raw data cleaning

```{r setup}

## Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')

source('~/github/bhi/baltic2015/prep/common.r')

## rprojroot
root <- rprojroot::is_rstudio_project


## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)



dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')


# root$find_file("README.md")
# 
# root$find_file("ao_need_gl2014.csv")
# 
# root <- find_root_file("install_ohicore.r", 
# 
# withr::with_dir(
#   root_file("DESCRIPTION"))



dir_cc    = file.path(dir_prep, 'pressures/climate_change')


## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_cc, 'climate_change_prep.rmd') 
```


## 1. Background 
The direct effects of climate change in the abiotic environment will be changes to water temperature and salinity. In the BHI framework, we will include sea surface temperature (SST), surface salinity (SS), and bottom water salinity (BWS) as climate change pressures.

## 2. Data

### 2.1 Data source
Data are from the [BALTSEM model](http://www.balticnest.org/balticnest/thenestsystem/baltsem.4.3186f824143d05551ad20ea.html), run by Bärbel Müller Karulis from the Baltic Sea Centre at Stockholm University.

### 2.2 Data cleaning and extraction
Data from hindcast and future scenarios by basin need to be organized into separate SST and salinity files. These data will then be prepared in the subfolder 'temperature_climatechange' and 'salinity_climatechange'



## 3. Data organization

### 3. Read in all data
```{r read in data}
## read in data...

## plot data...

```
