---
title: "ao_prep.rmd"
output: github_document
params: 
    datasource: csv
---
# Artisanal Opportunity (AO) Goal Data Prep

## Goal Overview

### Components
This goal has three sub-components: *stock, access, and need*. For BHI we focus first on the *stock* sub-component and will use this as a proxy for the entire goal initially

### Goal model
#### Status  
Xao = Mean Stock Indicator Value /  Reference pt  
Stock indicators = two HELCOM core indicators assessed for good environemental status (each scored between 0 and 1 by BHI)  
Reference pt = maximum possible good environmental status (value=1)  

#### Trend
**Background**  
CPUE time series are available for all stations used for the HELCOM coastal fish populations core indicators. These data were provided by Jens Olsson (FISH PRO II project). To calculate GES status, full time series were used. Therefore, only one status time point and cannot calculate trend of status over time. Instead, follow approach from Bergström et al 2016, but only focus on the final time period for the slope (2004-2013).  

[Bergstrom et al. 2016. Long term changes in the status of coastal fish in the Baltic Sea. Estuarin, Coast and Shelf Science. 169:74-84](http://www.sciencedirect.com/science/article/pii/S0272771415301700http://www.sciencedirect.com/science/article/pii/S0272771415301700)  

**Trend Method**
1.Select final time period of trend assessment (2004-2013)  
2. Use time series from both indicators, Key Species and Functional groups. For functional groups,include both cyprinid and piscivore time series.  
3. For each time series:  square-root transform data, z-score, fit linear regression, extract slope  
4. Within each time series group (key species, cyprinid, piscivore), take the mean slope for each group within each basin  
5. Within each basin take a mean functional group indicator slope (mean of cyprinid mean and piscivore mean)  
6. For each basin take overall mean slope - mean of key species and functional group  
7. Multiple by five for future year value?  
8. Apply trend value for basin to all BHI regions (except in Gulf of Finland, do not apply Finnish site value to Estonia and Russian regions.)  
*Steps 1-3 done in file `AO/ao_slope_calc.r`


## Data
[HELCOM Core Indicator Abundance of coastal fish key functional groups](http://helcom.fi/baltic-sea-trends/indicators/abundance-of-coastal-fish-key-functional-groups/)  

[HELCOM Core Indicator Abundance of key coastal fish species](http://helcom.fi/baltic-sea-trends/indicators/abundance-of-key-coastal-fish-species)  
Good Environmental Status (GES) is assessed as either *GES* or *sub-GES* based on data times series using either a baseline or a trend approach, [see explanation](http://helcom.fi/baltic-sea-trends/indicators/abundance-of-key-coastal-fish-species/good-environmental-status/). There is only a single assessment for each region.  

*status qualifying comments*: for one dataset if a monitoring station receives a "sub-GES" assessment, it is given a qualifier as "low" or "high".  

Environmental status assessments provided by Jens Olsson (SLU).  See [HELCOM FISH-PRO II](http://www.helcom.fi/helcom-at-work/projects/fish-pro/)  

CPUE data used in the GES assessment. Data provided by Jens Olsson, used in trend. Slopes from each analsysis available here, but CPUE data held internally in the BHI database.  

### Data locations
Data are from monitoring locations (described in the HELCOM core indicators). Finnish data are fisheries data from ICES assessment regions (ICES 29-32).

## Data Prep
Data prep below.  

### Data scoring
We will have to determine appropriate 0-1 scale for *GES* and *sub-GES*.  

*Note that in the OHI framework 'NA' means that an indicator is not applicable, not that there is no data. If there is no data, should either gap-fill or assign a score of 0 to highlight missing information*  


```{r setup, echo=FALSE}

## source common libraries, directories, functions, etc
source('~/github/bhi/baltic2015/prep/common.r')

## Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')

## rprojroot
root <- rprojroot::is_rstudio_project

## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)

dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')


# root$find_file("README.md")
# 
# root$find_file("ao_need_gl2014.csv")
# 
# root <- find_root_file("install_ohicore.r", 
# 
# withr::with_dir(
#   root_file("DESCRIPTION"))


dir_ao    = file.path(dir_prep,'AO')


## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_ao, 'ao_prep.rmd')
```


## Layer prep

## 1. Read in data
Read in status assessment, monitoring area locations, lookup table for BHI regions to HOLAS basins.  

Locations for the Finnish monitoring areas are not given by specific lat-lon because are assessment areas. I have assigned a single lat-lon for each Finnish monitoring area for plotting purposes.  
```{r read in data}
## read in data...
  ## data are in "ao_data_database" - currently csv loaded directly there, need to put in the database and then set up script to extract

#assessment of GES status, all 2 indicators
coastal_fish = readr::read_csv2(file.path(dir_ao, 'ao_data_database/ao_coastalfish_ges_status.csv'))

head(coastal_fish)
dim(coastal_fish)
str(coastal_fish)

#locations of specific monitoring stations
coastal_fish_loc = read.csv(file.path(dir_ao, 'ao_data_database/ao_coastalfish_locations.csv'),
                          sep=";") ## can't use read_csv2 because while includes sep = ";" as the default, also has the default decimal separator as a comma so does not know what to do with a period

head(coastal_fish_loc)
dim(coastal_fish_loc)
str(coastal_fish_loc)        

## Station cleaned is a column that I created to make the station names match those under "monitoring_area" in the status asssessment

## assign lat-lon for ICES area to allow map plotting - assign location in center of Finnish coasts for ICES areas (29-32)

##ICES 31
 ##64.207281
 ##23.511161
##ICES 30
 ##61.679815
 ##21.269112
##ICES 29
 ##59.705518
 ##21.264879
##ICES 32
 ##60.108130
 ##25.772410
##Finnish Rectangle 23 & 28 (The Quark)
 ##63.505817
 ##21.351243

coastal_fish_loc = coastal_fish_loc%>%
                    mutate(lat = DECWGSN,
                           lon = DECWGSE)%>%
                    mutate(lat = ifelse(station_cleaned== "ICES SD 31", 64.207281,
                                 ifelse(station_cleaned== "ICES SD 30", 61.679815,
                                 ifelse(station_cleaned== "ICES SD 29", 59.705518,
                                 ifelse(station_cleaned== "ICES SD 32", 60.108130,
                                 ifelse(station_cleaned== "Rectangle 23 & 28",  63.505817,lat ))))),
                           lon = ifelse(station_cleaned== "ICES SD 31", 23.511161,
                                 ifelse(station_cleaned== "ICES SD 30", 21.269112,
                                 ifelse(station_cleaned== "ICES SD 29", 21.264879,
                                 ifelse(station_cleaned== "ICES SD 32", 25.772410,
                                 ifelse(station_cleaned== "Rectangle 23 & 28",  21.351243,lon ))))))%>%
                    mutate(station_cleaned = as.character(station_cleaned),
                           station = as.character(station))
str(coastal_fish_loc)

#bhi region and HOLAS basin look up
 basin_lookup = readr::read_csv(file.path(
  dir_ao,"baltic_rgns_to_bhi_rgns_lookup_holas.csv"))
basin_lookup=basin_lookup %>% select(bhi_id = rgn_id, basin_name)%>%
  mutate(basin_name = str_replace_all(basin_name,"_"," ")) 
basin_lookup

```


## 2. Assign scores to GES status
Explore the consequences of different scoring schemes.  
In all cases, a score of 1 achieving highest status.  

### 2.1 Alternative scoring methods
**score"**: GES = 1, subGES = 0 *If value does not meet GES threshold so recieves 0"*  

**score2**: GES = 1, subGES = 0.2 *If value does not meet GES threshold but have data to assess status receive score of 0.2. This way, if regions are not assessed and use 0 for these regions, a distinction is made (NA in the OHI framework means "indicator not applicable", not "no data").*  

**score3**: GES =1, subGES (low) = 0.2, subGES(high) = 0.5, subGES (no comment) = 0.2. *Distinguish between subGES levels that have been ranked low or high. Has only been done for cyprinids functional group.*  

```{r assign scores}
## is status ever NA?
coastal_fish %>% filter(is.na(status)) #No

## Assign three alternative 0-1 scores
  ## score 1:  GES =1, subGES = 0
  ## score 2:  GES =1, subGES = 0.2
  ## score 3:  GES =1, subGES (low)= 0.2, subGES (high)=0.5, subGES = 0.2

coastal_fish_scores = coastal_fish %>% 
                      mutate(score1 = ifelse(status== "GES",1,0),
                             score2 = ifelse(status=="GES",1,.2),
                             score3 = ifelse(status== "GES",1,
                                      ifelse(status=="subGES" & is.na(status_comment)==TRUE,.2,      
                                      ifelse(status=="subGES" & status_comment == "Low",.2,
                                      ifelse(status=="subGES" & status_comment == "High",.5,0)))))
                                     
                                     

coastal_fish_scores


```

### 2.2 Plot alternative scores by location
Three separate plots for alternative scoring methods.  

The difference between **score1** and **score2** / **score3** is that **score1** is binary between 0 and 1, while **score2** is binary between 0.2 and 1. **Score3** is only slightly different from **score 2** where for the functional groups (cyprinids) some receive a slightly higher score as "high" subGES stats.  

```{r plot alt scores}

# make coastal_fish_scores in long format then plot
temp_long = coastal_fish_scores %>% 
            select(monitoring_area,core_indicator, score1,score2,score3) %>%
            group_by(monitoring_area, core_indicator) %>%
            gather(score_type,score,score1,score2,score3)%>%
            ungroup()

#Score 1
ggplot(filter(temp_long, score_type=="score1")) + 
  geom_point(aes(monitoring_area, score))+
  facet_wrap(~core_indicator)+
   theme(axis.text.x = element_text(colour="grey20",size=7,angle=90,hjust=.5,vjust=.5,face="plain"),
        plot.margin = unit(c(1,1,1,1), "cm")) +
  ggtitle("Scoring Method 1")


#Score 2
ggplot(filter(temp_long, score_type=="score2")) + 
  geom_point(aes(monitoring_area, score))+
  facet_wrap(~core_indicator)+
   theme(axis.text.x = element_text(colour="grey20",size=7,angle=90,hjust=.5,vjust=.5,face="plain"),
        plot.margin = unit(c(1,1,1,1), "cm")) +
  ggtitle("Scoring Method 2")


#Score 3
ggplot(filter(temp_long, score_type=="score3")) + 
  geom_point(aes(monitoring_area, score))+
  facet_wrap(~core_indicator)+
   theme(axis.text.x = element_text(colour="grey20",size=7,angle=90,hjust=.5,vjust=.5,face="plain"),
        plot.margin = unit(c(1,1,1,1), "cm")) +
  ggtitle("Scoring Method 3")



```


## 3. Unique indicators per monitoring location
*Summary information from code below.*  
1. Is more than one key species monitored at a given locations?  
**NO**
2. Is more than one function group monitored?  
**Depends on location, 1 or 2 groups monitored**  
3. Are both key species and functional groups monitored at all locations?
**No**, 4 monitoring areas without Functional status, 2 without Key_spp status  

```{r unique indicators monitoring region}

coastal_fish_scores_long = coastal_fish_scores %>% 
            group_by(Basin_HOLAS,Basin_assessment, country,monitoring_area,
                     period, coastal_water_type,core_indicator,taxa, assessment_method,
                     status,status_comment) %>%
            gather(score_type,score,score1,score2,score3)%>%
            ungroup()

coastal_fish_scores_long

## Number of indicators by monitoring location
indicator_count = coastal_fish_scores_long %>%
                  select(monitoring_area, core_indicator,taxa,score_type,score)%>%
                  group_by(monitoring_area)%>%
                  summarise(unique_indicator = length(unique(core_indicator)),
                            unique_taxa_func =  length(unique(taxa))) %>%
                ungroup()

indicator_count %>% print(n=60)

indicator_taxa_count= coastal_fish_scores_long %>% filter (score_type=="score1") %>% #remove duplicates based on scoring alternatives
                    select(monitoring_area, core_indicator,taxa)%>%
                  group_by(monitoring_area,core_indicator)%>%
                  summarise(unique_taxa_func =length(unique(taxa)))%>%
                  ungroup()
                  
indicator_taxa_count %>% print(n=60)

ggplot(indicator_taxa_count) + geom_point(aes(monitoring_area, unique_taxa_func))+
                              facet_wrap(~core_indicator)+
                              theme(axis.text.x = element_text(colour="grey20", size=6, angle=90, 
                                                               hjust=.5, vjust=.5, face = "plain"),
                                    plot.margin = unit(c(1,1,1,1), "cm")) 

##Number of NAs by each indicator
indicator_taxa_count %>%  spread(core_indicator,unique_taxa_func)%>% 
                          dplyr::rename(Key_species = `Key species`)%>%
                          summarise(Func_na = sum(is.na(Functional)),                                                                               KeySpp_na = sum(is.na(Key_species)))


```



## 4. BHI score
1. Take mean score for each indicator type in each monitoring region.  
2. Take mean score for each indicator at the HOLAS basin scale.  
3. Take mean of the two indicators for each basin.  
4. Apply basin score to each BHI region  

```{r status calculation}

## indicator mean by monitoring area
monitoring_indicator_mean = coastal_fish_scores_long %>% select(Basin_HOLAS, monitoring_area,
                                                                core_indicator,score_type,score)%>%
                            group_by(Basin_HOLAS, monitoring_area,core_indicator,score_type)%>%
                            summarise(mean_core_monitoring_score = mean(score, na.rm=TRUE))%>%
                            ungroup()
monitoring_indicator_mean


# indicator mean by HOLAS basin
basin_indicator_mean = monitoring_indicator_mean %>%
                        group_by(Basin_HOLAS,core_indicator, score_type)%>%
                        summarise(mean_core_basin_score = mean(mean_core_monitoring_score,na.rm=TRUE))%>%
                        ungroup()
basin_indicator_mean

#HOLAS basin score (mean across the two indicators)
basin_mean_score = basin_indicator_mean %>%
                    group_by(Basin_HOLAS,score_type)%>%
                    summarise(mean_basin_score = round(mean(mean_core_basin_score,na.rm=TRUE),2))%>%
                    ungroup()
  
basin_mean_score


## BHI score
    ## join basin lookup
  bhi_mean_score = basin_mean_score %>% full_join(.,basin_lookup, by=c("Basin_HOLAS" = "basin_name"))
  bhi_mean_score %>% print(n=45)
  
  ## Want to make sure that there are entries for each score type and basin (if not scored because no data, then only get a single NA)
      ## make a "master" list of scores and bhi_id
      score_list = basin_lookup %>% select(basin_name, bhi_id)%>% 
                                    mutate(score1 = "score1",
                                           score2 = "score2",
                                           score3 = "score3") %>%
                                    dplyr::rename(Basin_HOLAS = basin_name) %>%
                                    group_by(Basin_HOLAS,bhi_id) %>%
                                    gather(score_type,score,score1,score2,score3)%>%
                                    ungroup() %>%
                                    select(Basin_HOLAS,bhi_id,score_type)
      
      ## Join bhi_mean_score and score_list
          bhi_mean_score = bhi_mean_score %>% right_join(., score_list, by=c("Basin_HOLAS","bhi_id","score_type"))

  
```

### 4.1 Plot scores at each level
*e.g. plot the monitoring stations, basins, BHI regions*  


### 4.1a Plot monitoring area indicator mean scores
Scoring alternatives do not seem to lead to strong differences.  Clearly changing the subGES from 0 to 0.2 shifts the range of scores.  

**Note that points are "jittered" in the second plot to make visible**  

**Colors in the second plot are different monitoring locations**


```{r plot scores levels of aggregation}

##plot monitoring area indicator mean scores
ggplot(monitoring_indicator_mean) + 
  geom_point(aes(monitoring_area, mean_core_monitoring_score, color=core_indicator))+
  facet_wrap(~score_type)+
  theme(axis.text.x = element_text(colour="grey20", size=6, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"),
         plot.margin = unit(c(1,1,1,1), "cm")) 

##plot monitoring area indicator mean scores by HOLAS basin
ggplot(monitoring_indicator_mean) + 
  geom_jitter(aes(Basin_HOLAS, mean_core_monitoring_score, color=monitoring_area, shape=factor(core_indicator)))+
  facet_wrap(~score_type)+
  guides(colour=FALSE)+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"),
         plot.margin = unit(c(1,1,1,1), "cm")) 


```

### 4.1b Plot the monitoring area mean scores on a map
Each of the alternative score methods plotted on a separate map.

```{r monitoring mean score map}
##plot on map

## Join monitoring_indicator_mean to lat-lon
monitoring_indicator_mean_loc= monitoring_indicator_mean %>% left_join(., select(coastal_fish_loc, station_cleaned, lat, lon), by=c("monitoring_area" = "station_cleaned"))
str(monitoring_indicator_mean_loc)

## get the map
library('ggmap')
map = get_map(location = c(8.5, 53, 32, 67.5))


##Plot Scoring Alternative 1
    map_data1 = monitoring_indicator_mean_loc %>% select(monitoring_area, core_indicator,score_type,mean_core_monitoring_score,lat,lon ) %>% 
      filter(score_type == "score1")
    
    ##set up the plot
    plot_map1 = ggmap(map) +
      geom_point(aes(x=lon, y=lat, colour=mean_core_monitoring_score, shape=factor(core_indicator)), data=map_data1,size = 2.5)
    
    ##plot the map
    plot_map1 + scale_color_gradientn(colours=rainbow(2), limits=c(0,1)) +
      ggtitle('Coastal Fish Stock Status, Score Type 1') +
      theme(title = element_text(size = 12))


##Plot Scoring Alternative 2
    map_data2 = monitoring_indicator_mean_loc %>% select(monitoring_area, core_indicator,score_type,mean_core_monitoring_score,lat,lon ) %>% 
      filter(score_type == "score2")
    
    ##set up the plot
    plot_map2 = ggmap(map) +
      geom_point(aes(x=lon, y=lat, colour=mean_core_monitoring_score,shape=factor(core_indicator)), data=map_data2,size = 2.5)
    
    ##plot the map
    plot_map2 + scale_color_gradientn(colours=rainbow(2), limits=c(0,1)) +
      ggtitle('Coastal Fish Stock Status, Score Type 2') +
      theme(title = element_text(size = 12))



##Plot Scoring Alternative 3
    map_data3 = monitoring_indicator_mean_loc %>% select(monitoring_area, core_indicator,score_type,mean_core_monitoring_score,lat,lon ) %>% 
      filter(score_type == "score3")
    
    ##set up the plot
    plot_map3 = ggmap(map) +
      geom_point(aes(x=lon, y=lat, colour=mean_core_monitoring_score,shape=factor(core_indicator)), data=map_data3,size = 2.5)
    
    ##plot the map
    plot_map3+ scale_color_gradientn(colours=rainbow(2), limits=c(0,1)) +
      ggtitle('Coastal Fish Stock Status, Score Type 3') +
      theme(title = element_text(size = 12))

```



### 4.1c Plot Basin Indicator scores
Plot the mean basin score for each indicator *(e.g. on functional score per basin and one key species score per basin).*  
*Basins are currently alphabetically ordered on x-axis, not ordered geographically.*  
```{r plot basin indicator scores}

##basin_indicator_mean

##plot monitoring area indicator mean scores
ggplot(basin_indicator_mean) + 
  geom_point(aes(Basin_HOLAS, mean_core_basin_score, color=core_indicator,shape=core_indicator), size=1.5)+
  facet_wrap(~score_type)+
  scale_shape_manual(values=c(6,19))+
  scale_colour_manual(values = c("black","turquoise"))+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"),
         plot.margin = unit(c(1,1,1,1), "cm")) 



```

### 4.1d Plot Basin mean score across indicators
The mean of the two indicator scores was taken for each basin.

```{r plot basin mean across indicators}

##basin_mean_score

##plot monitoring area indicator mean scores
ggplot(basin_mean_score) + 
  geom_point(aes(Basin_HOLAS, mean_basin_score), size=2)+
  facet_wrap(~score_type)+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"),
         plot.margin = unit(c(1,1,1,1), "cm")) 


```



### 4.1e Plot BHI Scores
There are no scores for Kiel Bay and Gdansk Basin, there for no scores for BHI regions 7,8,18,19. For Kiel Bay, this could be because some monitoring locations should be assigned to Kiel Bay but are not. There is no Polish data which is why there is no scoring for Gdansk Basin.  
**Outcome among score type** does not seem to change at the basin or BHI scale. Should other score methods be considered?  

```{r plot BHI scores}
## BHI Data
library(rgdal)
BHIshp = readOGR("C:/Users/jgrif/Documents/StockholmUnivPostDoc/BalticHealthIndex/BHI_r/March2016WkshpPlots/shapefiles", "BHI_regions_plus_buffer_25km")
BHIshp2 = spTransform(BHIshp, CRS("+proj=longlat +init=epsg:4326"))
print(proj4string(BHIshp2))


## Assign colors to BHI ID based on score - these bins are not even, not sure how to do a gradient
## 0 - 0.19
## 0.2 - 0.49
## 0.5 - 0.74
## 0.75 - 1.0

bhi_mean_score_colors = bhi_mean_score %>% 
                        mutate(cols = ifelse(is.na(mean_basin_score) == TRUE, "grey",
                                      ifelse(mean_basin_score >= 0 & mean_basin_score < 0.2, "orange1",
                                      ifelse(mean_basin_score >= 0.2 & mean_basin_score < 0.5, "yellow2",
                                      ifelse(mean_basin_score >= 0.5 & mean_basin_score < 0.75, "light blue",
                                      ifelse(mean_basin_score >= .75 & mean_basin_score <=1.0, "blue", "grey"))))))


## Need separate shapefile for each score

shp_score1 = BHIshp2
shp_score2 = BHIshp2
shp_score3 = BHIshp2

#assign colors to shapefile data

shp_score1@data = shp_score1@data %>% full_join(., filter(bhi_mean_score_colors,score_type=="score1"), by=c("BHI_ID"= "bhi_id"))
head(shp_score1@data)

shp_score2@data = shp_score2@data %>% full_join(., filter(bhi_mean_score_colors,score_type=="score2"), by=c("BHI_ID"= "bhi_id"))
head(shp_score2@data)

shp_score3@data = shp_score3@data %>% full_join(., filter(bhi_mean_score_colors,score_type=="score3"), by=c("BHI_ID"= "bhi_id"))
head(shp_score3@data)

## Plot BHI regions colored by score value

  ## plot each score type separately

  par(mfrow=c(2,2), mar=c(.5,.2,.5,.2), oma=c(0,0,4,0))
  
 plot(shp_score1, col=shp_score1@data$cols, main = "score 1")
 plot(shp_score1, col=shp_score1@data$cols, main = "score 2")
 plot(shp_score1, col=shp_score1@data$cols, main = "score 3")

  plot(c(1,2,3),c(1,2,3), type='n', fg="white",bg="white", xaxt='n',yaxt='n')
  legend("center", 
         legend=c("No Score","0 - 0.19", "0.2 - 0.49", "0.5 - 0.74", "0.75 -1.0"), 
         fill=c("grey","orange1","yellow2","light blue", "blue"), bty='n', cex=1.5)

    mtext("AO Score", side = 3, outer=TRUE, line=1.5)


```


## Status review and decisions

1. Have Jens assess scoring options for GES status assessment - **Jens says alternative 2**

2. Have Jens assess method for scale up from monitoring location specific indicator status assessments to BHI region score - **Jens says generally reasonable, but GF score for Finland should not be applied to Russian and Estonian waters**  

3. Check assignment of monitoring regions / assessment basins to HOLAS basins.  
**Missing BHI scores** for Kiel Bay (should something be reassigned) and Gdansk Basin (no Polish data).  *Should we gap-fill with adjacent areas?*- **Jens say do not gap fill from adjacent areas, there is not data and populations too local to extrapolate**  

## 5. Calculate Trend


```{r  slope data read in}

slope = read.csv(file.path(dir_ao, 'ao_data_database/ao_cpue_slope.csv'))
dim(slope)
str(slope)
head(slope)
```


## TO DO Trend
1. Within each time series group (key species, cyprinid, piscivore), take the mean slope for each group within each basin  
2. Within each basin take a mean functional group indicator slope (mean of cyprinid mean and piscivore mean)
3. For each basin take overall mean slope - mean of key species and functional group  
4. Multiple by five for future year value?  
5. Apply trend value for basin to all BHI regions (except in Gulf of Finland, do not apply Finnish site value to Estonia and Russian regions.)  

