---
title: "temperature_climatechange_prep"
output:
  github_document:
    toc: true
    toc_depth: 3
params: 
    datasource: csv
---

# Prepare Pressure Layer - Sea Surface Temperature Climate Change pressure
```{r setup}

## Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')

source('~/github/bhi/baltic2015/prep/common.r')

## rprojroot
root <- rprojroot::is_rstudio_project


## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)

dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')


# root$find_file("README.md")
# 
# root$find_file("ao_need_gl2014.csv")
# 
# root <- find_root_file("install_ohicore.r", 
# 
# withr::with_dir(
#   root_file("DESCRIPTION"))

dir_tempcc    = file.path(dir_prep, 'pressures/climate_change/temperature_climatechange')
dir_cc    = file.path(dir_prep, 'pressures/climate_change')

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_tempcc, 'temperature_climatechange_prep.rmd') 
```

## 1. Background
Climate change is resulting in the warming of Baltic Sea surface waters. This poses a direct pressure on biodiversity and food provisioning as well as can indirectly affect water quality through changes in mixing/stratification, nutrient cycling and algal blooms. 

[Second Assessment of Climate Change for the Baltic Sea Basin](http://www.springer.com/gp/book/9783319160054)  


## 2. Data
Summer (July-Aug) sea surface temperature (SST, 0-5 m) by basin.

### 2.1 Data source
Data are from the [BALTSEM model](http://www.balticnest.org/balticnest/thenestsystem/baltsem.4.3186f824143d05551ad20ea.html), run by Bärbel Müller Karulis from the Baltic Sea Centre at Stockholm University.

#### 2.1.1 Hindcast


#### 2.1.2 Projections
There are two different projection scenarios.  

One set of projection scenarios use BALTSEM results run with forcing from the [ECHAM5](http://www.mpimet.mpg.de/en/science/models/echam/) global climate model for the scenario A1b. Project goes to year 2100.  

The second set of  projection scenarios use BALTSEM results run with forcing from the [HADCM3](http://www.metoffice.gov.uk/research/modelling-systems/unified-model/climate-models/hadcm3) global climate model for the scenario A1b. Projection goes to year 2099.  

### 2.2 Data to use
Data from all three above datasets have been explored. We have decided *to use the ECHAM5 dataset for all components of the pressure model (past min temperature, current temperature, future max temperature)*. While forecast data often captures past and future trends, specific values may not align with observed data.  However, each of these datasets have different forcing and previous data exploration showed that using both hindcaset and future projection datasets was problematic given that the datasets did not track one anothers specific value.  Included in the data prep are some of these data explorations.  


## 3. Pressure model

### 3.1 Current conditions
Current conditions = mean summer SST 2010-2014  
Use the hindcast data

### 3.2 rescaling data
Climate projections are for increased temperature. Greater pressure with higher temperature.  

min value = mininium summer SST duing reference period (1960-1990)

max value = maximum SST during the future projection period (2020-2050)

### 3.3 BHI region pressure
Apply basin values to BHI regions.  




## 4. Prepate Data Layer

### 4.1 Read in data

```{r read in data}
## read in data

hind_sst = read.csv(file.path(dir_tempcc, 'temp_data_database/hind_sst.csv'))

proj_sst = read.csv(file.path(dir_tempcc,'temp_data_database/proj_sst.csv')) ## projection from echam5

proj2_sst = read.csv(file.path(dir_tempcc,'temp_data_database/proj2_sst.csv')) ##projection from hadcm3

```

### 4.2 Plot data
```{r plot data}
ggplot(hind_sst)+
  geom_line(aes(year,sst_jul_aug)) + 
  facet_wrap(~basin_name_holas)+
  ggtitle("Hindcast SST time series")

ggplot(proj_sst)+
  geom_line(aes(year,sst_jul_aug)) + 
  facet_wrap(~basin_name_holas)+
  ggtitle("Projected A1b SST time series ECHAM5")

ggplot(proj2_sst)+
  geom_line(aes(year,sst_jul_aug)) + 
  facet_wrap(~basin_name_holas)+
  ggtitle("Projected A1b SST time series HADCM3")


## combine data and plot together

sst_all = bind_rows(
          mutate(hind_sst, type="hindcast"),
          mutate(proj_sst, type="echam5"),
          mutate(proj2_sst, type= "hadcm3"))

ggplot(sst_all)+
  geom_line(aes(year,sst_jul_aug,colour= type)) + 
  facet_wrap(~basin_name_holas)+
  ggtitle("All SST time series")
```



### 4.3 Clean data
```{r clean data}
## remove where year is NA - these are for basins from baltsem not used
hind_sst %>% filter(is.na(year))

hind_sst = hind_sst %>%
           filter(!is.na(year))


proj_sst %>% filter(is.na(year))

proj_sst = proj_sst %>%
           filter(!is.na(year))


proj2_sst %>% filter(is.na(year))

proj2_sst = proj2_sst %>%
           filter(!is.na(year))
```


### 4.3 Current conditions 2010-2014

#### 4.3.1 Using hindcast data
Five most recent years

```{r current sst}
max_year = hind_sst %>%
           select(year)%>%
           max()%>%
           as.numeric()

sst_current = hind_sst %>%
              filter(year >= (max_year-4))%>%
              select(basin_name_holas,sst_jul_aug)%>%
              group_by(basin_name_holas)%>%
              summarise(current_sst = mean(sst_jul_aug))%>%
              ungroup()
              

sst_current

```

#### 4.3.2 Using ECHAM5 data

```{r current sst echam5}
year_range = seq(2010,2014,1)

sst_current_echam5 = hind_sst %>%
             filter(year %in% year_range)%>%
              select(basin_name_holas,sst_jul_aug)%>%
              group_by(basin_name_holas)%>%
              summarise(current_sst = mean(sst_jul_aug))%>%
              ungroup()
              

sst_current_echam5

```


### 4.4.1 Plot Current condition - Compare datasets
Very similar
```{r plot current condition}
ggplot(bind_rows(
      mutate(sst_current, type ="hindcast"),
      mutate(sst_current_echam5, type="echam5")))+
  geom_point(aes(basin_name_holas, current_sst, colour = type, shape=type), size=2.5)+
   theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Mean SST 2010-2014")

```


### 4.5 Historic Min

#### 4.5.1 From Hindcast data
From Hindcast data, extract the minimum Jul-Aug SST for each basin between 1960-1990
```{r historic min sst}

hist_min = hind_sst %>%
           filter(year >= 1960 & year <= 1990)%>%
           select(basin_name_holas,sst_jul_aug)%>%
           group_by(basin_name_holas)%>%
           summarise(hist_min = min(sst_jul_aug))%>%
           ungroup()

hist_min

```

#### 4.5.2 From ECHAM 5 data
Extract the minimum Jul-Aug SST for each basin between 1960-1990
```{r historic min echam5 sst}

hist_min_echam5 = proj_sst %>%
           filter(year >= 1960 & year <= 1990)%>%
           select(basin_name_holas,sst_jul_aug)%>%
           group_by(basin_name_holas)%>%
           summarise(hist_min = min(sst_jul_aug))%>%
           ungroup()

hist_min_echam5

```

#### 4.5.3 Plot and Compare Historical Min value
```{r compare historical min value}
ggplot(bind_rows(
      mutate(hist_min, type ="hindcast"),
      mutate(hist_min_echam5, type="echam5")))+
  geom_point(aes(basin_name_holas, hist_min, colour = type, shape=type), size=2.5)+
   theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Historic Min SST 1960-1990")
```


### 4.6 Future max
From projection data, extract the maximum Jul-Aug SST for each basin between 2020-2050.

Do for each of the projection datasets
```{r future max sst}
future_max1 = proj_sst %>%
             filter(year >= 2020 & year <= 2050)%>%
             select(basin_name_holas,sst_jul_aug)%>%
             group_by(basin_name_holas)%>%
             summarise(future_max = max(sst_jul_aug))%>%
             ungroup()
future_max1

future_max2 = proj2_sst %>%
             filter(year >= 2020 & year <= 2050)%>%
             select(basin_name_holas,sst_jul_aug)%>%
             group_by(basin_name_holas)%>%
             summarise(future_max = max(sst_jul_aug))%>%
             ungroup()
future_max2
```


### 4.7 Plot min, max, and current

#### 4.7.1 Using different datasets
hindcast for historic min and current, different projections compared for the future
```{r plot min, max, current}
##join data for plot
sst_data = full_join(sst_current,hist_min, by="basin_name_holas")%>%
            full_join(.,future_max1, by="basin_name_holas")%>%
            full_join(.,future_max2, by="basin_name_holas") %>%
            dplyr::rename(future_echam5=future_max.x,
                          future_hadcm3 = future_max.y) %>%
            gather(temp_type, temperature, -basin_name_holas)

ggplot(sst_data)+
  geom_point(aes(basin_name_holas,temperature, colour = temp_type, shape=temp_type), size=2)+
  scale_shape_manual(values=c(19,19,17,19))+
  scale_colour_manual(values = c("black","red","orange","blue"))+
    theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Basin Jul-Aug SST Historic Min, Current, Future Max")

```

#### 4.7.2 Using only ECHAM5
```{r plot min, max, current echam 5}
sst_current_echam5 = sst_current_echam5 %>% 
                    dplyr::rename(sst = current_sst)%>%
                    mutate(type = "current")
hist_min_echam5 = hist_min_echam5 %>%
                  dplyr::rename(sst = hist_min)%>%
                  mutate(type = "min")
  
future_max1 = future_max1 %>%
              dplyr::rename(sst = future_max)%>%
                  mutate(type = "max")


sst_data_echam5 = bind_rows(sst_current_echam5,hist_min_echam5,future_max1)

ggplot(sst_data_echam5)+
  geom_point(aes(basin_name_holas,sst, colour = type, shape=type), size=2)+
  scale_shape_manual(values=c(19,17,19))+
  scale_colour_manual(values = c("black","red","blue"))+
    theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                    hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("ECHAM5 Basin Jul-Aug SST Historic Min, Current, Future Max")

```
**Are some issues that current value is warmer than future (will result in pressure value of 1). However, for consistency, continue using all data from ECHAM5** 

### 4.8 Rescale data for pressure layer

```{r rescale to min and max sst}
 sst_rescale = sst_data_echam5 %>%
               spread(type,sst)%>%
               mutate(sst_normalize =(current - min) / (max - min),
                      sst_rescale = pmin(1, sst_normalize))

 sst_rescale
```

#### 4.8.1 Plot Pressure layer by basin

```{r plot temp pressure layer by basin}
ggplot(sst_rescale)+
  geom_point(aes(basin_name_holas,sst_rescale),size=2.5)+
  ylim(0,1)+
  ylab("Pressure value")+
  xlab("Basin")+
  theme(axis.text.x = element_text(colour="grey20", size=8, angle=90, 
                                  hjust=.5, vjust=.5, face = "plain"))+
  ggtitle("Surface temperature pressure layer")



```

### 4.9 Assign basin values to BHI regions

#### 4.9.1 Read in lookup for BHI regions
```{r read in bhi lookup}
bhi_holas_lookup = read.csv(file.path(dir_cc, 'bhi_basin_country_lookup.csv'), sep=";")%>%
                    select(BHI_ID, Subbasin)%>%
                   dplyr::rename(rgn_id = BHI_ID, 
                                 basin_name= Subbasin)

```

#### 4.9.2 join lookup for BHI regions to sst_rescale
```{r join bhi regions to sst_rescale}
sst_rescale1 = sst_rescale %>%
              full_join(., bhi_holas_lookup, by=c("basin_name_holas" = "basin_name"))%>%
              select(rgn_id,sst_rescale)%>%
              dplyr::rename(pressure_score = sst_rescale)%>%
              arrange(rgn_id)

```

#### 4.9.3 Plot rescaled SST by BHI ID
```{r plot sst by bhi id}
ggplot(sst_rescale1)+
  geom_point(aes(rgn_id,pressure_score), size=2.5)+
  ylim(0,1)+
   ggtitle("SST pressure data layer")

```

### 4.10 Write to layers 
```{r write sst to layers}
write.csv(sst_rescale1, file.path(dir_layers, 'cc_sst_bhi2015.csv' ), row.names=FALSE)

```


