---
title: "bottom_trawling_prep"
output:
  github_document:
    toc: true
    toc_depth: 3
params: 
    datasource: csv
---
# Bottom Trawling Pressure Layer


```{r setup}

## Libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RMySQL)
library(stringr)
library(tools)
library(rprojroot) # install.packages('rprojroot')

source('~/github/bhi/baltic2015/prep/common.r')

## rprojroot
root <- rprojroot::is_rstudio_project


## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)



dir_layers = make_path('baltic2015/layers') # replaces  file.path(dir_baltic, 'layers')


# root$find_file("README.md")
# 
# root$find_file("ao_need_gl2014.csv")
# 
# root <- find_root_file("install_ohicore.r", 
# 
# withr::with_dir(
#   root_file("DESCRIPTION"))




dir_trawl    = file.path(dir_prep, 'pressures/bottom_trawling')


## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_trawl, 'bottom_trawling_prep.rmd') 
```

## 1. Background

## 2. Data 

### 2.1 Data sources
[HELCOM Pressures and Human Activities Map Service](http://maps.helcom.fi/website/Pressures/index.html) under 'Pressures and Human Activities' then 'Fisheries' then 'Effort' - there are 8 data layers on *fishing effort mobile bottom-contacting ground'.  

### 2.2 Data attributes
Layers are quarterly effort.  
Units are in hours fished.  

## 3. Pressure Model
Overlay trawl effort with BHI shape files so can assign effort per each BHI region.

### 3.1 Current pressure data
Trawl_quarterly = sum hours within a BHI region / area of BHI region (hours/km2)  

Xtrawl_r_y = annual hours / km2 = sum quarters in each year *y* in each region *r* to get total (hours/km2)  

Xtrawl_r = mean of hours / km2 across years

 
### 3.2 Rescale data
min value = 0  

max value = max(quarterly hours/km2 across all regions) multipled by 4 = max annual  



## 4. Prepare bottom trawl pressure layer

```{r TODO, echo=FALSE, message=FALSE}
## read in data...

## plot data...

```
